### 2018-10-11 code.tex
Initial ShareLaTeX Import

\chapter{Code}\label{cha:code}
You can put code in your document using the listings package, which is
loaded by default in \path{custom.tex}.  Be aware that the listings
package does not put code in your document if you are in draft mode
unless you set the \texttt{forcegraphics} option.

There is an example java (Listing~\ref{src:Data_Bus.java}) and XML
file (Listing~\ref{src:AndroidManifest.xml}).  Thanks to the
\texttt{url} package, you can typeset OSX and unix paths like this:
\path{/afs/rnd.ru.is/project/thesis-template}.  Windows paths:
\path{C:\windows\temp\ }.  You can also typeset them using the menukey
package, but it tends to delete the last separator and has other
complications.\footnote{The menukey package has issues with biblatex,
  read \path{custom.tex} for more information.}

If you are trying to include multiple different languages, you should
go read the documentation and set these up in \path{custom.tex}.  You
will save yourself a lot of effort, especially if you have to fix
anything.

%I have put the source code in the \directory{src/} folder.
\lstinputlisting[language=Java, firstline=1,
lastline=40, caption={Data\_Bus.java: Setting up the class.},
label={src:Data_Bus.java}]{src/Data_Bus.java}

\lstinputlisting[language={[android]XML}, firstline=1, lastline=20,
caption={AndroidManifest.xml: Configuration for the Android UI.},
label={src:AndroidManifest.xml}]{src/AndroidManifest.xml}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "DEGREE-NAME-YEAR"
%%% End:

### 2018-10-11 instructions.tex
Initial ShareLaTeX Import

\documentclass[projectreport,online,forcegraphics,draft]{ruthesis}

%%%%%%%%%%%%%%%%%%%% TeXStudio Magic Comments %%%%%%%%%%%%%%%%%%%%%
%% These comments that start with "!TeX" modify the way TeXStudio works
%% For details see http://texstudio.sourceforge.net/manual/current/usermanual_en.html   Section 4.10
%%
%% What encoding is the file in?
% !TeX encoding = UTF-8
%% What language should it be spellchecked?
% !TeX spellcheck = en_US
%% What program should I compile this document with?
% !TeX program = xelatex

%%%%%%%%%%%%%%%%%%%% Bibliography options %%%%%%%%%%%%%%%%%%%%%
%% We suggest switching from bibtex to biblatex/biber because it is better able
%% to deal with Icelandic characters and other bibliography issues
%% As long as you use biblatex instead of bibtex by itself, it will at least
%%  generate a document without errors.
%% !!!If you are using TeXStudio, don't forget to update the bibliography setting!!!
\usepackage[backend=biber,bibencoding=utf8,style=ieee]{biblatex}
%\DeclareLanguageMapping{american}{american-apa}  
% need to declare mapping for style=apa to alphabetize properly
% If you set backend=bibtex, it will use bibtex for processing (old way)
%    this can work with Icelandic characters, but you may get weird results.
%    bibtex does not know how to sort Þ and ð
% if you set backend=biber, you can use UTF8 characters such as Þ and
%     ð  but you will have to remember to switch from using bibtex to 
%     biber in your client
% If you use JabRef, make sure the file is encoded in UTF-8 which is
%    not the default.

%% This tells TeXStudio to use biber
% !TeX TXS-program:bibliography = txs:///biber
%% This also sets the bibliography program for TeXShop and TeXWorks
% !BIB program = biber

% Where is your reference library?
\addbibresource{references.bib}

\IfFileExists{custom.sty}{\usepackage{custom}}{}
\setTitle{Writing a Thesis in \LaTeX{}}{WRITING A THESIS IN LATEX}
%% Because datetime is broken, we always have to set a \whensigned{day}{MLmonth}{year}
\whensigned{3}{\MLdec}{2018}
\author{Joseph Timothy Foley}
\setCourse{Final Project Reports, Thesis, and Dissertations}
\setSchool{\MLSSE{} \& \MLSCS{}}

%% GLOSSARIES BROKEN!  DO NOT USE
%% Making glossaries needs perl installed
%% Usually only Windows needs this installed separately
%% http://www.activestate.com/activeperl
%\RequirePackage[acronym,toc]{glossaries}
%\RequirePackage[toc]{glossaries}
%% nomain: no main glossary
%% acronym: acronym glossary
%% toc: add entry in Table of Contents
%% automake: try to run makeindex using \write18 shell escape
%% without automake, run "makeglossaries" to process them
%% or the right makeindex options
%\newglossary[slg]{symbols}{sym}{sbl}{List of Symbols}
%\makeglossaries{}% intitialize the glossary.  Must be in the main file!!!
%\input{glossary}% load glossary definitions in preamble

%% You get indexes for free from the memoir package.  Don't use anything else!
%% You have been warned!
\makeindex{}

\finalifforcegraphics{hyperref} %hyperlinks even in draft mode
\usepackage[hidelinks]{hyperref}

\begin{document}
\frontmatter{}
\frontcover{}
\enableindents{}

\starttables{}% setup formatting
%% TOC, list of figures and list of tables are required
\tableofcontents{}\clearpage%%RUM: "Table of contents"
\listoffigures{}\clearpage%%RUM: "List of figures"
\listoftables{}\clearpage%%RUM: "List of tables"
\starttext{}
%\coverchapter{List of drawings and enclosed material}
%RUM: "List of drawings and enclosed material, e.g. CD(as appropriate)"

\listoffixmes{}
% if using fixme package, lists what needs to be done

\chapter{Getting started}
\section{Introduction}
Before using the template in \path{DEGREE-NAME-YEAR.tex}, the author
recommends reading the (in progress) guide to successful projects,
including \LaTeX{} at
\url{http://afs.rnd.ru.is/project/htgaru/trunk/how-to-get-around-projects.pdf}.
Full instructions on setting up \LaTeX{} and SubVersion are there.
Additionally, as issues are discovered, they are added to the
Frequently Asked Questions section in that document.

Text will be double spaced when in draft mode so that there is room
for reviewers to write comments.  Extra blank pages are also removed.
If you want to check formatting in a more final form, switch to
proofing mode.  This text in gray will disappear when you are not in
draft mode.  It is better to use FiXme's as a reminder mechanism,
though.  See Section~\ref{sec:fixme}

\textbf{WARNING} It is very very important that you print the generated PDF at 100\% scaling.   
On many computers, printing is set to adjust the size to fit the page.
Do not do this, because it will mess up the sizes of the font and the placement of the letters.

\subsection{Fonts}
The specified font in the MSc rules is Times or Times New Roman.
If you are in ubuntu linux and wish to use TNR, you will need to:
\verb|sudo apt install ttf-mscorefonts-installer|

\section{Citing properly}
Any content from external sources must be properly cited.
If any part of this document is plagiarized (from another group,
the internet, or anywhere else) or your references are not properly
cited, you will be in a lot of trouble.

This is in compliance with Reykjavík University's Code of
Ethics\footnote{\url{http://en.ru.is/the-university/ru-code-of-ethics/}
  -- see Item 8 and General Rules on Study and Examinations}
\footnote{\url{http://en.ru.is/studies/study--and-examination-rules/}
  -- see Article 4.5} \footnote{see also the document "Your Work -
  Your Contribution" \url{http://www.ru.is/hugverk}, available only in
  Icelandic}. For your reference, here is a link to guidelines
defining plagiarism and discussing how to avoid it (from the
University of Oxford):

\url{http://www.ox.ac.uk/students/academic/goodpractice/about/}

This template uses BibLaTeX and Biber to produce the Bibliography.
This is because they have better support for accented and Icelandic
characters in UTF-8.  They also do more checking on the bibliography
to look for problems.  Unfortunately, this means that if you have a
semi-broken references.bib, Biber will refuse to produce a
Bibliography.  This is a feature because it is telling you that you
need to fix the problems.  BibTeX will also give errors, but will
continue to generate a Bibliography.  This can cause problems later.
For more information read
\url{http://afs.rnd.ru.is/project/htgaru/trunk/how-to-get-around-projects.pdf}.

\section{Document Structure}
The default document structure is the style called IMRAD:\@ Introduction, Methods, Resarch, And, Discussion.  
This is also known as the American Psychological Association (APA) format and is commonly used in the sciences.\cite{web:uta.fi:imrad}.
The template's current editor believes that this format is a good starting point, but may not be appropriate for some engineering works.  
In those cases, following a more technical report style may be easier: Introduction, Background, Design, Implementation, Analysis,  Conclusion.
For some works, this is still too confining.
In those cases, the editor recommends the Introduction, Body, Conclusion (IBC) format where the body is a series of IMRAD or other structures.
Each section of the body should ask a question and answer it relating to the topic.
If you are writing a thesis that is primarily a collection of papers, then the body section would consist of these papers.

\subsection{Reports for classes}
If you are writing a report for a class using this template, you probably want to get rid of a lot of the extra pages.
The set used in these instructions are a good place to start.
See the file \path{minimal-document.tex} for a simpler example.
%See Listing~\ref{src:minimal-document} for a quick example.
% Unicode characters are causing problems for pdftex --foley
%\lstinputlisting[language={[LaTeX]TeX}, caption={Minimal document example},label={src:minimal-document}]{minimal-document.tex}

\chapter{\LaTeX{} techniques}
\section{Fitting things on the page\label{sec:fitting}}
If you see little black boxes on the right in draft mode, this is
\LaTeX{} trying to tell you that it couldn't make it fit nicely.

\section{FixMe\label{sec:fixme} reminders}
\fxnote{This is a note.  It will disappear in final mode, which makes it a safe place to put reminders.}
\fxwarning{This is something more important that needs to be corrected, but will also disappear in final mode.}
\fxerror{This is something very important to be fixed.  This note will disappear in final mode.}
%\fxfatal{This is a fatal issue that must be fixed before submission (final).  
If any of these still exist except for \verb|\fxnote| when we are not in draft mode, \LaTeX{} will have warnings during processing.
If the Fatal error is still there, \LaTeX{} will give errors while producing the document\footnote{This is a {\em feature\/} so that you don't submit the document without fixing really important issues.}.
Unfortunately, many \LaTeX{} UI's have processing set to continue no matter what the error status, so you may only notice this in a little window where the log is displayed.
The Fatal error has been commented out in this document so that we can successfully generate documents but it can be seen in the document's source.

\subsection{Indexes, glossaries, and abbreviations}
This section demonstrates indexes\index{index}.
Currently, the glossary capability is not working --- {\em Do not use it!}
Note that you may get errors in this section if you turn the index options off.
%In order to generate the glossary, you will need to run \verb|makeglossaries|, which requires the Perl language to be installed.  On windows, you can get a perl interpreter at \url{http://www.activestate.com/activeperl}
%See the \verb|README.TXT| for details.

%You might expect to find the term \gls{led} in a \gls{glossary}.
%For maximum safety, the glossary and acronyms should be declared using newglossaryentry before \verb|\begin{%document}| which is done automatically if you put them into \verb|glossary.tex|.
%This is {\em highly recommended\/} in case you reorder your text.
%The author did his \gls{dissertation} on \gls{rfid} security.  
%You can even have symbols, such as the one for \gls{ohm} which is \glssymbol{ohm}

\subsection{Examples}
Some examples of how to use features of \LaTeX{} follow.

\subsubsection{Equations and references}
\begin{figure}
\centering
\includegraphics[height=20mm]{ru-logo}
\caption{The logo of Reykjavík University}\label{fig:ru-logo}
\end{figure}
Figures and Tables need to be properly formatted and referenced in the
text.  Number figures/tables consecutively, include captions, and
refer to the figures/tables in the text
e.g. Figure~\ref{fig:ru-logo}. Equations need to be numbered
consecutively as well.  Equations need to have each of their variables
defined when they are first used or redefined.  If you need to refer
to particular places in the document, use numbered references.  Note
that the \verb|\ref{BLAH}| and the \verb|\label{BLAH}| must match.
When using \verb|\label{}|, it has to be AFTER the item you are trying
to point to without spaces or inside of its declaration.  It is often
safest to put inside e.g.:
\begin{verbatim}\caption{This is a caption\label{fig:mycaption}}\end{verbatim}

If you are wondering why the graphics are missing in your document, you should see if you are in draft mode.
\LaTeX{} speeds up processing in draft mode by just putting boxes so you can focus on the layout and text first.
The \path{rureport.cls} file defines an option \verb|forcegraphics| that can force most included graphics and elements to be enabled even in draft mode.

As an example of an equation, in Equation~\ref{eq:freq} is the
relationship between angular frequency and hertz:
\begin{equation}
f = \frac{1}{T} = \frac{1}{2\pi\omega}\label{eq:freq}
\end{equation}
where $f$ is frequency in \si{\hertz}, $T$ is period in \si{\second},
and $\omega$ is angular frequency in \si{\radian\per\second}.
Resistance is in units of \si{\ohm}.  This example uses \verb|\si|
from the siunitx package to format the units correctly.  If you want
to format it with a value, it uses \verb|\SI{value}{unit}| e.g.:
\verb|\SI{2.99E8}{\meter\per\second}| becomes
\SI{2.99E8}{\meter\per\second}

As an example of referring to a particular part of the document, see
Section~\ref{sec:fitting} on page~\pageref{sec:fitting} for information
on how to fit text on a page.

\subsubsection{Chemical Formulas}
Chemical formulas are best formatted by the mhchem package (included
in the \path{custom.sty} by default).
\href{http://mirrors.ctan.org/macros/latex/contrib/mhchem/mhchem.pdf}{Click
  here for the documentation}.

Here is an example in reaction~\ref{react:so4}:

\reaction[react:so4]{SO4^2- + Ba^2+ -> BaSO4 v}

\subsubsection{Tables}
When making tables, there are some very important rules for making
them look professional.  See
\url{http://ctan.uib.no/macros/latex/contrib/booktabs/booktabs.pdf}
For an example of a simple table, see Table~\ref{tab:example}.  Some general
rules are to always avoid \verb|\hrule| and vertical separator lines.

\begin{table}\centering
\caption{Listing of real name to username\label{tab:example}}
\begin{tabular}{lll}\toprule
Firstname &Lastname &username\\\midrule
Joseph &Foley &foley\\\bottomrule
\end{tabular}
\end{table}

\subsection{Mutilingual support}
In the main file, you have already seen the ML macros used to create multi-lingual documents.
You may want to read more about the babel and polyglossia packages to do even more.
Quotation marks can be made smart by using \enquote{encite}.
This will change the quotation marks to follow whatever the current language is set to.
The Context Sensite Quotes (csquote) package is full of useful macros of this sort. 
One caveat: don't use the shorcut "` and '" in the macros nor in the text\footnote{This is the shortcute for German style quotations, which is the same as Icelandic.}.
It doesn't work due to interactions with the multilanguage macros.
If you need to use those quotes, you can switch languages temporarily,  \selectlanguage{icelandic}
"`Like this'" \resetlanguage{}

\subsection{URLs and paths}
URLs are easy to use in both the bibliography and text.  In your
\path{.bib} file, make sure that you only put URLs into the \path{url}
field.  Putting it anywhere else is likely to cause it to go off the
page.  If you are using the \path{hyperref} package, then these will
be clickable.  You can put links in the text with
\verb|\url| like this: \url{http://www.ru.is}.  If you need to
specify a location of a file, use the
\verb|\path| similarly: \path{C:\Documents and Settings\ }.  The only
difference is that paths are not clickable.

\subsection{Hints}
When placing tables and figures, {\em do not\/} force \LaTeX{} to
place them in a given place.  Doing so almost always causes problems
later.  It is better to give hints \verb|[htb]| than to
force it \verb|!h]|.

Add linebreaks at least at the end of every sentence.  This is
especially important if you are using SVN or collaborating with
others.  Everything is done by line number, so if an error or
issue is on line 5 which happens to be a page of text, good luck
finding the problem.

Along those same lines, if you are not storing your thesis in version
control (SVN, git, etc) or on a backed up drive, you are asking for
trouble.
\subsection{Debugging}
When things go wrong, it is sometimes hard to figure out what to do.
The first step is always to go look at the log file.  If you are
making changes and they are not showing up, it is likely that you are
viewing an old file.  Delete the \path{.pdf} and \path{.aux} and try
again.  On some platforms, the package manager can get out of date.
Make sure you have updated/synchronized before running \path{pdflatex} again.

%% ---------------------------------------------------------------
\printbibliography{} %%RUM: "References"

%% If appendices are needed, uncomment the following line
%% and include the appendices in separate files
\appendix{}%%RUM: "Appendicies (as appropriate)
%ERROR: Unicode characters are giving errors!
%\input{code} % as an example, perhaps some of your code

%\backmatter{} % Sections after this don't get numbers
%% We prefer that all elements be numbered

%%%%%%%%%%%%% SHOW GLOSSARY %%%%%%%%%%%%%%%%%
%% makeindex -s basic.gst -o FILENAME.gls FILENAME.glo
%% Remember to replace FILENAME with the name of this file!
%% If you are using TeXStudio:
%% Options> Configure TexStudio> Build
%% click + next to "User Command"
%% left box: Makeglossary:  
%% right box:  makeindex -s basic.gst -o %.gls %.glo 
\printglossary{}
\printglossary{}%%RUM: Not mentioned

%% Glossary debugging code
%% Uncomment this to figure out of one of the glossary entries is broken
%% by putting all of the entries in the glossary.
%\ifthenelse{\boolean{debug}}{%
%  \glsaddall{} %put all entries in
%  \printglossaries{}}{}% dump them all
%% This command dumps all of the glossaries

%%%%%%%%%%%%% SHOW INDEX %%%%%%%%%%%%%%%%%%
%% Index, optional.  A good idea on longer documents

% You can put instructions at the beginning of the index:
%\renewcommand{\preindexhook}{%
%  The first page number is usually, but not always,
%  the primary reference to the indexed topic.\vskip\onelineskip}

%% You may have to run "makeindex" to have it be generated
%% Depending upon which package you chose.
%% 
\clearforchapter{}
\printindex{}%%RUM: Not mentioned

\backcover{}%%RUM: "Back cover (standard)
\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% TeX-engine: xetex
%%% End:

### 2018-11-16 introduction.tex
version for the meeting 15 nov.

\chapter{Introduction\label{cha:introduction}}
%% \ifdraft only shows the text in the first argument if you are in draft mode.
%% These directions will disappear in other modes.
\ifdraft{State the objectives of the exercise. Ask yourself:
  \underline{Why} did I design/create the item? What did I aim to
  achieve? What is the problem I am trying to solve?  How is my
  solution interesting or novel?}{}

Keflavik International Airport (KEF) is the major gateway to Iceland currently with two functioning runways.

There has been a gradual increase of passengers travelling through Keflavik in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017 \cite{isavia_facts_2017}. For 2018 the passenger numbers show a growth of 13.1\% on average (at the time of writing) from the previous year \cite{isavia_pass_statistics_2018}. The increase in traffic at Keflavik Airport can encounter constrained capacity of the runway during peak traffic periods. 
A significant constraint of this process is caused by separation regulations of aircraft due to wake-vortices. These are specified by rules established by the European Aviation Safety Agency (EASA)  for each type of aircraft based primarily on the weight of the aircraft. The general flight safety requirements are established by the International Civil Aviation Organisation (ICAO) for maintaining safe distance between aircraft

\section{Background and Literature Review}

%\ifdraft{Provide background about the subject matter (e.g. %How was morse code developed?  How is it used today?). 
%This is a place where there are usually many citations.
%It is suspicious when there is not.
%Include the purpose of the different equipment and your %design intent. 
%Include references to relevant scientific/technical work and %books.
%What other examples of similar designs exist?
%How is your approach distinctive?

%If you have specifications or related standards, these %mustscribed and cited also.  As an example, you might cite %the specific
%RoboSub competition website (and documents) if working on the %lighting system for an AUV

%%% Glossary is broken, do not use --foley
%% \gls{auv}\footnote{Autonomous Undersea Vehicle}.
%
%% Notice that there is now information on the AUV in the %Index %and Acronyms.
%% It isn't in the \gls{glossary} because we didn't put it %there.
%\index{AUV}
%}{}

\subsection{Current understandings of wake vortex behaviour}
The wing of an aircraft can be regarded as a beam of finite length with aerofoil as cross section \cite{hansen2015aerodynamics}. The lift effect is created by the differential pressure between the lower and the upper side of the wing. At the wing tips the high pressure flow from the lower side leaks around the tip and to the upper side of the wing. Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards. A finite lifting wing leaves behind a continuous sheet of stream-wise vorticity  \cite{hansen2015aerodynamics}.  This manifests itself in multiple trailing swirls of the airflow that mix together and roll up into two main counter-rotating vortices aft of the wing as illustrated in Figure \ref{fig:vortex_develop} \cite{magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}. Between the vortices the air moves downwards while outside the induced flow is upwards. Those vortices are also known as wake turbulence. 
The kinetic energy contained in the vortices is dependent on the weight and aerodynamics of the generating aircraft. The cross flow velocities in the core region of the trailing vortices can reach $360km/h$ and the vortices can stay effective up to hundred wing spans. This can result in wake vortices lasting for several minutes and up to $30km$ behind larger aircraft \cite{Breitsamter2011Feb, gerz_commercial_2002}.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{graphics/WakeVortexPlane.png}
    \caption[Wake vortex roll-up process]{Wing vortex evolution and roll-up process. Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one \cite{magazine_aibus_safety}} \label{fig:vortex_develop}
\end{figure}

\subsection{Ground effect and decay process}

Near ground on approach the wake vortices tend to descend slowly to an altitude of about one half of the initial separation.
Upon reaching this point the descent will stop and then ascend slowly. This "re-bounce" effect is caused by the presence of the ground and carries with it the formation of a second pair of induced vortices outside and bellow the main vortex. 
In the presence of stable cross-wind conditions the decay of the downwind vortex is not identical to the upwind vortex and the descent rate may vary significantly \cite{Hallock2018Apr}. 
The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure \ref{fig:vortex_ground_effect}).
Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229 m$ and $518 m$ and in some cases even $762 m$, which is the basis for the "2500-foot"  rule separation in parallel runways configuration \cite{Hallock2018Apr, hallock2004summary, hallock2003wake}. The vertical descent rate for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5m/s$ for the first $30s$ after which the descent slows and eventually approaches zero at $152-274m$ below the flight path, and for heavier aircraft decaying vortices have been observed at $305m$ and below of the flight path  \cite{lissaman1973aircraft, Hallock2018Apr}. 
Aerodynamic properties of the aircraft govern the vortex roll-up process and ambient atmospheric conditions dominate the behaviour of the vortices forcing instability and eventual decay of the vortices \cite{Hallock2018Apr}.
several factors that drive the decay rate of the vortex pair have formulated:
\begin{itemize}
    \item Atmospheric turbulence extracts energy from the vortex and reduces its strength leading to a faster wake decay \cite{Hallock2018Apr}.
    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence \cite{Hallock2018Apr}.
    \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level \cite{Holzapfel2001Feb, gerz_commercial_2002}.
    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may causes the vortices to link and decay faster \cite{Hallock2018Apr, crow2003stability}.
    \item Secondary vorticity structures, vertical rib-like counter rotating flow formations may appear between the vortex pair and eventually wrap around the main vortices, leading to turbulence build inside the system and rapid circulation decay\cite{Holzapfel2001Feb, Holzapfel2003Jun}.
\end{itemize}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{graphics/Hallock_vortex_evolution.jpg}
    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect \cite{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
\end{figure}

\subsection{Wake vortex separation}
An aircraft affected by the wake vortex can experience loss of lift or an induced rolling moment and velocity fluctuations \ref{fig:vortex_encounter}.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{graphics/reaction_in_wake.jpg}
    \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft. (A) induced rolling moment, (B) upward motion, (C) loss of lift  \cite{Hallock2018Apr}} \label{fig:vortex_encounter}
\end{figure}

This can lead to the up-set and potentially loss of control of a smaller aircraft following the flight path of a preceding aircraft.
To diminish the effect of such vortices the trailing aircraft must be maintained at a safe distance behind the lead aircraft as the vortices spread laterally to either side of the flight path and are dissipated.\cite{Breitsamter2011Feb}. \\
The distance between aircraft pairs in view of safety is referred to as wake vortex separation. 
Traditional separation standards, introduced in the seventies, are defined by ICAO  on the basis of certificated maximum take-off mass (MTOM) as shown in Table \ref{tab:WTC}. The prescribed categories are three (i.e.Heavy, Medium and Light). In addition for aircraft in the order of $560000kg$ ICAO provides a subcategory to the Heavy types called Super Heavy. Some states have chosen to further increase the number of categories to adapt the wake turbulence categories (WTC) to specific airport requirements. Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table \ref{tab:WTC}), while the Federal Aviation Administration (FAA) places the B757 into a separate category and increases the upper weight limit of the Light category. \cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}

\begin{table}
    \centering
    \resizebox{1\textwidth}{!} {
    \begin{tabular}{l|c|c|c|c|c|c}
    ~    & \multicolumn{6}{c}{Category} \\ \hline
    ICAO & Super Heavy & Heavy (H) & \multicolumn{2}{c|}{Medium (M)} & \multicolumn{2}{c}{Light (L)} \\

    ~    & A380-800    & $ > 136000 $  & \multicolumn{2}{c|}{$ 136000-7000 $ } & \multicolumn{2}{c}{$ \leq 7000 $} \\ \hline

    ICAO (UK)   & ~  & Heavy (H) & Upper Medium (UM) & Lower Medium (LM) & Small (S)  & Light (L) \\

    ~    & A380-800    & $ > 136000 $  & $ 136000-104000 $     & $ 104000-40000 $      & $ 40000-17000 $ & $ \leq 17000 $   \\ \hline

    ICAO \& FAA (US)   & Super      & Heavy (H) & B757   & Large   & \multicolumn{2}{c}{Small} \\

    ~    & A380        & $ > 136000 $  & ~                 & $ 136000-18600 $      & \multicolumn{2}{c}{$ \leq 18600 $} \\ 
    \end{tabular}}
    \caption[ICAO wake turbulence categories based on maximum take-off mass]{ICAO wake turbulence categories based on maximum take-off mass (MTOM) in kg. \cite{icao_wtc, uk_aeronautical_information_services_wake_2017, kolos2013influence}} \label{tab:WTC}
\end{table}

The ICAO wake turbulence categories outline the worst case for each category and thus generate over-separation in many cases\cite{noauthor_recat_2018}. The separation distances for instrumental flight rules are observed for the follower aircraft and are conventionally expressed in nautical miles (NM). The separation reverts to radar separation minimum (MRS) when wake turbulence restrictions are not required. As  prescribed  by  ICAO  the  horizontal MRS is 3NM(or 2.5NM under given conditions described in ICAO Doc 4444 PANS-AM) \cite{noauthor_icao_2018}, which translates to 79s (66s) time separation between leader and follower at approach speed of 70 m/s.

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{ICAO WTC scheme}} & \multicolumn{4}{c|}{Follower}              \\ \cline{3-6} 
\multicolumn{2}{|c|}{}                                 & Super (A380-800) & Heavy & Medium & Light \\ \hline
\multirow{4}{*}{Leader}       & Super (A380-800)       & ~                & 6 NM  & 7 NM   & 8 NM   \\ \cline{2-6} 
                              & Heavy                  & ~                & 4 NM  & 5 NM   & 6 NM   \\ \cline{2-6} 
                              & Medium                 & ~                & ~     & ~      & 5 NM   \\ \cline{2-6} 
                              & Light                  & ~                & ~     & ~      & ~      \\ \hline
\end{tabular}%
}
\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter. \cite{noauthor_recat_2018}} \label{tab:ICAO_WTC}
\end{table}

%There is an RU logo in Figure~\ref{fig:ru-logo}.
%This logo will scale according to the width of the text on the page.

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
\multicolumn{2}{|c|}{}                                 & CAT\_A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
\multirow{6}{*}{Leader}             & CAT-A            & 3 NM   & 4 NM  & 5 NM  & 5 NM  & 6 NM  & 8 NM  \\ \cline{2-8} 
                                    & CAT-B            &        & 3 NM  & 4 NM  & 4 NM  & 5 NM  & 7 NM  \\ \cline{2-8} 
                                    & CAT-C            &        & (*)   & 3 NM  & 3 NM  & 4 NM  & 6 NM  \\ \cline{2-8} 
                                    & CAT-D            &        &       &       &       &       & 5 NM  \\ \cline{2-8} 
                                    & CAT-E            &        &       &       &       &       & 4 NM  \\ \cline{2-8} 
                                    & CAT-F            &        &       &       &       &       & 3 NM  \\ \hline
\end{tabular}%
}
\caption[RECAT-EU  distance-based separation minima]{RECAT-EU WT distance-based separation minima on approach and departure (*) means minimum radar separation (MRS), set at 2.5 NM, is applicable as per current ICAO doc 4444 provisions.}
\label{tab:RECAT-dist}
\end{table}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
\multicolumn{2}{|c|}{}                                 & CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
\multirow{6}{*}{Leader}             & CAT-A            &        & 100s  & 120s  & 140s  & 160s  & 180s  \\ \cline{2-8} 
                                    & CAT-B            &        &       &       & 100s  & 120s  & 140s  \\ \cline{2-8} 
                                    & CAT-C            &        &       &       & 80s   & 100s  & 120s  \\ \cline{2-8} 
                                    & CAT-D            &        &       &       &       &       & 120s  \\ \cline{2-8} 
                                    & CAT-E            &        &       &       &       &       & 100s  \\ \cline{2-8} 
                                    & CAT-F            &        &       &       &       &       & 80s   \\ \hline
\end{tabular}%
}
\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure}
\label{tab:RECAT-time}
\end{table}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                                                                                                                   \\ \cline{3-8} 
\multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A                             & CAT-B                                & CAT-C                         & CAT-D                         & CAT-E                         & CAT-F                                                \\ \hline
                                                        & CAT-A & \cellcolor[HTML]{FD6864}(+0,5 NM) & \cellcolor[HTML]{67FD9A}-2 NM        & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8} 
                                                        & CAT-B &                                   & \cellcolor[HTML]{67FD9A}-1 NM        &                               & \cellcolor[HTML]{67FD9A}-1 NM &                               & \cellcolor[HTML]{FD6864}+1 NM                        \\ \cline{2-8} 
                                                        & CAT-C &                                   & \cellcolor[HTML]{67FD9A}-1 (-1,5) NM & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8} 
                                                        & CAT-D &                                   &                                      &                               &                               &                               &                                                      \\ \cline{2-8} 
                                                        & CAT-E &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{67FD9A}{\color[HTML]{000000} -1 NM} \\ \cline{2-8} 
\multirow{-6}{*}{Leader}                                & CAT-F &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{FD6864}(+0,5 NM)                    \\ \hline 
\end{tabular}%
}
\caption{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)}
\label{tab:delta_distance_wtc2recat}
\end{table}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                             \\ \cline{3-8} 
\multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A & CAT-B                        & CAT-C & CAT-D                        & CAT-E                        & CAT-F                        \\ \hline
                                                        & CAT-A &       & \cellcolor[HTML]{67FD9A}-20s &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8} 
                                                        & CAT-B &       &                              &       & \cellcolor[HTML]{67FD9A}-20s &                              & \cellcolor[HTML]{FD6864}+20s \\ \cline{2-8} 
                                                        & CAT-C &       &                              &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8} 
                                                        & CAT-D &       &                              &       &                              &                              &                              \\ \cline{2-8} 
                                                        & CAT-E &       &                              &       &                              &                              & \cellcolor[HTML]{67FD9A}-20s \\ \cline{2-8} 
\multirow{-6}{*}{Leader}                                & CAT-F &       &                              &       &                              &                              & \cellcolor[HTML]{FD6864}+20s \\ \hline
\end{tabular}
}
\caption{Difference in wake separation on departure between reference ICAO and RECAT-EU schemes(full proposal)}
\label{tab:delta_times_wtc2recat}
\end{table}

\subsection{Runway Occupancy}
\fxnote{describe rot in general and what factors have influence on rot, theory, models for rot, see validation for rot model paper}
The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway\cite{ROT_definition}. This project focuses on arrivals runway occupancy (AROT) in peak traffic hours, that is the time interval between the aircraft crossing the threshold and its tail vacating the runway\cite{AROT_definition}. The threshold is the beginning of that portion of the runway that is available for landing.
The peak traffic hours are characterised by the time intervals when the airport operates at high load. High load interval/window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavik airport. The time between flights, i.e. the time between start of runway occupancy, during the high load interval is less than or equal to X minutes (currently X = 4 in the script). The last flight in a window that fulfils these requirements is not considered to be a part of the peak as its behaviour is not affected by an instantaneous next flight after it.

The data collection on Runway Occupancy Time (ROT) in Isavia starts on 21.Nov 2014 with the acceptance of the Automatic Dependent Surveillance-Broadcast (ADS-B) system by the company. ADS-B was preferred to radar at Keflavik airport because of signal accuracy \cite{isavia_wiki}. The aircraft surveillance data specifications used for measurements are in the Eurocontrol ASTERIX Category 21 standard \cite{ASTERIX_ADS-B_specs}.
The reading from the position of the ADS-B antenna of the aircraft is used instead of the tail and nose positions. This simplification could result in a few seconds difference. The ADS-B equipped aircraft at Keflavik airport is estimated around $90\%$ \cite{isavia-rounardeild_rannsoknir_2018}.  Certain filtering of the ADS-B data is applied before commencing measurements: 
\begin{itemize}
    \item All records with no latitude, longitude and/or time data are omitted.
    \item The records are filtered with regards to quality, i.e. filtered with regards to Target Surveillance status, MOPS version, NUC/NIC, NACp, SIL and PIC.
    \item An ADS-B data record within 0.5 second from another is omitted, i.e. if there are less than 0.5 seconds between two successive locations one of the records is omitted.
    \item An algorithm is used to analyse the data and determine if it originates from two different flights (the flight stops for 15 minutes or more after/before taxiing). If so the data for the second flight is omitted
    \item The velocity is calculated from ADS-B location data and ADS-B time data (velocity = distance travelled/time). The ADS-B velocity record is optional in the ASTERIX category 021 standard and is therefore unreliable in measurements. Since the velocity is calculated from measured data, it can exhibit spikes. To get a more realistic velocity curve, a Savitzky-Golay filter is applied to smooth the data.
    \item An aircraft main gear lift-off is considered to be the point where the ADS-B ground bit is removed, i.e. the ADS-B ground bit goes from a value of 1 to a value of 0.
    \item An aircraft touchdown is considered to be the point where the ADS-B ground bit is set, i.e. the ADS-B ground bit goes from a value of 0 to a value of 1.
    \item An aircraft is considered to be stationary when the velocity is below 0.5 knots and to be moving if the velocity is greater than 0.5 knots.
    \cite{isavia_wiki}
\end{itemize}

\fxnote{The ROT measurements have been checked by: comparing them to measurements made by hand.
Who measures and who checks them and how???}

% Please add the following required packages to your document preamble:
% \usepackage{graphicx}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/AROT_hist.png}
    \caption[AROT histogram]{AROT histogram}
    \label{fig:AROT_hist}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/AROT_hist_cumul.png}
    \caption[AROT histogram and cumulative distribution]{AROT histogram and cumulative distribution}
    \label{fig:AROT_hist_cumul}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/pdf_wtc.png}
    \caption[PDF for WTC]{PDF for WTC}
    \label{fig:pdf_wtc}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/pdf_recats.png}
    \caption[PDF for RACAT]{PDF fro RACAT}
    \label{fig:pdf_recats}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/each_recat_box.png}
    \caption[Boxplot of AROT for RECAT categories]{Boxplot of AROT for RECAT categories}
    \label{fig:each_recat_box}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/each_recat_hist.png}
    \caption[AROT histogram for each RECAT]{AROT histogram for each RECAT}
    \label{fig:each_recat_hist}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/each_recat_hist_dist.png}
    \caption[AROT histogram and pdf for each RECAT]{AROT histogram and pdf for each RECAT}
    \label{fig:each_recat_hist_dist}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/each_recat_rug_cumdist.png}
    \caption[AROT rug plot and cumulative pdf for each RECAT]{AROT rug plot and cumulative pdf for each RECAT}
    \label{fig:each_recat_rug_cumdist}
\end{figure}

\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{l|r|r|r|r|r|r|r|r|}
\cline{2-9}
                               & \multicolumn{8}{c|}{AROT}                                                              \\ \hline
\multicolumn{1}{|l|}{RECAT-EU} & count  & mean      & std       & min       & 25\%      & 50\%      & 75\%      & max   \\ \hline
\multicolumn{1}{|l|}{CAT-A}    & 3      & 108       & 22        & 88        & 96        & 105       & 119       & 132   \\ \hline
\multicolumn{1}{|l|}{CAT-B}    & 147    & 86        & 19        & 51        & 70        & 85        & 100       & 139   \\ \hline
\multicolumn{1}{|l|}{CAT-C}    & 8909   & 80        & 17        & 14        & 67        & 79        & 91        & 156   \\ \hline
\multicolumn{1}{|l|}{CAT-D}    & 2177   & 78        & 19        & 46        & 64        & 77        & 92        & 155   \\ \hline
\multicolumn{1}{|l|}{CAT-E}    & 168    & 86        & 24        & 37        & 71        & 85        & 101       & 149   \\ \hline
\multicolumn{1}{|l|}{CAT-F}    & 96     & 85        & 26        & 43        & 69        & 81        & 97        & 155   \\ \hline
\end{tabular}%
}
\caption{AROT statistics for airtrafic mix at KEF by RECAT categories}
\label{tab:AROT_RECAT_ stats}
\end{table}

\subsection{Runway Capacity}

\fxnote{operational procedures for air traffic control at kef, }
\fxnote{equipment specifications in the tower that might influence the WTC sepatarion decision-talk to Haraldur}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "DEGREE-NAME-YEAR"
%%% End:

### 2018-12-16 code.tex
Graphs and tables included,. Started the Methods section,

@@
+\chapter{Arrival Runway Occupancy Times}\label{cha:AROTs}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+\multicolumn{2}{l|}{RWY01}  & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
+\multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{768} & 85 & 20 & 44 & 69 & 85 & 99 & 153 \\ \hline
+\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{263} & 100 & 17 & 55 & 87 & 98 & 111 & 156 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs RWY-01 pre fast exit by season]{AROTs for runway RWY-01 pre-fast-exit TWY A-1 by season}
+\label{tab:season_AROT_stats_RWY01_pre_fast_exit}
+\end{table}
+
+% ----------------------------
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+\multicolumn{2}{l|}{RWY01, TWY A-1}  & \multicolumn{7}{c|}{AROT  {[}s{]}} \\ \hline
+\multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{1900} & 83 & 22 & 41 & 63 & 86 & 98 & 152 \\ \hline
+\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{286} & 88 & 26 & 43 & 62 & 90 & 106 & 153 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs RWY-01 post fast exit by season]{AROTs for runway RWY-01 post-fast-exit TWY A-1 by season}
+\label{tab:season_AROT_stats_RWY01_post_fast_exit}
+\end{table}
+
+% ----------------------------
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+\multicolumn{2}{l|}{RWY28} & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
+\multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{554} & 99 & 13 & 37 & 91 & 98 & 106 & 153 \\ \hline
+\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{250} & 109 & 14 & 37 & 100 & 108 & 117 & 155 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs RWY-28 pre fast exit by season]{AROTs for runway RWY-28 pre-fast-exit TWY B-1 by season}
+\label{tab:season_AROT_stats_RWY28_pre_fast_exit}
+\end{table}
+
+% ----------------------------
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+\multicolumn{2}{l|}{RWY28, TWY B-1} & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
+\multicolumn{1}{|l|}{Season} & {count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{Summer} & 402 & 68 & 14 & 42 & 60 & 66 & 72 & 155 \\ \hline
+\multicolumn{1}{|l|}{Winter} & 69 & 73 & 13 & 49 & 63 & 72 & 79 & 105 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs RWY-28 post fast exit by season]{AROTs for runway RWY-28 post-fast-exit TWY B-1 by season}
+\label{tab:season_AROT_stats_RWY28_post_fast_exit}
+\end{table}
+
+% ----------------------------
+
 \chapter{Code}\label{cha:code}

### 2018-12-16 introduction.tex
Graphs and tables included,. Started the Methods section,

@@
 
-Keflavik International Airport (KEF) is the major gateway to Iceland currently with two functioning runways.
-
-There has been a gradual increase of passengers travelling through Keflavik in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017 \cite{isavia_facts_2017}.
+Keflavík International Airport (IATA:KEF, ICAO:BIKF) is the major gateway to Iceland currently with two functioning runways. 
+There has been a gradual increase of passengers travelling through Keflavik Airport in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017 \cite{isavia_facts_2017}.
 For 2018 the passenger numbers show a growth of 13.1\% on average (at the time of writing) from the previous year \cite{isavia_pass_statistics_2018}.
-The increase in traffic at Keflavik Airport can encounter constrained capacity of the runway during peak traffic periods.
+The increase in traffic at Keflavík Airport can encounter constrained capacity of the runway during peak traffic periods.
 A significant constraint of this process is caused by separation regulations of aircraft due to wake-vortices.
@@
 \subsection{Current understandings of wake vortex behaviour}
-The wing of an aircraft can be regarded as a beam of finite length with aerofoil as cross section \cite{hansen2015aerodynamics}.
+The wing of an aircraft can be regarded as a beam of finite length with aerofoil as cross section \cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation or drop it if unnecessary.
+Read about induced vertical flow in a book on aerodynamics}.
 The lift effect is created by the differential pressure between the lower and the upper side of the wing.
@@
 Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards.
-A finite lifting wing leaves behind a continuous sheet of stream-wise vorticity  \cite{hansen2015aerodynamics}.
+A finite lifting wing leaves behind a continuous sheet of stream-wise vorticity  \cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation }.
 This manifests itself in multiple trailing swirls of the airflow that mix together and roll up into two main counter-rotating vortices aft of the wing as illustrated in Figure \ref{fig:vortex_develop} \cite{magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
@@
 The kinetic energy contained in the vortices is dependent on the weight and aerodynamics of the generating aircraft.
-The cross flow velocities in the core region of the trailing vortices can reach $360km/h$ and the vortices can stay effective up to hundred wing spans.
-This can result in wake vortices lasting for several minutes and up to $30km$ behind larger aircraft \cite{Breitsamter2011Feb, gerz_commercial_2002}.
+The cross flow velocities in the core region of the trailing vortices can reach $360$ km/h and the vortices can stay effective up to hundred wing spans.
+This can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft \cite{Breitsamter2011Feb, gerz_commercial_2002}.
 
@@
     \caption[Wake vortex roll-up process]{Wing vortex evolution and roll-up process.
-Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one \cite{magazine_aibus_safety}} \label{fig:vortex_develop}
+Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one \cite[p.
+043]{magazine_aibus_safety}} \label{fig:vortex_develop}
 \end{figure}
@@
 The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure \ref{fig:vortex_ground_effect}).
-Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229 m$ and $518 m$ and in some cases even $762 m$, which is the basis for the "2500-foot"  rule separation in parallel runways configuration \cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
-The vertical descent rate for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5m/s$ for the first $30s$ after which the descent slows and eventually approaches zero at $152-274m$ below the flight path, and for heavier aircraft decaying vortices have been observed at $305m$ and below of the flight path  \cite{lissaman1973aircraft, Hallock2018Apr}.
+Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$ m and $518$ m and in some cases even $762$ m, which is the basis for the "2500-foot"  rule separation in parallel runways configuration \cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
+The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5$ m/s for the first $30$ s after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$ m and below of the flight path  \cite{lissaman1973aircraft, Hallock2018Apr}.
 Aerodynamic properties of the aircraft govern the vortex roll-up process and ambient atmospheric conditions dominate the behaviour of the vortices forcing instability and eventual decay of the vortices \cite{Hallock2018Apr}.
-several factors that drive the decay rate of the vortex pair have formulated:
+Several factors that drive the decay rate of the vortex pair have been formulated:
 \begin{itemize}
     \item Atmospheric turbulence extracts energy from the vortex and reduces its strength leading to a faster wake decay \cite{Hallock2018Apr}.
-    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence \cite{Hallock2018Apr}.
+    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence \cite{Hallock2018Apr}.\fxnote{This needs to be explained (Ármann)}
     \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level \cite{Holzapfel2001Feb, gerz_commercial_2002}.
-    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may causes the vortices to link and decay faster \cite{Hallock2018Apr, crow2003stability}.
+    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may cause the vortices to link and decay faster \cite{Hallock2018Apr, crow2003stability}.
     \item Secondary vorticity structures, vertical rib-like counter rotating flow formations may appear between the vortex pair and eventually wrap around the main vortices, leading to turbulence build inside the system and rapid circulation decay\cite{Holzapfel2001Feb, Holzapfel2003Jun}.
@@
     \includegraphics[width=0.8\textwidth]{graphics/Hallock_vortex_evolution.jpg}
-    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect \cite{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
+    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect \cite[p. 29]{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
 \end{figure}
@@
 \subsection{Wake vortex separation}
-An aircraft affected by the wake vortex can experience loss of lift or an induced rolling moment and velocity fluctuations \ref{fig:vortex_encounter}.
+An aircraft affected by the wake vortex can experience loss of lift or an induced rolling moment and velocity fluctuations (Figure \ref{fig:vortex_encounter}).
+This can lead to the up-set and potentially loss of control of an aircraft following the flight path of a preceding aircraft.
+To diminish the effect of such vortices the trailing aircraft must be maintained at a safe distance behind the lead aircraft as the vortices spread laterally to either side of the flight path and are dissipated \cite{Breitsamter2011Feb}.
+The distance between aircraft pairs in view of safety is referred to as wake vortex separation.
 
@@
     \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft.
-(A) induced rolling moment, (B) upward motion, (C) loss of lift  \cite{Hallock2018Apr}} \label{fig:vortex_encounter}
+(A) induced rolling moment, (B) upward motion, (C) loss of lift  \cite[p.
+33]{Hallock2018Apr}} \label{fig:vortex_encounter}
 \end{figure}
 
-This can lead to the up-set and potentially loss of control of a smaller aircraft following the flight path of a preceding aircraft.
-To diminish the effect of such vortices the trailing aircraft must be maintained at a safe distance behind the lead aircraft as the vortices spread laterally to either side of the flight path and are dissipated.\cite{Breitsamter2011Feb}.
-\\
-The distance between aircraft pairs in view of safety is referred to as wake vortex separation. 
 Traditional separation standards, introduced in the seventies, are defined by ICAO  on the basis of certificated maximum take-off mass (MTOM) as shown in Table \ref{tab:WTC}.
 The prescribed categories are three (i.e.Heavy, Medium and Light).
-In addition for aircraft in the order of $560000kg$ ICAO provides a subcategory to the Heavy types called Super Heavy.
+In addition for aircraft in the order of $560000$ kg ICAO provides a subcategory to the Heavy types called Super Heavy.
 Some states have chosen to further increase the number of categories to adapt the wake turbulence categories (WTC) to specific airport requirements.
-Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table \ref{tab:WTC}), while the Federal Aviation Administration (FAA) places the B757 into a separate category and increases the upper weight limit of the Light category.
+Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table \ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing B757 into a separate category and increases the upper weight limit of the Light category.
 \cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
 
-\begin{table}
+\begin{table}[]
     \centering
@@
 
-The ICAO wake turbulence categories outline the worst case for each category and thus generate over-separation in many cases\cite{noauthor_recat_2018}.
+The ICAO wake turbulence categories outline the worst case for each category and thus generate over-separation in many cases \cite{noauthor_recat_2018}.
 The separation distances for instrumental flight rules are observed for the follower aircraft and are conventionally expressed in nautical miles (NM).
 The separation reverts to radar separation minimum (MRS) when wake turbulence restrictions are not required.
-As  prescribed  by  ICAO  the  horizontal MRS is 3NM(or 2.5NM under given conditions described in ICAO Doc 4444 PANS-AM) \cite{noauthor_icao_2018}, which translates to 79s (66s) time separation between leader and follower at approach speed of 70 m/s.
+As  prescribed  by  ICAO  the  horizontal MRS is 3 NM (or a reduced separation minimum of 2.5 NM may be applied under given conditions described in ICAO Doc 4444 PANS-AM \cite{doc44444}), which translates to 79s (66s) time separation between leader and follower at approach speed of 70 m/s.
 
@@
 \multicolumn{2}{|c|}{}                                 & Super (A380-800) & Heavy & Medium & Light \\ \hline
-\multirow{4}{*}{Leader}       & Super (A380-800)       & ~                & 6 NM  & 7 NM   & 8 NM   \\ \cline{2-6} 
-                              & Heavy                  & ~                & 4 NM  & 5 NM   & 6 NM   \\ \cline{2-6} 
-                              & Medium                 & ~                & ~     & ~      & 5 NM   \\ \cline{2-6} 
-                              & Light                  & ~                & ~     & ~      & ~      \\ \hline
+\multirow{4}{*}{Leader}       & Super (A380-800)       & (*)                & 6 NM  & 7 NM   & 8 NM   \\ \cline{2-6} 
+                              & Heavy                  & (*)                & 4 NM  & 5 NM   & 6 NM   \\ \cline{2-6} 
+                              & Medium                 & (*)                & (*)     & (*)      & 5 NM   \\ \cline{2-6} 
+                              & Light                  & (*)                & (*)     & (*)      & (*)      \\ \hline
 \end{tabular}%
 }
-\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter. \cite{noauthor_recat_2018}} \label{tab:ICAO_WTC}
+\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter.
+\cite{noauthor_recat_2018, rooseleer2015recat}} \label{tab:ICAO_WTC}
 \end{table}
@@
 
-% Please add the following required packages to your document preamble:
-% \usepackage{multirow}
-% \usepackage{graphicx}
 % Please add the following required packages to your document preamble:
@@
 \multirow{6}{*}{Leader}             & CAT-A            & 3 NM   & 4 NM  & 5 NM  & 5 NM  & 6 NM  & 8 NM  \\ \cline{2-8} 
-                                    & CAT-B            &        & 3 NM  & 4 NM  & 4 NM  & 5 NM  & 7 NM  \\ \cline{2-8} 
-                                    & CAT-C            &        & (*)   & 3 NM  & 3 NM  & 4 NM  & 6 NM  \\ \cline{2-8} 
-                                    & CAT-D            &        &       &       &       &       & 5 NM  \\ \cline{2-8} 
-                                    & CAT-E            &        &       &       &       &       & 4 NM  \\ \cline{2-8} 
-                                    & CAT-F            &        &       &       &       &       & 3 NM  \\ \hline
+                                    & CAT-B            &    (*)    & 3 NM  & 4 NM  & 4 NM  & 5 NM  & 7 NM  \\ \cline{2-8} 
+                                    & CAT-C            &    (*)    & (*)   & 3 NM  & 3 NM  & 4 NM  & 6 NM  \\ \cline{2-8} 
+                                    & CAT-D            &    (*)    &   (*)    &    (*)   &   (*)    &   (*)    & 5 NM  \\ \cline{2-8} 
+                                    & CAT-E            &    (*)    &   (*)    &   (*)    &   (*)    &   (*)    & 4 NM  \\ \cline{2-8} 
+                                    & CAT-F            &    (*)    &   (*)    &    (*)   &   (*)    &   (*)    & 3 NM  \\ \hline
 \end{tabular}%
 }
-\caption[RECAT-EU  distance-based separation minima]{RECAT-EU WT distance-based separation minima on approach and departure (*) means minimum radar separation (MRS), set at 2.5 NM, is applicable as per current ICAO doc 4444 provisions.}
+\caption[RECAT-EU distance-based separation minima]{RECAT-EU WT distance-based separation minima on approach and departure.
+(*) means minimum radar separation (MRS), set at 2.5 NM, applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat}.}
 \label{tab:RECAT-dist}
@@
 }
-\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure}
+\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure \cite{rooseleer2015recat}}
 \label{tab:RECAT-time}
@@
 }
-\caption{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)}
+\caption[Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal) \cite{rooseleer2015recat}}
 \label{tab:delta_distance_wtc2recat}
@@
 }
-\caption{Difference in wake separation on departure between reference ICAO and RECAT-EU schemes(full proposal)}
+\caption[Time difference between reference ICAO and RECAT-EU schemes]{Time difference in wake separation on departure between reference ICAO and RECAT-EU schemes(full proposal) \cite{rooseleer2015recat}}
 \label{tab:delta_times_wtc2recat}
@@
 
-\subsection{Runway Occupancy}
+\subsection{Runway Occupancy Time}
+
 \fxnote{describe rot in general and what factors have influence on rot, theory, models for rot, see validation for rot model paper}
-The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway\cite{ROT_definition}.
-This project focuses on arrivals runway occupancy (AROT) in peak traffic hours, that is the time interval between the aircraft crossing the threshold and its tail vacating the runway\cite{AROT_definition}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{graphics/fig_RECAT_AROTs_boxplot.png}
+    \caption[AROTs boxplot for RECAT categories, all runways]{Arrival Runway Occupancy Times for the different RECAT-EU categories based on data gathered for a period of one year (since October, 2017).
+The box plot shows the AROTs for all runways at BIKF.}
+    \label{fig:RECAT_AROTs_boxplot}
+\end{figure}
+
+The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway \cite{ROT_definition}. \\
+This project focuses on Arrivals Runway Occupancy Time (AROT) in peak traffic hours, that is the time interval between the aircraft crossing the threshold and its tail vacating the runway \cite{AROT_definition}.
 The threshold is the beginning of that portion of the runway that is available for landing.
 The peak traffic hours are characterised by the time intervals when the airport operates at high load.
-High load interval/window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavik airport.
-The time between flights, i.e.
-the time between start of runway occupancy, during the high load interval is less than or equal to X minutes (currently X = 4 in the script).
-The last flight in a window that fulfils these requirements is not considered to be a part of the peak as its behaviour is not affected by an instantaneous next flight after it.
+High load interval or window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavik Airport.
+The time between flights during the high load interval is set at $\leq$4 minutes.
+This value is based on statistical data by Isavia in order to achieve two distinct peak hours: one in the morning and one in the afternoon.
+The last flight in a high load window that fulfils these requirements is not considered to be a part of the peak as its behaviour is not affected by an instantaneous next flight after it.
+\fxnote{Peak hours need to be explained, Include picture}
 
-The data collection on Runway Occupancy Time (ROT) in Isavia starts on 21.Nov 2014 with the acceptance of the Automatic Dependent Surveillance-Broadcast (ADS-B) system by the company.
-ADS-B was preferred to radar at Keflavik airport because of signal accuracy \cite{isavia_wiki}.
-The aircraft surveillance data specifications used for measurements are in the Eurocontrol ASTERIX Category 21 standard \cite{ASTERIX_ADS-B_specs}.
-The reading from the position of the ADS-B antenna of the aircraft is used instead of the tail and nose positions.
-This simplification could result in a few seconds difference.
-The ADS-B equipped aircraft at Keflavik airport is estimated around $90\%$ \cite{isavia-rounardeild_rannsoknir_2018}.
-Certain filtering of the ADS-B data is applied before commencing measurements:
-\begin{itemize}
-    \item All records with no latitude, longitude and/or time data are omitted.
-    \item The records are filtered with regards to quality, i.e. filtered with regards to Target Surveillance status, MOPS version, NUC/NIC, NACp, SIL and PIC.
-    \item An ADS-B data record within 0.5 second from another is omitted, i.e. if there are less than 0.5 seconds between two successive locations one of the records is omitted.
-    \item An algorithm is used to analyse the data and determine if it originates from two different flights (the flight stops for 15 minutes or more after/before taxiing).
-If so the data for the second flight is omitted
-    \item The velocity is calculated from ADS-B location data and ADS-B time data (velocity = distance travelled/time).
-The ADS-B velocity record is optional in the ASTERIX category 021 standard and is therefore unreliable in measurements.
-Since the velocity is calculated from measured data, it can exhibit spikes.
-To get a more realistic velocity curve, a Savitzky-Golay filter is applied to smooth the data.
-    \item An aircraft main gear lift-off is considered to be the point where the ADS-B ground bit is removed, i.e. the ADS-B ground bit goes from a value of 1 to a value of 0.
-    \item An aircraft touchdown is considered to be the point where the ADS-B ground bit is set, i.e. the ADS-B ground bit goes from a value of 0 to a value of 1.
-    \item An aircraft is considered to be stationary when the velocity is below 0.5 knots and to be moving if the velocity is greater than 0.5 knots.
-    \cite{isavia_wiki}
-\end{itemize}
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{graphics/Peak_Diagram.png}
+    \caption[Rules defining a peak hour]{The rules used by Isavia to mark a high load interval as peak or not.
+The time separating each two aircraft in a peak cluster (red) is set as less than or equal to 4 minutes.
+The last red aircraft is not counted as part of the peak.
+Clusters are separated by time intervals larger than four minutes.}
+    \label{fig:Peak_Diagram}
+\end{figure}
 
-\fxnote{The ROT measurements have been checked by: comparing them to measurements made by hand.
-Who measures and who checks them and how???}
+\subsubsection{Factors affecting AROT}
+\fxnote{Airfield surface condition, weather conditions, fast exits, season,  , }
 
@@
 % \usepackage{graphicx}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/AROT_hist.png}
-    \caption[AROT histogram]{AROT histogram}
-    \label{fig:AROT_hist}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/AROT_hist_cumul.png}
-    \caption[AROT histogram and cumulative distribution]{AROT histogram and cumulative distribution}
-    \label{fig:AROT_hist_cumul}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/pdf_wtc.png}
-    \caption[PDF for WTC]{PDF for WTC}
-    \label{fig:pdf_wtc}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/pdf_recats.png}
-    \caption[PDF for RACAT]{PDF fro RACAT}
-    \label{fig:pdf_recats}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/each_recat_box.png}
-    \caption[Boxplot of AROT for RECAT categories]{Boxplot of AROT for RECAT categories}
-    \label{fig:each_recat_box}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/each_recat_hist.png}
-    \caption[AROT histogram for each RECAT]{AROT histogram for each RECAT}
-    \label{fig:each_recat_hist}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/each_recat_hist_dist.png}
-    \caption[AROT histogram and pdf for each RECAT]{AROT histogram and pdf for each RECAT}
-    \label{fig:each_recat_hist_dist}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/each_recat_rug_cumdist.png}
-    \caption[AROT rug plot and cumulative pdf for each RECAT]{AROT rug plot and cumulative pdf for each RECAT}
-    \label{fig:each_recat_rug_cumdist}
-\end{figure}
-
 \begin{table}[]
 \centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{l|r|r|r|r|r|r|r|r|}
-\cline{2-9}
-                               & \multicolumn{8}{c|}{AROT}                                                              \\ \hline
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+                          & \multicolumn{1}{c|}{}   & \multicolumn{7}{c|}{AROT [s]} \\ \hline
 \multicolumn{1}{|l|}{RECAT-EU} & count & mean & std & min & 25\% & 50\% & 75\% & max \\ \hline
-\multicolumn{1}{|l|}{CAT-A}    & 3      & 108       & 22        & 88        & 96        & 105       & 119       & 132   \\ \hline
-\multicolumn{1}{|l|}{CAT-B}    & 147    & 86        & 19        & 51        & 70        & 85        & 100       & 139   \\ \hline
-\multicolumn{1}{|l|}{CAT-C}    & 8909   & 80        & 17        & 14        & 67        & 79        & 91        & 156   \\ \hline
-\multicolumn{1}{|l|}{CAT-D}    & 2177   & 78        & 19        & 46        & 64        & 77        & 92        & 155   \\ \hline
-\multicolumn{1}{|l|}{CAT-E}    & 168    & 86        & 24        & 37        & 71        & 85        & 101       & 149   \\ \hline
-\multicolumn{1}{|l|}{CAT-F}    & 96     & 85        & 26        & 43        & 69        & 81        & 97        & 155   \\ \hline
+\multicolumn{1}{|l|}{CAT-A}    & 2     & 119  & 19  & 105 & 112  & 119  & 125  & 132 \\ \hline
+\multicolumn{1}{|l|}{CAT-B}    & 53    & 81   & 20  & 51  & 65   & 77   & 91   & 136 \\ \hline
+\multicolumn{1}{|l|}{CAT-C}    & 3449  & 78   & 17  & 41  & 65   & 75   & 89   & 155 \\ \hline
+\multicolumn{1}{|l|}{CAT-D}    & 1029  & 75   & 18  & 46  & 62   & 71   & 87   & 155 \\ \hline
+\multicolumn{1}{|l|}{CAT-E}    & 98    & 86   & 25  & 42  & 71   & 84   & 101  & 151 \\ \hline
+\multicolumn{1}{|l|}{CAT-F}    & 50    & 83   & 23  & 43  & 69   & 81   & 91   & 155 \\ \hline
 \end{tabular}%
 }
-\caption{AROT statistics for airtrafic mix at KEF by RECAT categories}
-\label{tab:AROT_RECAT_ stats}
+\caption[AROTs for the air traffic mix by RECAT]{AROT statistics for the air traffic mix at KEF by RECAT-EU categories. The count is the number of landings in peak hours since October 2017}
+\label{tab:AROT_RECAT_stats}
+\end{table}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{ll|r|r|r|r|r|r|r|}
+\cline{3-9}
+                             &                           & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
+\multicolumn{1}{|l|}{season} & count                     & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{SUMMER} & \multicolumn{1}{r|}{3741} & 76                        & 17                       & 41                       & 63                        & 72                        & 86                        & 155                      \\ \hline
+\multicolumn{1}{|l|}{WINTER} & \multicolumn{1}{r|}{940}  & 84                        & 20                       & 43                       & 69                        & 84                        & 96                        & 153                      \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by season]{AROT statistics for the air traffic mix at KEF by season. The count is the number of landings in peak hours since October 2017}
+\label{my-label2}
+\end{table}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
+\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{RWY 01} & 974                        & 85                        & 22                       & 42                       & 67                        & 88                        & 99                        & 152                      \\ \hline
+\multicolumn{1}{|l|}{RWY 10} & 852                        & 85                        & 11                       & 61                       & 78                        & 84                        & 91                        & 155                      \\ \hline
+\multicolumn{1}{|l|}{RWY 19} & 1513                       & 67                        & 10                       & 41                       & 61                        & 66                        & 71                        & 155                      \\ \hline
+\multicolumn{1}{|l|}{RWY 28} & 402                        & 68                        & 14                       & 42                       & 60                        & 66                        & 72                        & 155                      \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by runway for the summer]{AROT statistics for the air traffic mix at KEF by runway for the summer of 2018. The count is the number of landings in peak hours.}
+\label{my-label3}
+\end{table}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
+\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{RWY 01} & 286                        & 88                        & 26                       & 43                       & 62                        & 90                        & 106                       & 153                      \\ \hline
+\multicolumn{1}{|l|}{RWY 10} & 333                        & 93                        & 12                       & 70                       & 85                        & 91                        & 99                        & 144                      \\ \hline
+\multicolumn{1}{|l|}{RWY 19} & 252                        & 72                        & 12                       & 45                       & 64                        & 70                        & 75                        & 140                      \\ \hline
+\multicolumn{1}{|l|}{RWY 28} & 69                         & 73                        & 13                       & 49                       & 63                        & 72                        & 79                        & 105                      \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by runway for the winter]{AROT statistics for the air traffic mix at KEF by runway for the winter season.
+The count is the number of landings in peak hours since October 2017}
+\label{my-label4}
+\end{table}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
+\begin{table}[]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+ & \multicolumn{1}{l|}{} & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
+\multicolumn{1}{|l|}{month} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}January} & 95 & 86 & 20 & 54 & 70 & 85 & 96 & 136 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}February} & 80 & 87 & 16 & 52 & 75 & 86 & 93 & 144 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}March} & 215 & 84 & 18 & 47 & 70 & 85 & 95 & 150 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}April} & 222 & 82 & 17 & 45 & 69 & 81 & 93 & 139 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}May} & 387 & 75 & 13 & 42 & 66 & 74 & 82 & 140 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}June} & 649 & 73 & 16 & 44 & 62 & 68 & 80 & 155 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}July} & 731 & 72 & 16 & 43 & 61 & 68 & 77 & 155 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}August} & 727 & 79 & 18 & 42 & 64 & 81 & 91 & 148 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}September} & 636 & 77 & 18 & 41 & 64 & 75 & 88 & 152 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}October} & 611 & 79 & 17 & 43 & 66 & 76 & 91 & 155 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}November} & 221 & 84 & 23 & 43 & 66 & 82 & 98 & 151 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}December} & 107 & 87 & 24 & 45 & 68 & 86 & 99 & 153 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by month]{AROT statistics for the air traffic mix at KEF by month. The count is the number of landings in peak hours since October 2017}
+\label{my-label5}
 \end{table}
@@
 
+subsection{AROT and LTI Objective}
+
+\fxnote{State my hypothesis. Research objectives/research question comes in concluding part of introduction}
+
 %%% Local Variables: 

### 2018-12-16 introduction.tex
aded Data manipulation to Methods

@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{graphics/fig_RECAT_AROTs_boxplot.png}
-    \caption[AROTs boxplot for RECAT categories, all runways]{Arrival Runway Occupancy Times for the different RECAT-EU categories based on data gathered for a period of one year (since October, 2017).
-The box plot shows the AROTs for all runways at BIKF.}
-    \label{fig:RECAT_AROTs_boxplot}
-\end{figure}
-
 The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway \cite{ROT_definition}. \\
@@
 
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
-                          & \multicolumn{1}{c|}{}   & \multicolumn{7}{c|}{AROT [s]} \\ \hline
-\multicolumn{1}{|l|}{RECAT-EU} & count & mean & std & min & 25\% & 50\% & 75\% & max \\ \hline
-\multicolumn{1}{|l|}{CAT-A}    & 2     & 119  & 19  & 105 & 112  & 119  & 125  & 132 \\ \hline
-\multicolumn{1}{|l|}{CAT-B}    & 53    & 81   & 20  & 51  & 65   & 77   & 91   & 136 \\ \hline
-\multicolumn{1}{|l|}{CAT-C}    & 3449  & 78   & 17  & 41  & 65   & 75   & 89   & 155 \\ \hline
-\multicolumn{1}{|l|}{CAT-D}    & 1029  & 75   & 18  & 46  & 62   & 71   & 87   & 155 \\ \hline
-\multicolumn{1}{|l|}{CAT-E}    & 98    & 86   & 25  & 42  & 71   & 84   & 101  & 151 \\ \hline
-\multicolumn{1}{|l|}{CAT-F}    & 50    & 83   & 23  & 43  & 69   & 81   & 91   & 155 \\ \hline
-\end{tabular}%
-}
-\caption[AROTs for the air traffic mix by RECAT]{AROT statistics for the air traffic mix at KEF by RECAT-EU categories. The count is the number of landings in peak hours since October 2017}
-\label{tab:AROT_RECAT_stats}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{ll|r|r|r|r|r|r|r|}
-\cline{3-9}
-                             &                           & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
-\multicolumn{1}{|l|}{season} & count                     & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{SUMMER} & \multicolumn{1}{r|}{3741} & 76                        & 17                       & 41                       & 63                        & 72                        & 86                        & 155                      \\ \hline
-\multicolumn{1}{|l|}{WINTER} & \multicolumn{1}{r|}{940}  & 84                        & 20                       & 43                       & 69                        & 84                        & 96                        & 153                      \\ \hline
-\end{tabular}%
-}
-\caption[AROTs for the air traffic mix by season]{AROT statistics for the air traffic mix at KEF by season. The count is the number of landings in peak hours since October 2017}
-\label{my-label2}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
-                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
-\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{RWY 01} & 974                        & 85                        & 22                       & 42                       & 67                        & 88                        & 99                        & 152                      \\ \hline
-\multicolumn{1}{|l|}{RWY 10} & 852                        & 85                        & 11                       & 61                       & 78                        & 84                        & 91                        & 155                      \\ \hline
-\multicolumn{1}{|l|}{RWY 19} & 1513                       & 67                        & 10                       & 41                       & 61                        & 66                        & 71                        & 155                      \\ \hline
-\multicolumn{1}{|l|}{RWY 28} & 402                        & 68                        & 14                       & 42                       & 60                        & 66                        & 72                        & 155                      \\ \hline
-\end{tabular}%
-}
-\caption[AROTs for the air traffic mix by runway for the summer]{AROT statistics for the air traffic mix at KEF by runway for the summer of 2018. The count is the number of landings in peak hours.}
-\label{my-label3}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
-                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
-\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{RWY 01} & 286                        & 88                        & 26                       & 43                       & 62                        & 90                        & 106                       & 153                      \\ \hline
-\multicolumn{1}{|l|}{RWY 10} & 333                        & 93                        & 12                       & 70                       & 85                        & 91                        & 99                        & 144                      \\ \hline
-\multicolumn{1}{|l|}{RWY 19} & 252                        & 72                        & 12                       & 45                       & 64                        & 70                        & 75                        & 140                      \\ \hline
-\multicolumn{1}{|l|}{RWY 28} & 69                         & 73                        & 13                       & 49                       & 63                        & 72                        & 79                        & 105                      \\ \hline
-\end{tabular}%
-}
-\caption[AROTs for the air traffic mix by runway for the winter]{AROT statistics for the air traffic mix at KEF by runway for the winter season.
-The count is the number of landings in peak hours since October 2017}
-\label{my-label4}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-% \usepackage[table,xcdraw]{xcolor}
-% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
-\begin{table}[]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
- & \multicolumn{1}{l|}{} & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
-\multicolumn{1}{|l|}{month} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\rowcolor[HTML]{DAE8FC} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}January} & 95 & 86 & 20 & 54 & 70 & 85 & 96 & 136 \\ \hline
-\rowcolor[HTML]{DAE8FC} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}February} & 80 & 87 & 16 & 52 & 75 & 86 & 93 & 144 \\ \hline
-\rowcolor[HTML]{DAE8FC} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}March} & 215 & 84 & 18 & 47 & 70 & 85 & 95 & 150 \\ \hline
-\rowcolor[HTML]{DAE8FC} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}April} & 222 & 82 & 17 & 45 & 69 & 81 & 93 & 139 \\ \hline
-\rowcolor[HTML]{FFFC9E} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}May} & 387 & 75 & 13 & 42 & 66 & 74 & 82 & 140 \\ \hline
-\rowcolor[HTML]{FFFC9E} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}June} & 649 & 73 & 16 & 44 & 62 & 68 & 80 & 155 \\ \hline
-\rowcolor[HTML]{FFFC9E} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}July} & 731 & 72 & 16 & 43 & 61 & 68 & 77 & 155 \\ \hline
-\rowcolor[HTML]{FFFC9E} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}August} & 727 & 79 & 18 & 42 & 64 & 81 & 91 & 148 \\ \hline
-\rowcolor[HTML]{FFFC9E} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}September} & 636 & 77 & 18 & 41 & 64 & 75 & 88 & 152 \\ \hline
-\rowcolor[HTML]{FFFC9E} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}October} & 611 & 79 & 17 & 43 & 66 & 76 & 91 & 155 \\ \hline
-\rowcolor[HTML]{DAE8FC} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}November} & 221 & 84 & 23 & 43 & 66 & 82 & 98 & 151 \\ \hline
-\rowcolor[HTML]{DAE8FC} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}December} & 107 & 87 & 24 & 45 & 68 & 86 & 99 & 153 \\ \hline
-\end{tabular}%
-}
-\caption[AROTs for the air traffic mix by month]{AROT statistics for the air traffic mix at KEF by month. The count is the number of landings in peak hours since October 2017}
-\label{my-label5}
-\end{table}
-
 \subsection{Runway Capacity}

### 2018-12-18 introduction.tex
capacity envelope and beginning of hypothesis

@@
 
-\subsection{Runway Occupancy Time}
+\subsection{Runway Occupancy Time\label{sec:ROT}}
 
@@
 
-\subsubsection{Factors affecting AROT}
-\fxnote{Airfield surface condition, weather conditions, fast exits, season,  , }
-
 \subsection{Runway Capacity}
 
-\fxnote{operational procedures for air traffic control at kef, }
-\fxnote{equipment specifications in the tower that might influence the WTC sepatarion decision-talk to Haraldur}
+ICAO generally defines airport capacity as the number of movements per unit of time that can be accepted during different meteorological conditions \cite{airport_capacity_methodology}. 
+Several measures are currently used to estimate the amount of aircraft movements on the runways of an airport in a specified time interval.
+R.
+de Neufville \cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure maximum throughput capacity.
+It indicates the average number of movements that can be performed on the runway system in 1 hour in the presence of continuous demand, while adhering to all the separation requirements imposed by the air traffic management (ATM) system \cite{de_neufville_airport_2013}.
+The method used by Isavia to estimate the throughput capacity at BIKF is the Capacity Envelope that takes into consideration the number of arrivals and departures within a selected time frame.
+This approach allows to measure the throughput capacity and model the maximum possible values or the potential of the airfield.
+A summary of the capacity envelope for a period of one year is shown in Figure \ref{fig:capacity_evnelope} with a 15 minute time interval.
 
-subsection{AROT and LTI Objective}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.8\textwidth]{graphics/fig_Capacity_Envelope_2017-10-04_to_2018-11-30_15min_occurrences_limits.png}
+    \caption[Capacity envelope for BIKF]{Capacity envelope at BIKF since October 2017.
+The measured time interval is 15 minutes and includes both the morning and afternoon peaks.
+The capacity limits of the airfield are also defined and noted on the figure for various sequence modes.
+The mixed modes indicate equal arrivals and departures (Mixed Mode A) or two times more arrivals than departures (Mixed Mode B), the later being a tendency during the afternoon peaks.
+The script producing the plot is on the Isavia´s Víkingaskipið.}  \label{fig:capacity_evnelope}
+\end{figure}
+
+This approach indicates a maximum measured capacity of four arrivals to four departures in a "Mixed Mode A".
+This equilibrium mode is defined as the primary measure for the throughput capacity of BIKF.
+The values are in accord with the defined peak hour intervals in the previous section (\ref{sec:ROT})
+
+There are many aspects of a runway system that can affect the number of aircraft that can land or depart from an airfield and some of those are listed bellow \cite{de_neufville_airport_2013, kim_validation_2010}.
+\begin{itemize}
+    \item Number and geometric layout of the runways.
+    \item State and performance of the ATM system.
+    \item Separation requirements between aircraft pairs imposed by the ATM.
+    \item Weather conditions (visibility, precipitation, cloud ceiling). 
+    \item Wind direction and strength.
+    \item Mix of aircraft using the airport.
+    \item Sequence of movement on each runway (arrivals - departures split)
+    \item Type and location of taxiway exits from the runway.
+    \item Runway occupancy time.
+    \item Controller workload.
+    \item Noise-related constraints and other environmental considerations.
+\end{itemize}
+
+Two of the above mentioned factors are principal in determining airfield capacity, namely the wake vortex separation requirements and runway occupancy \cite{kolos2013influence}.
+Those are in turn influenced by traffic mix, seasonal factors and runway exits, to mention a few.
+The following chapter (\ref{cha:methods}) on Methods  will deal with some of those aspects.
+
+\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur}
+
+\subsection{Runway Occupancy and Landing Time Interval Study Objective}
+The objective of examining the arrival runway occupancy times and the landing time intervals (LTI) is to determine whether the implementation of the RECAT-EU scheme at BIKF under certain conditions, would affect the throughput of the airfield.
+Additionally the analysis could indicate whether limiting element is the runway occupancy or the separation requirements.
+The RECAT scheme suggests a decrease in distance and in time separation of aircraft pairs (\ref{tab:delta_distance_wtc2recat}, \ref{tab:delta_times_wtc2recat})
 

### 2018-12-18 introduction.tex
hypothesis statement

@@
 %% These directions will disappear in other modes.
-\ifdraft{State the objectives of the exercise. Ask yourself:
-  \underline{Why} did I design/create the item? What did I aim to
-  achieve? What is the problem I am trying to solve?  How is my
-  solution interesting or novel?}{}
+% \ifdraft{State the objectives of the exercise. Ask yourself:
+%   \underline{Why} did I design/create the item? What did I aim to
+%   achieve? What is the problem I am trying to solve?  How is my
+%   solution interesting or novel?}{}
 
 Keflavík International Airport (IATA:KEF, ICAO:BIKF) is the major gateway to Iceland currently with two functioning runways.
-There has been a gradual increase of passengers travelling through Keflavik Airport in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017 \cite{isavia_facts_2017}.
+The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South) , RWY-10 (East-West) and RWY-28 (West-East).
+There has been a gradual increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017 \cite{isavia_facts_2017}.
 For 2018 the passenger numbers show a growth of 13.1\% on average (at the time of writing) from the previous year \cite{isavia_pass_statistics_2018}.
-The increase in traffic at Keflavík Airport can encounter constrained capacity of the runway during peak traffic periods.
-A significant constraint of this process is caused by separation regulations of aircraft due to wake-vortices.
+The increase in traffic at Keflavík Airport can encounter constrained capacity of the runways during peak traffic periods.
+A significant constraint of airport capacity is caused by separation regulations of aircraft due to wake-vortices.
 These are specified by rules established by the European Aviation Safety Agency (EASA) for each type of aircraft based primarily on the weight of the aircraft.
-The general flight safety requirements are established by the International Civil Aviation Organisation (ICAO) for maintaining safe distance between aircraft
+The general flight safety requirements are established by the International Civil Aviation Organisation (ICAO) for maintaining safe distance between aircraft.\fxnote{Introduction needs more work, incomplete.}
 
 \section{Background and Literature Review}
-
+The research parameters for the literate review included but were not limited to the following concepts: runway occupancy time, runway capacity, wake vortex separation.
+The project addresses in particular the arrival runway occupancy and its seasonal fluctuations as well as the variations between the different runways at Keflavík International Airport.
 %\ifdraft{Provide background about the subject matter (e.g. %How was morse code developed?  How is it used today?). 
@@
 
-% Please add the following required packages to your document preamble:
-% \usepackage{multirow}
-% \usepackage{graphicx}
-\begin{table}[]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-\multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                                                                                                                   \\ \cline{3-8}
-\multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A                             & CAT-B                                & CAT-C                         & CAT-D                         & CAT-E                         & CAT-F                                                \\ \hline
-                                                        & CAT-A & \cellcolor[HTML]{FD6864}(+0,5 NM) & \cellcolor[HTML]{67FD9A}-2 NM        & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
-                                                        & CAT-B &                                   & \cellcolor[HTML]{67FD9A}-1 NM        &                               & \cellcolor[HTML]{67FD9A}-1 NM &                               & \cellcolor[HTML]{FD6864}+1 NM                        \\ \cline{2-8}
-                                                        & CAT-C &                                   & \cellcolor[HTML]{67FD9A}-1 (-1,5) NM & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
-                                                        & CAT-D &                                   &                                      &                               &                               &                               &                                                      \\ \cline{2-8}
-                                                        & CAT-E &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{67FD9A}{\color[HTML]{000000} -1 NM} \\ \cline{2-8}
-\multirow{-6}{*}{Leader}                                & CAT-F &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{FD6864}(+0,5 NM)                    \\ \hline
-\end{tabular}%
-}
-\caption[Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal) \cite{rooseleer2015recat}}
-\label{tab:delta_distance_wtc2recat}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{multirow}
-% \usepackage{graphicx}
-\begin{table}[]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-\multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                             \\ \cline{3-8}
-\multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A & CAT-B                        & CAT-C & CAT-D                        & CAT-E                        & CAT-F                        \\ \hline
-                                                        & CAT-A &       & \cellcolor[HTML]{67FD9A}-20s &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8}
-                                                        & CAT-B &       &                              &       & \cellcolor[HTML]{67FD9A}-20s &                              & \cellcolor[HTML]{FD6864}+20s \\ \cline{2-8}
-                                                        & CAT-C &       &                              &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8}
-                                                        & CAT-D &       &                              &       &                              &                              &                              \\ \cline{2-8}
-                                                        & CAT-E &       &                              &       &                              &                              & \cellcolor[HTML]{67FD9A}-20s \\ \cline{2-8}
-\multirow{-6}{*}{Leader}                                & CAT-F &       &                              &       &                              &                              & \cellcolor[HTML]{FD6864}+20s \\ \hline
-\end{tabular}
-}
-\caption[Time difference between reference ICAO and RECAT-EU schemes]{Time difference in wake separation on departure between reference ICAO and RECAT-EU schemes(full proposal) \cite{rooseleer2015recat}}
-\label{tab:delta_times_wtc2recat}
-\end{table}
-
-\subsection{Runway Occupancy Time\label{sec:ROT}}
+\section{Runway Occupancy Time\label{sec:ROT}}
 
@@
 The peak traffic hours are characterised by the time intervals when the airport operates at high load.
-High load interval or window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavik Airport.
+High load interval or window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavík Airport.
 The time between flights during the high load interval is set at $\leq$4 minutes.
@@
 The last flight in a high load window that fulfils these requirements is not considered to be a part of the peak as its behaviour is not affected by an instantaneous next flight after it.
-\fxnote{Peak hours need to be explained, Include picture}
 
@@
 The time separating each two aircraft in a peak cluster (red) is set as less than or equal to 4 minutes.
-The last red aircraft is not counted as part of the peak.
-Clusters are separated by time intervals larger than four minutes.}
+The last red aircraft in the peak cluster is not counted as part of the peak.
+Cluster separators are defined as time intervals larger than four minutes.}
     \label{fig:Peak_Diagram}
@@
 
-\subsection{Runway Capacity}
+\section{Runway Capacity}
 
@@
 The mixed modes indicate equal arrivals and departures (Mixed Mode A) or two times more arrivals than departures (Mixed Mode B), the later being a tendency during the afternoon peaks.
-The script producing the plot is on the Isavia´s Víkingaskipið.}  \label{fig:capacity_evnelope}
+The script producing this plot is on the Isavia´s internal GUI Víkingaskipið.}  \label{fig:capacity_evnelope}
 \end{figure}
@@
 Those are in turn influenced by traffic mix, seasonal factors and runway exits, to mention a few.
-The following chapter (\ref{cha:methods}) on Methods  will deal with some of those aspects.
+The following chapter \ref{cha:methods} on Methods  will deal with some of those aspects.
 
-\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur}
+\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur, Talk to Hjalti about the 2,5 - 3 NM MRS issue!}
 
-\subsection{Runway Occupancy and Landing Time Interval Study Objective}
-The objective of examining the arrival runway occupancy times and the landing time intervals (LTI) is to determine whether the implementation of the RECAT-EU scheme at BIKF under certain conditions, would affect the throughput of the airfield.
-Additionally the analysis could indicate whether limiting element is the runway occupancy or the separation requirements.
-The RECAT scheme suggests a decrease in distance and in time separation of aircraft pairs (\ref{tab:delta_distance_wtc2recat}, \ref{tab:delta_times_wtc2recat})
+\section{Runway Occupancy and Landing Time Interval Study Objective}
+The goal of examining the arrival runway occupancy times and the landing time intervals (LTI) is to determine whether the implementation of the RECAT-EU scheme at BIKF under certain conditions, would affect the throughput of the airfield.
+Here the LTI is a directly influenced by the distance separation of arrival pairs and the final approach speed of the aircraft.
+Additionally the analysis could indicate whether the limiting element is the runway occupancy or the separation requirement.
+The transition from ICAO WTS to RECAT-EU scheme suggests a decrease in distance separation for a number of the aircraft pairs (Tables \ref{tab:delta_distance_wtc2recat}).
 
+% Please add the following required packages to your document preamble:
+% \usepackage{multirow}
+% \usepackage{graphicx}
+\begin{table}[]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
+\hline
+\multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                                                                                                                   \\ \cline{3-8}
+\multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A                             & CAT-B                                & CAT-C                         & CAT-D                         & CAT-E                         & CAT-F                                                \\ \hline
+                                                        & CAT-A & \cellcolor[HTML]{FD6864}(+0,5 NM) & \cellcolor[HTML]{67FD9A}-2 NM        & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
+                                                        & CAT-B &                                   & \cellcolor[HTML]{67FD9A}-1 NM        &                               & \cellcolor[HTML]{67FD9A}-1 NM &                               & \cellcolor[HTML]{FD6864}+1 NM                        \\ \cline{2-8}
+                                                        & CAT-C &                                   & \cellcolor[HTML]{67FD9A}-1 (-1,5) NM & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
+                                                        & CAT-D &                                   &                                      &                               &                               &                               &                                                      \\ \cline{2-8}
+                                                        & CAT-E &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{67FD9A}{\color[HTML]{000000} -1 NM} \\ \cline{2-8}
+\multirow{-6}{*}{Leader}                                & CAT-F &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{FD6864}(+0,5 NM)                    \\ \hline
+\end{tabular}%
+}
+\caption[Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal) \cite{rooseleer2015recat}}
+\label{tab:delta_distance_wtc2recat}
+\end{table}
+
+% % Please add the following required packages to your document preamble:
+% % \usepackage{multirow}
+% % \usepackage{graphicx}
+% \begin{table}[]
+% \centering
+% \resizebox{\textwidth}{!}{%
+% \begin{tabular}{|c|c|c|c|c|c|c|c|}
+% \hline
+% \multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                             \\ \cline{3-8}
+% \multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A & CAT-B                        & CAT-C & CAT-D                        & CAT-E                        & CAT-F                        \\ \hline
+%                                                         & CAT-A &       & \cellcolor[HTML]{67FD9A}-20s &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8}
+%                                                         & CAT-B &       &                              &       & \cellcolor[HTML]{67FD9A}-20s &                              & \cellcolor[HTML]{FD6864}+20s \\ \cline{2-8}
+%                                                         & CAT-C &       &                              &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8}
+%                                                         & CAT-D &       &                              &       &                              &                              &                              \\ \cline{2-8}
+%                                                         & CAT-E &       &                              &       &                              &                              & \cellcolor[HTML]{67FD9A}-20s \\ \cline{2-8}
+% \multirow{-6}{*}{Leader}                                & CAT-F &       &                              &       &                              &                              & \cellcolor[HTML]{FD6864}+20s \\ \hline
+% \end{tabular}
+% }
+% \caption[Time difference between reference ICAO and RECAT-EU schemes]{Time difference in wake separation on departure between reference ICAO and RECAT-EU schemes(full proposal) \cite{rooseleer2015recat}}
+% \label{tab:delta_times_wtc2recat}
+% \end{table}
+
+The implementation of RECAT shifts the reference separation minima for arrival pairs thus creating the potential for reduced time separation between the aircraft.
+The reduction of time interval between landings leads to increased throughput of the runways.
 \fxnote{State my hypothesis. Research objectives/research question comes in concluding part of introduction}

### 2018-12-18 introduction.tex
cosmetic changes

@@
 The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South) , RWY-10 (East-West) and RWY-28 (West-East).
-There has been a gradual increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017 \cite{isavia_facts_2017}.
-For 2018 the passenger numbers show a growth of 13.1\% on average (at the time of writing) from the previous year \cite{isavia_pass_statistics_2018}.
+There has been a gradual increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017~\cite{isavia_facts_2017}.
+For 2018 the passenger numbers show a growth of 13.1\% on average (at the time of writing) from the previous year~\cite{isavia_pass_statistics_2018}.
 The increase in traffic at Keflavík Airport can encounter constrained capacity of the runways during peak traffic periods.
@@
 
-\subsection{Current understandings of wake vortex behaviour}
-The wing of an aircraft can be regarded as a beam of finite length with aerofoil as cross section \cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation or drop it if unnecessary.
+\subsection{Current Understandings of Wake Vortex Behaviour}
+The wing of an aircraft can be regarded as a beam of finite length with aerofoil as cross section~\cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation or drop it if unnecessary.
 Read about induced vertical flow in a book on aerodynamics}.
@@
 Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards.
-A finite lifting wing leaves behind a continuous sheet of stream-wise vorticity  \cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation }.
-This manifests itself in multiple trailing swirls of the airflow that mix together and roll up into two main counter-rotating vortices aft of the wing as illustrated in Figure \ref{fig:vortex_develop} \cite{magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
+A finite lifting wing leaves behind a continuous sheet of stream-wise vorticity ~\cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation }.
+This manifests itself in multiple trailing swirls of the airflow that mix together and roll up into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop}~\cite{magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
 Between the vortices the air moves downwards while outside the induced flow is upwards.
@@
 The cross flow velocities in the core region of the trailing vortices can reach $360$ km/h and the vortices can stay effective up to hundred wing spans.
-This can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft \cite{Breitsamter2011Feb, gerz_commercial_2002}.
+This can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft~\cite{Breitsamter2011Feb, gerz_commercial_2002}.
 
@@
     \caption[Wake vortex roll-up process]{Wing vortex evolution and roll-up process.
-Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one \cite[p.
+Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one~\cite[p.
 043]{magazine_aibus_safety}} \label{fig:vortex_develop}
@@
 
-\subsection{Ground effect and decay process}
+\subsection{Ground Effect and Decay Process}
 
@@
 In the presence of stable cross-wind conditions the decay of the downwind vortex is not identical to the upwind vortex and the descent rate may vary significantly \cite{Hallock2018Apr}. 
-The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure \ref{fig:vortex_ground_effect}).
-Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$ m and $518$ m and in some cases even $762$ m, which is the basis for the "2500-foot"  rule separation in parallel runways configuration \cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
-The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5$ m/s for the first $30$ s after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$ m and below of the flight path  \cite{lissaman1973aircraft, Hallock2018Apr}.
-Aerodynamic properties of the aircraft govern the vortex roll-up process and ambient atmospheric conditions dominate the behaviour of the vortices forcing instability and eventual decay of the vortices \cite{Hallock2018Apr}.
+The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure~\ref{fig:vortex_ground_effect}).
+Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$~m and $518$~m and in some cases even $762$~m, which is the basis for the "2500-foot"  rule separation in parallel runways configuration~\cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
+The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5$~m/s for the first $30$~s after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m and below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}.
+Aerodynamic properties of the aircraft govern the vortex roll-up process and ambient atmospheric conditions dominate the behaviour of the vortices forcing instability and eventual decay of the vortices~\cite{Hallock2018Apr}.
 Several factors that drive the decay rate of the vortex pair have been formulated:
 \begin{itemize}
-    \item Atmospheric turbulence extracts energy from the vortex and reduces its strength leading to a faster wake decay \cite{Hallock2018Apr}.
-    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence \cite{Hallock2018Apr}.\fxnote{This needs to be explained (Ármann)}
-    \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level \cite{Holzapfel2001Feb, gerz_commercial_2002}.
-    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may cause the vortices to link and decay faster \cite{Hallock2018Apr, crow2003stability}.
-    \item Secondary vorticity structures, vertical rib-like counter rotating flow formations may appear between the vortex pair and eventually wrap around the main vortices, leading to turbulence build inside the system and rapid circulation decay\cite{Holzapfel2001Feb, Holzapfel2003Jun}.
+    \item Atmospheric turbulence extracts energy from the vortex and reduces its strength leading to a faster wake decay ~\cite{Hallock2018Apr}.
+    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence~\cite{Hallock2018Apr}.\fxnote{This needs to be explained better (Ármann)}
+    \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level~\cite{Holzapfel2001Feb, gerz_commercial_2002}.
+    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may cause the vortices to link and decay faster~\cite{Hallock2018Apr, crow2003stability}.
+    \item Secondary vorticity structures, vertical rib-like counter rotating flow formations may appear between the vortex pair and eventually wrap around the main vortices, leading to turbulence build inside the system and rapid circulation decay~\cite{Holzapfel2001Feb, Holzapfel2003Jun}.
 \end{itemize}
@@
     \includegraphics[width=0.8\textwidth]{graphics/Hallock_vortex_evolution.jpg}
-    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect \cite[p. 29]{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
+    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect~\cite[p. 29]{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
 \end{figure}
 
-\subsection{Wake vortex separation}
-An aircraft affected by the wake vortex can experience loss of lift or an induced rolling moment and velocity fluctuations (Figure \ref{fig:vortex_encounter}).
+\subsection{Wake Vortex Separation}
+An aircraft affected by the wake vortex can experience loss of lift or an induced rolling moment and velocity fluctuations (Figure~\ref{fig:vortex_encounter}).
 This can lead to the up-set and potentially loss of control of an aircraft following the flight path of a preceding aircraft.
-To diminish the effect of such vortices the trailing aircraft must be maintained at a safe distance behind the lead aircraft as the vortices spread laterally to either side of the flight path and are dissipated \cite{Breitsamter2011Feb}.
+To diminish the effect of such vortices the trailing aircraft must be maintained at a safe distance behind the leading aircraft as the vortices spread laterally to either side of the flight path and are dissipated~\cite{Breitsamter2011Feb}.
 The distance between aircraft pairs in view of safety is referred to as wake vortex separation.
@@
     \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft.
-(A) induced rolling moment, (B) upward motion, (C) loss of lift  \cite[p.
+(A) induced rolling moment, (B) upward motion, (C) loss of lift~\cite[p.
 33]{Hallock2018Apr}} \label{fig:vortex_encounter}
@@
 
-Traditional separation standards, introduced in the seventies, are defined by ICAO  on the basis of certificated maximum take-off mass (MTOM) as shown in Table \ref{tab:WTC}.
+Traditional separation standards, introduced in the seventies, are defined by ICAO  on the basis of certificated maximum take-off mass (MTOM) as shown in Table~\ref{tab:WTC}.
 The prescribed categories are three (i.e.Heavy, Medium and Light).
-In addition for aircraft in the order of $560000$ kg ICAO provides a subcategory to the Heavy types called Super Heavy.
+In addition for aircraft in the order of $560000$~kg ICAO provides a subcategory to the Heavy types called Super Heavy.
 Some states have chosen to further increase the number of categories to adapt the wake turbulence categories (WTC) to specific airport requirements.
-Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table \ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing B757 into a separate category and increases the upper weight limit of the Light category.
-\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
+Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing B757 into a separate category and increases the upper weight limit of the Light category.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
 
@@
     \end{tabular}}
-    \caption[ICAO wake turbulence categories based on maximum take-off mass]{ICAO wake turbulence categories based on maximum take-off mass (MTOM) in kg.
-\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, kolos2013influence}} \label{tab:WTC}
+    \caption[ICAO wake turbulence categories based on maximum take-off mass]{ICAO wake turbulence categories based on maximum take-off mass (MTOM) in kg~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, kolos2013influence}.} \label{tab:WTC}
 \end{table}
 
-The ICAO wake turbulence categories outline the worst case for each category and thus generate over-separation in many cases \cite{noauthor_recat_2018}.
+The ICAO wake turbulence categories outline the worst case for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018}.
 The separation distances for instrumental flight rules are observed for the follower aircraft and are conventionally expressed in nautical miles (NM).
 The separation reverts to radar separation minimum (MRS) when wake turbulence restrictions are not required.
-As  prescribed  by  ICAO  the  horizontal MRS is 3 NM (or a reduced separation minimum of 2.5 NM may be applied under given conditions described in ICAO Doc 4444 PANS-AM \cite{doc44444}), which translates to 79s (66s) time separation between leader and follower at approach speed of 70 m/s.
+As  prescribed  by  ICAO  the  horizontal MRS is 3~NM (or a reduced separation minimum of 2.5~NM may be applied under given conditions described in ICAO Doc 4444 PANS-AM~\cite{doc44444}), which translates to 79~s (66~s) time separation between leader and follower at final approach speed of 70~m/s.
 
@@
 }
-\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter.
-\cite{noauthor_recat_2018, rooseleer2015recat}} \label{tab:ICAO_WTC}
+\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter~\cite{noauthor_recat_2018, rooseleer2015recat}.} \label{tab:ICAO_WTC}
 \end{table}
+
+It must be noted that the division of WTC categories into a RECAT scheme is not achieved by splitting each category in half.
+The criteria used for categorisation of existing and new aircraft types are provided in detail in~\cite{noauthor_recat_2018}.
+The case for the traffic mix at BIKF presents the arrangement shown in Table~\ref{tab:wtc2recat_division}.
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
+\begin{table}[]
+\centering
+\resizebox{0.7\textwidth}{!}{%
+\begin{tabular}{clcllc}
+\multicolumn{3}{c}{\cellcolor[HTML]{34CDF9}HEAVY} &  &  & \cellcolor[HTML]{FD6864}LIGHT \\ \hline
+\multicolumn{1}{|c|}{\cellcolor[HTML]{34CDF9}CAT-A} & \multicolumn{1}{c|}{\cellcolor[HTML]{34CDF9}CAT-B} & \multicolumn{1}{c|}{\cellcolor[HTML]{32CB00}CAT-C} & \multicolumn{1}{c|}{\cellcolor[HTML]{F8FF00}CAT-D} & \multicolumn{1}{c|}{\cellcolor[HTML]{F8FF00}CAT-E} & \multicolumn{1}{c|}{\cellcolor[HTML]{FFC702}CAT-F} \\ \hline
+\multicolumn{1}{l}{} &  & \multicolumn{4}{c}{\cellcolor[HTML]{F8FF00}MEDIUM}
+\end{tabular}%
+}
+\caption{Transition from ICAO WTC to RECAT-EU categories.
+The categorisation process and criteria for assigning an existing aircraft type into RECAT-EU scheme is illustrated in detail in~\cite{noauthor_recat_2018}}
+\label{tab:wtc2recat_division}
+\end{table}
+
+\fxnote{explain the transition from WTC to REACT , and point the case for BIKF: H to ABC, M to CDEF, ....}
 
@@
 }
-\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure \cite{rooseleer2015recat}}
+\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure~\cite{rooseleer2015recat}}
 \label{tab:RECAT-time}
@@
 
-The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway \cite{ROT_definition}. \\
-This project focuses on Arrivals Runway Occupancy Time (AROT) in peak traffic hours, that is the time interval between the aircraft crossing the threshold and its tail vacating the runway \cite{AROT_definition}.
+The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway~\cite{ROT_definition}. \\
+This project focuses on Arrivals Runway Occupancy Time (AROT) in peak traffic hours, that is the time interval between the aircraft crossing the threshold and its tail vacating the runway~\cite{AROT_definition}.
 The threshold is the beginning of that portion of the runway that is available for landing.
@@
 High load interval or window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavík Airport.
-The time between flights during the high load interval is set at $\leq$4 minutes.
+The time between flights during the high load interval is set at $\leq$4~minutes.
 This value is based on statistical data by Isavia in order to achieve two distinct peak hours: one in the morning and one in the afternoon.
@@
     \includegraphics[width=1\textwidth]{graphics/Peak_Diagram.png}
-    \caption[Rules defining a peak hour]{The rules used by Isavia to mark a high load interval as peak or not.
+    \caption[Rules defining a peak hour]{Rules used by Isavia to identify a high load interval as peak or not.
 The time separating each two aircraft in a peak cluster (red) is set as less than or equal to 4 minutes.
@@
 
-ICAO generally defines airport capacity as the number of movements per unit of time that can be accepted during different meteorological conditions \cite{airport_capacity_methodology}. 
+ICAO generally defines airport capacity as the number of movements per unit of time that can be accepted during different meteorological conditions~\cite{airport_capacity_methodology}. 
 Several measures are currently used to estimate the amount of aircraft movements on the runways of an airport in a specified time interval.
 R.
-de Neufville \cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure maximum throughput capacity.
-It indicates the average number of movements that can be performed on the runway system in 1 hour in the presence of continuous demand, while adhering to all the separation requirements imposed by the air traffic management (ATM) system \cite{de_neufville_airport_2013}.
-The method used by Isavia to estimate the throughput capacity at BIKF is the Capacity Envelope that takes into consideration the number of arrivals and departures within a selected time frame.
+de Neufville~\cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure maximum throughput capacity.
+It indicates the average number of movements that can be performed on the runway system in one hour in the presence of continuous demand, while adhering to all the separation requirements imposed by the air traffic management (ATM) system~\cite{de_neufville_airport_2013}.
+The Capacity Envelope method used by Isavia to estimate the throughput capacity at BIKF takes into consideration the number of arrivals and departures within a selected time frame.
 This approach allows to measure the throughput capacity and model the maximum possible values or the potential of the airfield.
-A summary of the capacity envelope for a period of one year is shown in Figure \ref{fig:capacity_evnelope} with a 15 minute time interval.
+A summary of the capacity envelope for a period of one year is shown in Figure~\ref{fig:capacity_evnelope} for a 15~minute time interval.
 
@@
 The capacity limits of the airfield are also defined and noted on the figure for various sequence modes.
-The mixed modes indicate equal arrivals and departures (Mixed Mode A) or two times more arrivals than departures (Mixed Mode B), the later being a tendency during the afternoon peaks.
+The mixed modes indicate equal arrivals and departures (Mixed~Mode~A) or two times more arrivals than departures (Mixed Mode B), the later being a tendency during the afternoon peaks.
 The script producing this plot is on the Isavia´s internal GUI Víkingaskipið.}  \label{fig:capacity_evnelope}
@@
 
-This approach indicates a maximum measured capacity of four arrivals to four departures in a "Mixed Mode A".
+This approach indicates a maximum measured capacity of four arrivals to four departures in a "Mixed~Mode~A".
 This equilibrium mode is defined as the primary measure for the throughput capacity of BIKF.
-The values are in accord with the defined peak hour intervals in the previous section (\ref{sec:ROT})
+The values are in accord with the defined peak hour intervals in the previous section~\ref{sec:ROT}
 
-There are many aspects of a runway system that can affect the number of aircraft that can land or depart from an airfield and some of those are listed bellow \cite{de_neufville_airport_2013, kim_validation_2010}.
+There are many aspects of a runway system that can affect the number of aircraft that can land or depart from an airfield and some of those are listed bellow~\cite{de_neufville_airport_2013, kim_validation_2010}.
 \begin{itemize}
@@
 
-Two of the above mentioned factors are principal in determining airfield capacity, namely the wake vortex separation requirements and runway occupancy \cite{kolos2013influence}.
+Two of the above mentioned factors are principal in determining airfield capacity, namely the wake vortex separation requirements and runway occupancy~\cite{kolos2013influence}.
 Those are in turn influenced by traffic mix, seasonal factors and runway exits, to mention a few.
-The following chapter \ref{cha:methods} on Methods  will deal with some of those aspects.
+The following chapter~\ref{cha:methods} on Methods  will deal with some of those aspects.
 
-\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur, Talk to Hjalti about the 2,5 - 3 NM MRS issue!}
+\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur, Talk to Hjalti about the 2,5~-~3~NM MRS issue!}
 
@@
 The goal of examining the arrival runway occupancy times and the landing time intervals (LTI) is to determine whether the implementation of the RECAT-EU scheme at BIKF under certain conditions, would affect the throughput of the airfield.
-Here the LTI is a directly influenced by the distance separation of arrival pairs and the final approach speed of the aircraft.
+LTI is the a measure of the time separation between aircraft pairs derived from the distance separation between the aircraft and the final approach speed (time travelled is distance travelled divided by velocity).
+% The LTI is directly influenced by the distance separation of arrival pairs and the final approach speed of the aircraft .
 Additionally the analysis could indicate whether the limiting element is the runway occupancy or the separation requirement.
-The transition from ICAO WTS to RECAT-EU scheme suggests a decrease in distance separation for a number of the aircraft pairs (Tables \ref{tab:delta_distance_wtc2recat}).
+The proposed transition by Eurocontrol from ICAO WTS to RECAT-EU scheme suggests a decrease in distance separation for a number of the aircraft pairs (Table~\ref{tab:delta_distance_wtc2recat}).
 
@@
 }
-\caption[Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal) \cite{rooseleer2015recat}}
+\caption[Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)~\cite{rooseleer2015recat}}
 \label{tab:delta_distance_wtc2recat}
@@
 
-The implementation of RECAT shifts the reference separation minima for arrival pairs thus creating the potential for reduced time separation between the aircraft.
-The reduction of time interval between landings leads to increased throughput of the runways.
-\fxnote{State my hypothesis. Research objectives/research question comes in concluding part of introduction}
+The implementation of RECAT shifts the wake separation minima for arrival pairs thus creating the potential for reduced time separation between the aircraft.
+The reduction of the time interval between landings implies increased throughput of the runways.
+Figure~\ref{fig:AROT_LTI_rwy19_H_M} may be used to illustrate the hypothesis.
+The current ICAO time separation reference, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT scheme.
+This relocation will be more noticeable for RECAT C-C and C-D pairs formed from the ICAO HEAVY-MEDIUM pairs, in which case the separation is reduced from 5 NM to 3 NM.
+A shift of the reference separation line creates the potential for shift in the distribution of the LTI, as long as it does not overlap with the ROT.
+Furthermore study on the runway capacity of significantly larger airports suggests that the frequency distribution of LTI tends to compress or squeeze to the right of the reference line with less standard deviation about the mean when the air traffic is intensified \cite{kolos2013influence}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.8\textwidth]{graphics/fig_rot_landig_time_interval_RWY19_leader_H_follower_M_peak-hour_BAR_20171004_20181130.png}
+    \caption[Caption table]{Arrival runway occupancy time (blue) and landing time intervals (green) for Heavy leader - Medium follower pairs on BIKF runway RWY-19.
+The red line indicates the ICAO WTC reference time separation for the selected H-M aircraft pair.}\label{fig:AROT_LTI_rwy19_H_M}
+\end{figure}
 

### 2018-12-19 code.tex
corrected quantiles on arot to exclude low value ouliers.
work on traffic mix, wtc to recat transition.

@@
+\chapter{Runway Usage}\label{cha:RWY_use}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=0.8\textwidth]{graphics/fig_runway_usage_summer}
+    \caption[Summer runway usage at BIKF]{The summer traffic favours RWY-19 (52,8\%), followed by RWY-01 (23,3\%), while RWY-10 and RWY-28 are with 11,6\% and 12,3\% respectively.}
+    \label{fig:runway_usage_summer}
+\end{figure}
+
+\begin{figure}[]
+    \centering
+    \includegraphics[width=0.8\textwidth]{graphics/fig_runway_usage_winter}
+    \caption[Winter runway usage at BIKF]{The winter traffic at BIKF is a bit more evenly distributed among runways bit still favours RWY-19 (38,8\%),followed by RWY-01 with 28,9\%, RWY-10 (20,8\%) and RWY-28 (11,4\%).}
+    \label{fig:runway_usage_winter}
+\end{figure}
+
 \chapter{Arrival Runway Occupancy Times}\label{cha:AROTs}
@@
 \multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{768} & 85 & 20 & 44 & 69 & 85 & 99 & 153 \\ \hline
+\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{764} & 85 & 20 & 46 & 70 & 86 & 99 & 153 \\ \hline
 \multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{263} & 100 & 17 & 55 & 87 & 98 & 111 & 156 \\ \hline
@@
 \multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{1900} & 83 & 22 & 41 & 63 & 86 & 98 & 152 \\ \hline
-\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{286} & 88 & 26 & 43 & 62 & 90 & 106 & 153 \\ \hline
+\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{1879} & 84 & 22 & 46 & 64 & 86 & 98 & 152 \\ \hline
+\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{284} & 88 & 26 & 47 & 63 & 90 & 106 & 153 \\ \hline
 \end{tabular}%
@@
 \multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{554} & 99 & 13 & 37 & 91 & 98 & 106 & 153 \\ \hline
-\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{250} & 109 & 14 & 37 & 100 & 108 & 117 & 155 \\ \hline
+\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{553} & 99 & 12 & 46 & 91 & 98 & 106 & 153 \\ \hline
+\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{249} & 110 & 13 & 60 & 100 & 108 & 117 & 155 \\ \hline
 \end{tabular}%
@@
 \multicolumn{1}{|l|}{Season} & {count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & 402 & 68 & 14 & 42 & 60 & 66 & 72 & 155 \\ \hline
+\multicolumn{1}{|l|}{Summer} & 401 & 68 & 14 & 49 & 60 & 66 & 72 & 155 \\ \hline
 \multicolumn{1}{|l|}{Winter} & 69 & 73 & 13 & 49 & 63 & 72 & 79 & 105 \\ \hline

### 2018-12-19 introduction.tex
work on traffic mix, wtc to recat transition.

@@
 
-It must be noted that the division of WTC categories into a RECAT scheme is not achieved by splitting each category in half.
+It must be noted that the division of ICAO WTC categories into the RECAT scheme is not achieved by splitting each category in half.
 The criteria used for categorisation of existing and new aircraft types are provided in detail in~\cite{noauthor_recat_2018}.
-The case for the traffic mix at BIKF presents the arrangement shown in Table~\ref{tab:wtc2recat_division}.
+The case from the traffic mix at BIKF presents the arrangement shown in Table~\ref{tab:wtc2recat_division}.
+A more detailed illustration of the re-categorisation is presented in the coming section \ref{ssec:traffic_mix} on aircraft traffic mix (Figure~\ref{fig:post_fast_exit_mix_pie_v2}).
 
@@
 
-\fxnote{explain the transition from WTC to REACT , and point the case for BIKF: H to ABC, M to CDEF, ....}
+\fxnote{explain the transition from WTC to REACT , and point the case for BIKF: H to ABC, M to CDEF, ....reference the pie chart on mix}
 
@@
 
-\section{Runway Capacity}
+\section{Runway Capacity\label{sec:runway_capacity}}
 
@@
 The implementation of RECAT shifts the wake separation minima for arrival pairs thus creating the potential for reduced time separation between the aircraft.
-The reduction of the time interval between landings implies increased throughput of the runways.
+The reduction of the time interval between landings facilitates increas in the throughput capacity of the runways.
 Figure~\ref{fig:AROT_LTI_rwy19_H_M} may be used to illustrate the hypothesis.
@@
 A shift of the reference separation line creates the potential for shift in the distribution of the LTI, as long as it does not overlap with the ROT.
-Furthermore study on the runway capacity of significantly larger airports suggests that the frequency distribution of LTI tends to compress or squeeze to the right of the reference line with less standard deviation about the mean when the air traffic is intensified \cite{kolos2013influence}.
+Furthermore study on the runway capacity of significantly larger airports~\cite{kolos2013influence} suggests that the frequency distribution of LTI tends to compress or squeeze to the right of the reference line with less standard deviation about the mean when the air traffic is intensified.
 
-\begin{figure}
+\begin{figure}[ht]
     \centering
@@
     \caption[Caption table]{Arrival runway occupancy time (blue) and landing time intervals (green) for Heavy leader - Medium follower pairs on BIKF runway RWY-19.
+The observed time period is from October 2017 to November 2018.
 The red line indicates the ICAO WTC reference time separation for the selected H-M aircraft pair.}\label{fig:AROT_LTI_rwy19_H_M}

### 2018-12-19 code.tex
finished writing on Traffic mix

@@
-\chapter{Runway Usage}\label{cha:RWY_use}
+\chapter{Runway Usage}\label{cha:RWY_usage}
 
-\begin{figure}[]
+\begin{figure}[h]
     \centering
-    \includegraphics[width=0.8\textwidth]{graphics/fig_runway_usage_summer}
+    \includegraphics[width=0.8\textwidth]{graphics/fig_runway_usage_2017-10-04_to_2018-11-30.png}
+    \caption[Runway usage at BIKF]{Overall runway usage at BIKF for a period of one year since October 2017.
+Almost half of all arrivals during that period (44,7\%) use RWY-19, followed by RWY-01 (25,7\%) and RWY-10 (17.6\%).
+The summer traffic favours RWY-19 with 52.8\%~(Figure~\ref{fig:runway_usage_summer}), while during the winter season some of those arrivals are diverted towards RWY-01 and RWY-10~(Figure~\ref{fig:runway_usage_summer}).
+RWY-28 is least used despite its fast-exit TWY B-1.}
+    \label{fig:runway_usage}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.7\textwidth]{graphics/fig_runway_usage_summer}
     \caption[Summer runway usage at BIKF]{The summer traffic favours RWY-19 (52,8\%), followed by RWY-01 (23,3\%), while RWY-10 and RWY-28 are with 11,6\% and 12,3\% respectively.}
@@
 
-\begin{figure}[]
+\begin{figure}[h]
     \centering
-    \includegraphics[width=0.8\textwidth]{graphics/fig_runway_usage_winter}
+    \includegraphics[width=0.7\textwidth]{graphics/fig_runway_usage_winter}
     \caption[Winter runway usage at BIKF]{The winter traffic at BIKF is a bit more evenly distributed among runways bit still favours RWY-19 (38,8\%),followed by RWY-01 with 28,9\%, RWY-10 (20,8\%) and RWY-28 (11,4\%).}
@@
 
-\chapter{Arrival Runway Occupancy Times}\label{cha:AROTs}
+\chapter{Arrival Runway Occupancy Times\label{cha:AROTs}}
 
@@
 % \usepackage{graphicx}
-\begin{table}[]
+\begin{table}[h]
 \centering
@@
 
-\chapter{Code}\label{cha:code}
-You can put code in your document using the listings package, which is
-loaded by default in \path{custom.tex}.  Be aware that the listings
-package does not put code in your document if you are in draft mode
-unless you set the \texttt{forcegraphics} option.
+% \chapter{Code}\label{cha:code}
+% You can put code in your document using the listings package, which is
+% loaded by default in \path{custom.tex}.  Be aware that the listings
+% package does not put code in your document if you are in draft mode
+% unless you set the \texttt{forcegraphics} option.
 
-There is an example java (Listing~\ref{src:Data_Bus.java}) and XML
-file (Listing~\ref{src:AndroidManifest.xml}).  Thanks to the
-\texttt{url} package, you can typeset OSX and unix paths like this:
-\path{/afs/rnd.ru.is/project/thesis-template}.  Windows paths:
-\path{C:\windows\temp\ }.  You can also typeset them using the menukey
-package, but it tends to delete the last separator and has other
-complications.\footnote{The menukey package has issues with biblatex,
-  read \path{custom.tex} for more information.}
+% There is an example java (Listing~\ref{src:Data_Bus.java}) and XML
+% file (Listing~\ref{src:AndroidManifest.xml}).  Thanks to the
+% \texttt{url} package, you can typeset OSX and unix paths like this:
+% \path{/afs/rnd.ru.is/project/thesis-template}.  Windows paths:
+% \path{C:\windows\temp\ }.  You can also typeset them using the menukey
+% package, but it tends to delete the last separator and has other
+% complications.\footnote{The menukey package has issues with biblatex,
+%   read \path{custom.tex} for more information.}
 
-If you are trying to include multiple different languages, you should
-go read the documentation and set these up in \path{custom.tex}.  You
-will save yourself a lot of effort, especially if you have to fix
-anything.
+% If you are trying to include multiple different languages, you should
+% go read the documentation and set these up in \path{custom.tex}.  You
+% will save yourself a lot of effort, especially if you have to fix
+% anything.
 
-%I have put the source code in the \directory{src/} folder.
-\lstinputlisting[language=Java, firstline=1,
-lastline=40, caption={Data\_Bus.java: Setting up the class.},
-label={src:Data_Bus.java}]{src/Data_Bus.java}
+% %I have put the source code in the \directory{src/} folder.
+% \lstinputlisting[language=Java, firstline=1,
+% lastline=40, caption={Data\_Bus.java: Setting up the class.},
+% label={src:Data_Bus.java}]{src/Data_Bus.java}
 
-\lstinputlisting[language={[android]XML}, firstline=1, lastline=20,
-caption={AndroidManifest.xml: Configuration for the Android UI.},
-label={src:AndroidManifest.xml}]{src/AndroidManifest.xml}
+% \lstinputlisting[language={[android]XML}, firstline=1, lastline=20,
+% caption={AndroidManifest.xml: Configuration for the Android UI.},
+% label={src:AndroidManifest.xml}]{src/AndroidManifest.xml}
 

### 2018-12-20 code.tex
fixed title fonts on figures, sorted boxplot, BIKF schematic.

@@
 \chapter{Arrival Runway Occupancy Times\label{cha:AROTs}}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[h]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+                          & \multicolumn{1}{c|}{}   & \multicolumn{7}{c|}{AROT [s]} \\ \hline
+\multicolumn{1}{|l|}{RECAT-EU} & count & mean & std & min & 25\% & 50\% & 75\% & max \\ \hline
+\multicolumn{1}{|l|}{CAT-A}    & 2     & 119  & 19  & 105 & 112  & 119  & 125  & 132 \\ \hline
+\multicolumn{1}{|l|}{CAT-B}    & 53    & 81   & 20  & 51  & 65   & 77   & 91   & 136 \\ \hline
+\multicolumn{1}{|l|}{CAT-C}    & 3436  & 78   & 17  & 46  & 65   & 75   & 89   & 155 \\ \hline
+\multicolumn{1}{|l|}{CAT-D}    & 1029  & 75   & 18  & 46  & 62   & 71   & 87   & 155 \\ \hline
+\multicolumn{1}{|l|}{CAT-E}    & 95   & 88   & 24  & 46  & 71   & 85   & 101  & 151 \\ \hline
+\multicolumn{1}{|l|}{CAT-F}    & 49    & 84   & 22  & 48  & 72   & 81   & 91   & 155 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by RECAT]{AROT statistics for the air traffic mix at BIKF by RECAT-EU categories. The count is the number of landings in peak hours since October 2017}
+\label{tab:AROT_RECAT_stats}
+\end{table}
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{graphics/BIKF_schematic.png}
+    \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways.}
+    \label{fig:BIKF_schematic}
+\end{figure}
 % ----------------------------

### 2018-12-20 introduction.tex
finished writing on Traffic mix
fixed title fonts on figures, sorted boxplot, BIKF schematic.

@@
 
-\begin{figure}
+\begin{figure}[h]
     \centering
@@
 
-\begin{table}[]
+\begin{table}[h]
     \centering
@@
 
-The ICAO wake turbulence categories outline the worst case for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018}.
+The ICAO wake turbulence categories (Table~\ref{tab:ICAO_WTC}) outline the worst case for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018}.
 The separation distances for instrumental flight rules are observed for the follower aircraft and are conventionally expressed in nautical miles (NM).
@@
 \multicolumn{2}{|c|}{}                                 & Super (A380-800) & Heavy & Medium & Light \\ \hline
-\multirow{4}{*}{Leader}       & Super (A380-800)       & (*)                & 6 NM  & 7 NM   & 8 NM   \\ \cline{2-6} 
+\multirow{4}{*}{\rotatebox[origin=c]{90}{Leader}}       & Super (A380-800)       & (*)                & 6 NM  & 7 NM   & 8 NM   \\ \cline{2-6} 
                               & Heavy                  & (*)                & 4 NM  & 5 NM   & 6 NM   \\ \cline{2-6} 
@@
 
-It must be noted that the division of ICAO WTC categories into the RECAT scheme is not achieved by splitting each category in half.
-The criteria used for categorisation of existing and new aircraft types are provided in detail in~\cite{noauthor_recat_2018}.
+It must be noted that the division of ICAO WTC into the RECAT scheme (Table~\ref{tab:RECAT-dist}) is not achieved by splitting each category in half.
+The criteria used for categorisation of existing and new aircraft types are provided in detail in~\cite{rooseleer2015recat}.
 The case from the traffic mix at BIKF presents the arrangement shown in Table~\ref{tab:wtc2recat_division}.
-A more detailed illustration of the re-categorisation is presented in the coming section \ref{ssec:traffic_mix} on aircraft traffic mix (Figure~\ref{fig:post_fast_exit_mix_pie_v2}).
+A more detailed illustration of the re-categorisation is presented in the upcoming section \ref{ssec:traffic_mix} on aircraft traffic mix (Figure~\ref{fig:post_fast_exit_mix_pie_v2}).
 
@@
 % If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
-\begin{table}[]
+\begin{table}[h]
 \centering
@@
 \caption{Transition from ICAO WTC to RECAT-EU categories.
-The categorisation process and criteria for assigning an existing aircraft type into RECAT-EU scheme is illustrated in detail in~\cite{noauthor_recat_2018}}
+The categorisation process and criteria for assigning an existing aircraft type into RECAT-EU scheme is illustrated in detail in~\cite{rooseleer2015recat}}.
 \label{tab:wtc2recat_division}
@@
 
-\fxnote{explain the transition from WTC to REACT , and point the case for BIKF: H to ABC, M to CDEF, ....reference the pie chart on mix}
+The noticeable changes in Table~\ref{tab:RECAT-dist} from Table~\ref{tab:ICAO_WTC} affect a follower aircraft, whose relative position from the ICAO scheme is shifted to the left in the new table after re-categorisation.
+Such benefit experiences for example a Heavy follower aircraft that is re-categorised as CAT-C, in which case the separation from the leader is reduced with up to 3 NM.
 
@@
 \multicolumn{2}{|c|}{}                                 & CAT\_A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
-\multirow{6}{*}{Leader}             & CAT-A            & 3 NM   & 4 NM  & 5 NM  & 5 NM  & 6 NM  & 8 NM  \\ \cline{2-8} 
+\multirow{6}{*}{\rotatebox[origin=c]{90}{Leader}}            & CAT-A            & 3 NM   & 4 NM  & 5 NM  & 5 NM  & 6 NM  & 8 NM  \\ \cline{2-8} 
                                     & CAT-B            &    (*)    & 3 NM  & 4 NM  & 4 NM  & 5 NM  & 7 NM  \\ \cline{2-8} 
@@
 \caption[RECAT-EU distance-based separation minima]{RECAT-EU WT distance-based separation minima on approach and departure.
-(*) means minimum radar separation (MRS), set at 2.5 NM, applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat}.}
+(*) means minimum radar separation (MRS), set at 2.5 NM, applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat, noauthor_recat_2018}.}
 \label{tab:RECAT-dist}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{multirow}
-% \usepackage{graphicx}
-\begin{table}[]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-\multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
-\multicolumn{2}{|c|}{}                                 & CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
-\multirow{6}{*}{Leader}             & CAT-A            &        & 100s  & 120s  & 140s  & 160s  & 180s  \\ \cline{2-8} 
-                                    & CAT-B            &        &       &       & 100s  & 120s  & 140s  \\ \cline{2-8} 
-                                    & CAT-C            &        &       &       & 80s   & 100s  & 120s  \\ \cline{2-8} 
-                                    & CAT-D            &        &       &       &       &       & 120s  \\ \cline{2-8} 
-                                    & CAT-E            &        &       &       &       &       & 100s  \\ \cline{2-8} 
-                                    & CAT-F            &        &       &       &       &       & 80s   \\ \hline
-\end{tabular}%
-}
-\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure~\cite{rooseleer2015recat}}
-\label{tab:RECAT-time}
 \end{table}
@@
 High load interval or window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavík Airport.
-The time between flights during the high load interval is set at $\leq$4~minutes.
+The time between flights during the high load interval is set at $\leq$4~minutes (Figure~\ref{fig:Peak_Diagram}).
 This value is based on statistical data by Isavia in order to achieve two distinct peak hours: one in the morning and one in the afternoon.
 The last flight in a high load window that fulfils these requirements is not considered to be a part of the peak as its behaviour is not affected by an instantaneous next flight after it.
+The AROT metric is essential for the project because a following aircraft is not allowed to land on the same runway before it has been vacated by the leading aircraft.
+The interval to the next landing aircraft is specified as the Landing Time Interval (LTI) which is directly linked to the inter-arrival distance or in other words the wake turbulence separation.
 
-\begin{figure}
+\begin{figure}[h]
     \centering
@@
 The mixed modes indicate equal arrivals and departures (Mixed~Mode~A) or two times more arrivals than departures (Mixed Mode B), the later being a tendency during the afternoon peaks.
-The script producing this plot is on the Isavia´s internal GUI Víkingaskipið.}  \label{fig:capacity_evnelope}
+The figure is from Isavia´s internal GUI Víkingaskipið.}  \label{fig:capacity_evnelope}
 \end{figure}
@@
                                                         & CAT-E &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{67FD9A}{\color[HTML]{000000} -1 NM} \\ \cline{2-8}
-\multirow{-6}{*}{Leader}                                & CAT-F &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{FD6864}(+0,5 NM)                    \\ \hline
+\multirow{-6}{*}{\rotatebox[origin=c]{90}{Leader}}                                & CAT-F &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{FD6864}(+0,5 NM)                    \\ \hline
 \end{tabular}%
@@
 
-The implementation of RECAT shifts the wake separation minima for arrival pairs thus creating the potential for reduced time separation between the aircraft.
+The implementation of RECAT shifts the wake separation minima for arrival and departure pairs thus creating the potential for reduced time separation between the aircraft.
 The reduction of the time interval between landings facilitates increas in the throughput capacity of the runways.
@@
 
-\begin{figure}[ht]
+\begin{figure}[h]
     \centering
@@
 The observed time period is from October 2017 to November 2018.
-The red line indicates the ICAO WTC reference time separation for the selected H-M aircraft pair.}\label{fig:AROT_LTI_rwy19_H_M}
+The red line indicates the ICAO WTC reference time separation for the selected H-M aircraft pairs.
+The figure is from the Isavia´s internal GUI Víkingaskipið.}\label{fig:AROT_LTI_rwy19_H_M}
 \end{figure}

### 2018-12-21 code.tex
finished arot considerations, rwy usage.

@@
 
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[h]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
+\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{RWY 01} & 963 & 85 & 21 & 46 & 68 & 88 & 100 & 152 \\ \hline
+\multicolumn{1}{|l|}{RWY 10} & 852 & 85 & 11 & 61 & 78 & 84 & 91 & 155 \\ \hline
+\multicolumn{1}{|l|}{RWY 19} & 1511 & 67 & 10 & 46 & 61 & 66 & 71 & 155 \\ \hline
+\multicolumn{1}{|l|}{RWY 28} & 401 & 68 & 14 & 49 & 60 & 66 & 72 & 155 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by runway for the summer]{AROT statistics for the air traffic mix at KEF by runway for the summer of 2018. The count is the number of landings in peak hours.}
+\label{my-label3}
+\end{table}
+\fxnote{Include a schematic of the airport, overview for runway and taxiway reference}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+\begin{table}[h]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
+\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\multicolumn{1}{|l|}{RWY 01} & 284 & 88 & 26 & 47 & 63 & 90 & 106 & 153 \\ \hline
+\multicolumn{1}{|l|}{RWY 10} & 333 & 93 & 12 & 70 & 85 & 91 & 99 & 144 \\ \hline
+\multicolumn{1}{|l|}{RWY 19} & 251 & 72 & 12 & 51 & 65 & 70 & 75 & 140 \\ \hline
+\multicolumn{1}{|l|}{RWY 28} & 69 & 73 & 13 & 49 & 63 & 72 & 79 & 105 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by runway for the winter]{AROT statistics for the air traffic mix at KEF by runway for the winter season.
+The count is the number of landings in peak hours since October 2017}
+\label{my-label4}
+\end{table}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
+\begin{table}[h]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{lr|r|r|r|r|r|r|r|}
+\cline{3-9}
+ & \multicolumn{1}{l|}{} & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
+\multicolumn{1}{|l|}{month} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}January} & 95 & 86 & 20 & 54 & 70 & 85 & 96 & 136 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}February} & 80 & 87 & 16 & 52 & 75 & 86 & 93 & 144 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}March} & 215 & 84 & 18 & 47 & 70 & 85 & 95 & 150 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}April} & 221 & 82 & 17 & 47 & 69 & 81 & 93 & 139 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}May} & 386 & 75 & 13 & 46 & 67 & 74 & 83 & 140 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}June} & 647 & 73 & 16 & 47 & 62 & 68 & 80 & 155 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}July} & 729 & 72 & 16 & 46 & 61 & 68 & 77 & 155 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}August} & 724 & 80 & 18 & 46 & 64 & 81 & 91 & 148 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}September} & 632 & 78 & 18 & 46 & 64 & 75 & 88 & 152 \\ \hline
+\rowcolor[HTML]{FFFC9E} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{FFFC9E}October} & 609 & 79 & 17 & 46 & 66 & 76 & 91 & 155 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}November} & 220 & 84 & 23 & 48 & 66 & 82 & 99 & 151 \\ \hline
+\rowcolor[HTML]{DAE8FC} 
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}December} & 106 & 87 & 23 & 51 & 69 & 86 & 99 & 153 \\ \hline
+\end{tabular}%
+}
+\caption[AROTs for the air traffic mix by month]{AROT statistics for the air traffic mix at BIKF by month.
+The count is the number of landings in peak hours from October 2017 til November 2018.
+The colour fields indicate a subjective separation of the data into summer and winter season, based on mean AROT value.
+Months with mean AROT $\leq$ 80 seconds are classified as summer, and the remaining as winter.}
+\label{tab:month2season_arot}
+\end{table}
+
 \begin{figure}

### 2018-12-22 introduction.tex
finished arot considerations, rwy usage.
began LTI method

@@
 Keflavík International Airport (IATA:KEF, ICAO:BIKF) is the major gateway to Iceland currently with two functioning runways.
-The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South) , RWY-10 (East-West) and RWY-28 (West-East).
+The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South) , RWY-10 (West-East) and RWY-28 (East-West).
 There has been a gradual increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017~\cite{isavia_facts_2017}.
@@
 
-\fxnote{describe rot in general and what factors have influence on rot, theory, models for rot, see validation for rot model paper}
-
 The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway~\cite{ROT_definition}. \\

### 2018-12-22 introduction.tex
finished method and stared result section

@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \centering
@@
 
-\section{Runway Occupancy and Landing Time Interval Study Objective}
+\section{Runway Occupancy and Landing Time Interval Study Objective}\label{sec:study_objective}
 The goal of examining the arrival runway occupancy times and the landing time intervals (LTI) is to determine whether the implementation of the RECAT-EU scheme at BIKF under certain conditions, would affect the throughput of the airfield.

### 2018-12-22 code.tex
start of results

@@
 Almost half of all arrivals during that period (44,7\%) use RWY-19, followed by RWY-01 (25,7\%) and RWY-10 (17.6\%).
-The summer traffic favours RWY-19 with 52.8\%~(Figure~\ref{fig:runway_usage_summer}), while during the winter season some of those arrivals are diverted towards RWY-01 and RWY-10~(Figure~\ref{fig:runway_usage_summer}).
-RWY-28 is least used despite its fast-exit TWY B-1.}
+The summer traffic favoured RWY-19 with 52.8\%~(Figure~\ref{fig:runway_usage_summer}), while during the winter season some of those arrivals were diverted towards RWY-01 and RWY-10~(Figure~\ref{fig:runway_usage_summer}).
+RWY-28 was least used despite its fast-exit TWY B-1.}
     \label{fig:runway_usage}
@@
     \includegraphics[width=0.7\textwidth]{graphics/fig_runway_usage_winter}
-    \caption[Winter runway usage at BIKF]{The winter traffic at BIKF is a bit more evenly distributed among runways bit still favours RWY-19 (38,8\%),followed by RWY-01 with 28,9\%, RWY-10 (20,8\%) and RWY-28 (11,4\%).}
+    \caption[Winter runway usage at BIKF]{The winter traffic at BIKF is slightly more evenly distributed among runways but still favours RWY-19 (38,8\%),followed by RWY-01 with 28,9\%, RWY-10 (20,8\%) and RWY-28 (11,4\%).}
     \label{fig:runway_usage_winter}
@@
 The colour fields indicate a subjective separation of the data into summer and winter season, based on mean AROT value.
-Months with mean AROT $\leq$ 80 seconds are classified as summer, and the remaining as winter.}
+Months with mean AROT $\leq$ 80 seconds were classified as summer, and the remaining as winter.}
 \label{tab:month2season_arot}

### 2018-12-29 code.tex
finished results and analysis chapter

@@
 
-% ----------------------------
-
 % Please add the following required packages to your document preamble:
@@
 \end{table}
-
-% ----------------------------
 
@@
 \end{table}
-\fxnote{Include a schematic of the airport, overview for runway and taxiway reference}
 
@@
 
+\chapter{Inter-arrival distance separation}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{graphics/fig_HM_to_DD_pairs_dist_separ.png}
+    \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs ti D-D pairs into the RECAT scheme.}
+    \label{fig:HM_to_DD_pairs_dist_separ}
+\end{figure}
+
+\chapter{BIKF Schemati}
+
 \begin{figure}
@@
 \end{figure}
+
 % ----------------------------

### 2018-12-29 introduction.tex
finished results and analysis chapter

@@
 \subsection{Current Understandings of Wake Vortex Behaviour}
-The wing of an aircraft can be regarded as a beam of finite length with aerofoil as cross section~\cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation or drop it if unnecessary.
-Read about induced vertical flow in a book on aerodynamics}.
-The lift effect is created by the differential pressure between the lower and the upper side of the wing.
+
+The lift effect on wing is created by the differential pressure between the lower and the upper surface of the wing.
 At the wing tips the high pressure flow from the lower side leaks around the tip and to the upper side of the wing.
 Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards.
-A finite lifting wing leaves behind a continuous sheet of stream-wise vorticity ~\cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation }.
-This manifests itself in multiple trailing swirls of the airflow that mix together and roll up into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop}~\cite{magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
-Between the vortices the air moves downwards while outside the induced flow is upwards.
+When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and combine into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
+Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards.
 Those vortices are also known as wake turbulence.
 The kinetic energy contained in the vortices is dependent on the weight and aerodynamics of the generating aircraft.
-The cross flow velocities in the core region of the trailing vortices can reach $360$ km/h and the vortices can stay effective up to hundred wing spans.
-This can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft~\cite{Breitsamter2011Feb, gerz_commercial_2002}.
+The cross flow velocities in the core region of the trailing vortices can reach $360$ km/h and the vortices can stay effective up to hundred wing spans, which can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft~\cite{Breitsamter2011Feb, gerz_commercial_2002}.
 
@@
     \item Atmospheric turbulence extracts energy from the vortex and reduces its strength leading to a faster wake decay ~\cite{Hallock2018Apr}.
-    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence~\cite{Hallock2018Apr}.\fxnote{This needs to be explained better (Ármann)}
+    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence.
+The so called dissipative action of viscosity  effectively removes energy from a disturbance, in this case a vortex, thereby causing it to decay~\cite{Hallock2018Apr, houghton2012aerodynamics}.
     \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level~\cite{Holzapfel2001Feb, gerz_commercial_2002}.

### 2018-12-29 code.tex
typos fixed

@@
     \includegraphics[width=0.5\textwidth]{graphics/fig_HM_to_DD_pairs_dist_separ.png}
-    \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs ti D-D pairs into the RECAT scheme.}
+    \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs to D-D pairs into the RECAT scheme.}
     \label{fig:HM_to_DD_pairs_dist_separ}
@@
 
-\chapter{BIKF Schemati}
+\chapter{BIKF Schematic}
 

### 2018-12-29 appendix_1.tex
minor changes to appendix

--- 2018-12-29 code.tex
@@
 
+% Please add the following required packages to your document preamble:
+% \usepackage{multirow}
+% \usepackage{graphicx}
+\begin{table}[h]
+\centering
+\resizebox{0.8\textwidth}{!}{%
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
+\hline
+\multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
+\multicolumn{2}{|c|}{}& CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
+\multirow{6}{*}{\rotatebox[origin=c]{90}{Leader}}& CAT-A&& 100s  & 120s  & 140s  & 160s  & 180s  \\ \cline{2-8}
+& CAT-B&&&& 100s  & 120s  & 140s  \\ \cline{2-8} 
+& CAT-C&&&& 80s   & 100s  & 120s  \\ \cline{2-8} 
+& CAT-D&&&&&& 120s  \\ \cline{2-8} 
+& CAT-E&&&&&& 100s  \\ \cline{2-8} 
+& CAT-F&&&&&& 80s   \\ \hline
+\end{tabular}%
+}
+\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure~\cite{rooseleer2015recat}}
+\label{tab:RECAT-time}
+\end{table}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=1.0\textwidth]{graphics/fig_best_CC_from_HM_pairs_time_sep.png}
+    \caption[AROT and LTI of C-C pairs from ICAO H-M pairs on RWY~19]{AROT and landing time intervals of C-C pairs originating from ICAO H-M pairs for RWY-19 during the summer season.
+The dashed red line indicates the RECAT reference time separation for the C-C pairs.}
+    \label{fig:HM_to_DD_pairs_dist_separ}
+\end{figure}
+
 \chapter{BIKF Schematic}
 
-\begin{figure}
+\begin{figure}[h]
     \centering
-    \includegraphics[width=1\textwidth]{graphics/BIKF_schematic.png}
+    \includegraphics[width=0.8\textwidth]{graphics/BIKF_schematic.png}
     \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways.}

### 2018-12-29 introduction.tex
minor changes to appendix

@@
 }
-\caption{Transition from ICAO WTC to RECAT-EU categories.
+\caption[Transition from ICAO WTC to RECAT-EU categories]{Transition from ICAO WTC to RECAT-EU categories.
 The categorisation process and criteria for assigning an existing aircraft type into RECAT-EU scheme is illustrated in detail in~\cite{rooseleer2015recat}}.
@@
 }
-\caption[Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)~\cite{rooseleer2015recat}}
+\caption[Difference in wake separation minima between ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)~\cite{rooseleer2015recat}}
 \label{tab:delta_distance_wtc2recat}

### 2019-01-04 appendix_1.tex
minor changes
beginning of Summary

@@
 
+\clearpage
 \chapter{Arrival Runway Occupancy Times\label{cha:AROTs}}
@@
 \centering
-\resizebox{\textwidth}{!}{%
+\resizebox{0.8\textwidth}{!}{%
 \begin{tabular}{lr|r|r|r|r|r|r|r|}
@@
 
-\chapter{Inter-arrival distance separation}
+\clearpage
+\chapter{Inter-arrival time and distance separation}
 
@@
 The dashed red line indicates the RECAT reference time separation for the C-C pairs.}
-    \label{fig:HM_to_DD_pairs_dist_separ}
+    \label{fig:best_CC_from_HM_pairs_time_sep}
 \end{figure}
 
+\clearpage
 \chapter{BIKF Schematic}
@@
     \includegraphics[width=0.8\textwidth]{graphics/BIKF_schematic.png}
-    \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways.}
+    \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways \cite{BIKF_scheme}.}
     \label{fig:BIKF_schematic}
@@
 
+\clearpage
+\chapter{BIKF Wind Rose}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.7\textwidth]{graphics/BIKF_windrose_2005-2014.pdf}
+    \caption[BIKF Wind Rose]{Wind rose - frequency, direction and strength of winds at BIKF aerodrome \cite{wind_rose_2014}.}
+    \label{fig:BIKF_wind_rose}
+\end{figure}
 % ----------------------------

### 2019-01-04 introduction.tex
beginning of Summary

@@
 Traditional separation standards, introduced in the seventies, are defined by ICAO  on the basis of certificated maximum take-off mass (MTOM) as shown in Table~\ref{tab:WTC}.
-The prescribed categories are three (i.e.Heavy, Medium and Light).
+The prescribed categories are three i.e.
+Heavy, Medium and Light.
 In addition for aircraft in the order of $560000$~kg ICAO provides a subcategory to the Heavy types called Super Heavy.
@@
 \multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
-\multicolumn{2}{|c|}{}                                 & CAT\_A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
+\multicolumn{2}{|c|}{}                                 & CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
 \multirow{6}{*}{\rotatebox[origin=c]{90}{Leader}}            & CAT-A            & 3 NM   & 4 NM  & 5 NM  & 5 NM  & 6 NM  & 8 NM  \\ \cline{2-8} 
@@
 }
-\caption[RECAT-EU distance-based separation minima]{RECAT-EU WT distance-based separation minima on approach and departure.
+\caption[RECAT-EU distance-based separation minima]{RECAT-EU wake turbulence distance-based separation minima on approach and departure.
 (*) means minimum radar separation (MRS), set at 2.5 NM, applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat, noauthor_recat_2018}.}
@@
 \end{table}
+
+The different category code names in this study will also be referred to on a single letter basis for simplification e.g.
+"CAT-C" will be abbreviated to "C", "CAT-D" to "D", "Heavy" to "H", Medium to "M" and so forth.
+This applies to pair categories by abbreviating e.g.
+"CAT-C leader - CAT-D follower"  to "C-D pair" or simply "C-D", and "Medium-Medium" pair to "M-M".
 

### 2019-01-04 introduction.tex
fixed some of Armann's comments

@@
 
-The lift effect on wing is created by the differential pressure between the lower and the upper surface of the wing.
+The lift effect on an aircraft wing is created by the differential pressure between the lower and the upper surface of the wing.
 At the wing tips the high pressure flow from the lower side leaks around the tip and to the upper side of the wing.
@@
     \caption[Wake vortex roll-up process]{Wing vortex evolution and roll-up process.
-Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one~\cite[p.
-043]{magazine_aibus_safety}} \label{fig:vortex_develop}
+Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one~\cite[p.~043]{magazine_aibus_safety}} \label{fig:vortex_develop}
 \end{figure}
@@
 
-Near ground on approach the wake vortices tend to descend slowly to an altitude of about one half of the initial separation.
+Wake vortices tend to descend slowly to an altitude of about one half of the initial separation, near ground on approach for landing,.
 Upon reaching this point the descent will stop and then ascend slowly.
 This "re-bounce" effect is caused by the presence of the ground and carries with it the formation of a second pair of induced vortices outside and bellow the main vortex.
-In the presence of stable cross-wind conditions the decay of the downwind vortex is not identical to the upwind vortex and the descent rate may vary significantly \cite{Hallock2018Apr}. 
+In the presence of stable cross-wind conditions the decay of the downwind vortex is not identical to the upwind vortex and the descent rate may vary significantly~\cite{Hallock2018Apr}. 
 The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure~\ref{fig:vortex_ground_effect}).
@@
     \includegraphics[width=0.8\textwidth]{graphics/Hallock_vortex_evolution.jpg}
-    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect~\cite[p. 29]{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
+    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect~\cite[p.~29]{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
 \end{figure}
@@
     \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft.
-(A) induced rolling moment, (B) upward motion, (C) loss of lift~\cite[p.
-33]{Hallock2018Apr}} \label{fig:vortex_encounter}
+(A) induced rolling moment, (B) upward motion, (C) loss of lift~\cite[p.~33]{Hallock2018Apr}} \label{fig:vortex_encounter}
 \end{figure}
@@
     \includegraphics[width=0.8\textwidth]{graphics/fig_rot_landig_time_interval_RWY19_leader_H_follower_M_peak-hour_BAR_20171004_20181130.png}
-    \caption[Caption table]{Arrival runway occupancy time (blue) and landing time intervals (green) for Heavy leader - Medium follower pairs on BIKF runway RWY-19.
+    \caption[AROT and LTI of H-M pairs on RWY~19]{Arrival runway occupancy time (blue) and landing time intervals (green) for Heavy leader - Medium follower pairs on BIKF runway RWY-19.
 The observed time period is from October 2017 to November 2018.

### 2019-01-05 introduction.tex
initial corrections on Intro

@@
 The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South) , RWY-10 (West-East) and RWY-28 (East-West).
-There has been a gradual increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62,3 thousand air transport movements in 2017~\cite{isavia_facts_2017}.
-For 2018 the passenger numbers show a growth of 13.1\% on average (at the time of writing) from the previous year~\cite{isavia_pass_statistics_2018}.
-The increase in traffic at Keflavík Airport can encounter constrained capacity of the runways during peak traffic periods.
-A significant constraint of airport capacity is caused by separation regulations of aircraft due to wake-vortices.
+There has been a unforeseen increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined by the International Civil Aviation Organisation (ICAO) as an aircraft take-off or landing at an airport.
+For airport traffic purposes one arrival and one departure is counted as two movements~\cite{aircraft_movement}.
+For 2018 the passenger numbers show an yearly growth of 13.44\% (at the time of writing, for the same time period) from the previous year~\cite{isavia_pass_statistics_2018}.
+The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods.
+A significant constraint of airport capacity is caused by distance separation regulations of aircraft due to wake turbulence.
+This term describes the effect of the rotating air masses generated behind the wing tips of large jet aircraft.
+
 These are specified by rules established by the European Aviation Safety Agency (EASA) for each type of aircraft based primarily on the weight of the aircraft.
@@
 \subsection{Current Understandings of Wake Vortex Behaviour}
-
+Wake vortices describe the nature of the air flow generated behind the wing tips of a jet aircraft~\cite{doc44444}.
 The lift effect on an aircraft wing is created by the differential pressure between the lower and the upper surface of the wing.

### 2019-01-06 introduction.tex
finished summary and conclusion

@@
 
-Keflavík International Airport (IATA:KEF, ICAO:BIKF) is the major gateway to Iceland currently with two functioning runways.
-The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South) , RWY-10 (West-East) and RWY-28 (East-West).
+Keflavík International Airport (IATA: KEF, ICAO: BIKF) is the major gateway to Iceland currently with two functioning runways.
+The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West).
 There has been a unforeseen increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined by the International Civil Aviation Organisation (ICAO) as an aircraft take-off or landing at an airport.
@@
 The implementation of RECAT shifts the wake separation minima for arrival and departure pairs thus creating the potential for reduced time separation between the aircraft.
-The reduction of the time interval between landings facilitates increas in the throughput capacity of the runways.
+The reduction of the time interval between landings facilitates increase in the throughput capacity of the runways.
 Figure~\ref{fig:AROT_LTI_rwy19_H_M} may be used to illustrate the hypothesis.
-The current ICAO time separation reference, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT scheme.
+The current ICAO time reference separation, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT scheme.
 This relocation will be more noticeable for RECAT C-C and C-D pairs formed from the ICAO HEAVY-MEDIUM pairs, in which case the separation is reduced from 5 NM to 3 NM.
@@
 The red line indicates the ICAO WTC reference time separation for the selected H-M aircraft pairs.
-The figure is from the Isavia´s internal GUI Víkingaskipið.}\label{fig:AROT_LTI_rwy19_H_M}
+The figure is from the Isavia's internal GUI Víkingaskipið.}\label{fig:AROT_LTI_rwy19_H_M}
 \end{figure}

### 2019-01-07 introduction.tex
fixed Intro up to 1.1.3. WV separation

@@
 \chapter{Introduction\label{cha:introduction}}
-%% \ifdraft only shows the text in the first argument if you are in draft mode.
-%% These directions will disappear in other modes.
-% \ifdraft{State the objectives of the exercise. Ask yourself:
-%   \underline{Why} did I design/create the item? What did I aim to
-%   achieve? What is the problem I am trying to solve?  How is my
-%   solution interesting or novel?}{}
 
@@
 The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West).
-There has been a unforeseen increase of passengers travelling through Keflavík Airport in the recent years with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined by the International Civil Aviation Organisation (ICAO) as an aircraft take-off or landing at an airport.
-For airport traffic purposes one arrival and one departure is counted as two movements~\cite{aircraft_movement}.
-For 2018 the passenger numbers show an yearly growth of 13.44\% (at the time of writing, for the same time period) from the previous year~\cite{isavia_pass_statistics_2018}.
+There has been an unforeseen increase of passengers travelling through Keflavík Airport in the recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
+For 2018 the passenger numbers show an yearly growth of 13.44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
 The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods.
-A significant constraint of airport capacity is caused by distance separation regulations of aircraft due to wake turbulence.
-This term describes the effect of the rotating air masses generated behind the wing tips of large jet aircraft.
-
-These are specified by rules established by the European Aviation Safety Agency (EASA) for each type of aircraft based primarily on the weight of the aircraft.
-The general flight safety requirements are established by the International Civil Aviation Organisation (ICAO) for maintaining safe distance between aircraft.\fxnote{Introduction needs more work, incomplete.}
+Airport capacity is generally defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology}.
+\\
+A significant constraint of airport capacity is caused by distance separation regulations between aircraft in flight due to wake turbulence.
+The term "wake turbulence" describes the effect of the rotating air masses generated behind the wing tips of large jet aircraft~\cite{doc4444full}.
+A smaller aircraft can experience loss of control in close proximity behind a larger aircraft because of those swirling air masses.
+The general flight safety requirements for maintaining safe distance between aircraft are established by the International Civil Aviation Organisation (ICAO)~\cite{doc4444full}.
+The European Aviation Safety Agency (EASA) provides oversight of those requirements and their application by the Air Traffic Management (ATM).
+\\
+The wake separation due to wake turbulence is especially critical during take-off and on approach for landing.
+ICAO has specified wake separation minima requirements that are applied to aircraft in the flight-path of another aircraft.
+Those requirements classify each aircraft in a wake turbulence category (WTC) and the wake separation is applied to a following aircraft based on the wake category of a leading aircraft.
+Thus the aircraft form a pair composed of a "leader" and a "follower", that observes a certain wake separation distance in nautical miles (NM).
+Generally the ICAO wake categories are three - Heavy, Medium and Light.
+The classification is based primarily on the weight of the aircraft, but this can vary slightly for different states.\\
+The ICAO WTC categorisation is based on observations and calculations performed on aircraft models that are some 50 years old.
+Even though those separations are considered reliable, they are often conservative and outdated for newer aircraft models and as such generate unnecessary over-separation in some cases.
+In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} or just RECAT has been devised by EUROCONTROL~\cite{EUROCONTROL_who_is}, an intergovernmental organisation assisting in safe and efficient air traffic operations.
+RECAT is based on the ICAO wake turbulence categories, but takes into account the wake specifications of current aircraft models.
+The RECAT scheme splits aircraft into six wake categories, generally referred to as CAT-A, CAT-B, CAT-C, CAT-D, CAT-E and CAT-F.
+Mostly, re-categorisation leads to decreased wake separation requirements for aircraft pairs, which in turn can lead to accommodating more landing and departures per time interval and increase airport capacity (throughput).\\
+Apart from wake separation, the other major element that affects capacity is runway occupancy time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
+The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, per se, can limit the throughput of a runway.
+The emphasis in this project will be on runway occupancy times for arriving aircraft (AROT) during peak traffic hours at the airport\\
+The runway occupancy time and the wake separation requirements are the two main metrics that directly influence the throughput of a runway.
+The aim of this study is to estimate the constraints for implementing the RECAT wake turbulence categorisation and separation minima for Keflavík International Airport, the effect this can produce for airfield throughput and when is the runway occupancy a limiting factor for increased capacity.
 
 \section{Background and Literature Review}
-The research parameters for the literate review included but were not limited to the following concepts: runway occupancy time, runway capacity, wake vortex separation.
-The project addresses in particular the arrival runway occupancy and its seasonal fluctuations as well as the variations between the different runways at Keflavík International Airport.
-%\ifdraft{Provide background about the subject matter (e.g. %How was morse code developed?  How is it used today?). 
-%This is a place where there are usually many citations.
-%It is suspicious when there is not.
-%Include the purpose of the different equipment and your %design intent. 
-%Include references to relevant scientific/technical work and %books.
-%What other examples of similar designs exist?
-%How is your approach distinctive?
-
-%If you have specifications or related standards, these %mustscribed and cited also.  As an example, you might cite %the specific
-%RoboSub competition website (and documents) if working on the %lighting system for an AUV
-
-%%% Glossary is broken, do not use --foley
-%% \gls{auv}\footnote{Autonomous Undersea Vehicle}.
-%
-%% Notice that there is now information on the AUV in the %Index %and Acronyms.
-%% It isn't in the \gls{glossary} because we didn't put it %there.
-%\index{AUV}
-%}{}
 
 \subsection{Current Understandings of Wake Vortex Behaviour}
-Wake vortices describe the nature of the air flow generated behind the wing tips of a jet aircraft~\cite{doc44444}.
+Wake vortices describe the nature of the air flow generated behind the wing tips of a jet aircraft~\cite{doc4444full}.
 The lift effect on an aircraft wing is created by the differential pressure between the lower and the upper surface of the wing.
@@
 Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards.
-When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and combine into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
+When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and are combined into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
 Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards.
@@
 The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure~\ref{fig:vortex_ground_effect}).
-Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$~m and $518$~m and in some cases even $762$~m, which is the basis for the "2500-foot"  rule separation in parallel runways configuration~\cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
+Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$~m and $518$~m and in some cases even $762$~m, which is the basis for the "2500-foot" rule separation in parallel runway configurations~\cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
 The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5$~m/s for the first $30$~s after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m and below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}.
@@
 \subsection{Wake Vortex Separation}
-An aircraft affected by the wake vortex can experience loss of lift or an induced rolling moment and velocity fluctuations (Figure~\ref{fig:vortex_encounter}).
+A second aircraft following the aircraft generating wake vortices can experience loss of lift or an induced rolling moment and velocity fluctuations (Figure~\ref{fig:vortex_encounter}).
+
 This can lead to the up-set and potentially loss of control of an aircraft following the flight path of a preceding aircraft.
@@
 R.
-de Neufville~\cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure maximum throughput capacity.
+de Neufville~\cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure is maximum throughput capacity.
 It indicates the average number of movements that can be performed on the runway system in one hour in the presence of continuous demand, while adhering to all the separation requirements imposed by the air traffic management (ATM) system~\cite{de_neufville_airport_2013}.

### 2019-01-07 appendix_1.tex
added new BIKF schematic

@@
     \centering
-    \includegraphics[width=0.8\textwidth]{graphics/BIKF_schematic.png}
-    \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways \cite{BIKF_scheme}.}
+    \includegraphics[width=1\textwidth]{graphics/BIKF_schematic.png}
+    \caption[BIKF schematic]{BIKF schematic with four marked runways in perpendicular configuration (dashed yellow lines) and two fast-exit taxiways (red arrows): TWY~A-1 on RWY~01 and TWY~B-1 on RWY~28 (source:~ Isavia).}
     \label{fig:BIKF_schematic}

### 2019-01-08 introduction.tex
fixed intro up to WVS

@@
 \\
-The wake separation due to wake turbulence is especially critical during take-off and on approach for landing.
+The distance separation due to wake turbulence is especially critical during take-off and on approach for landing.
 ICAO has specified wake separation minima requirements that are applied to aircraft in the flight-path of another aircraft.
@@
 \section{Background and Literature Review}
+This section briefly reviews the necessity of adopting a wake vortex separation scheme for aircraft from the perspective of fluid dynamics.
+Wake vortex characteristics of conventional aircraft have been studied for several decades now and "it is generally acknowledged that the near wake of a large jet airliner poses a significant hazard to any smaller aircraft that follows it into of out of an air terminal"~\cite{dup._donamdson_vortex_1975}.
 
@@
 When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and are combined into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
-Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards.
-Those vortices are also known as wake turbulence.
+Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards.\\
+The effect of wake vortices on air masses is known as wake turbulence. 
 The kinetic energy contained in the vortices is dependent on the weight and aerodynamics of the generating aircraft.
 The cross flow velocities in the core region of the trailing vortices can reach $360$ km/h and the vortices can stay effective up to hundred wing spans, which can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft~\cite{Breitsamter2011Feb, gerz_commercial_2002}.
-
-\begin{figure}[ht]
+\begin{figure}[h]
     \centering
@@
 \subsection{Ground Effect and Decay Process}
-
-Wake vortices tend to descend slowly to an altitude of about one half of the initial separation, near ground on approach for landing,.
+Wake vortices tend to descend slowly to an altitude of about one half of the initial separation, near ground on approach for landing.
 Upon reaching this point the descent will stop and then ascend slowly.
@@
     \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level~\cite{Holzapfel2001Feb, gerz_commercial_2002}.
-    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may cause the vortices to link and decay faster~\cite{Hallock2018Apr, crow2003stability}.
+    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may cause the vortices to link and decay faster~\cite{dup._donamdson_vortex_1975, Hallock2018Apr, crow2003stability}.
     \item Secondary vorticity structures, vertical rib-like counter rotating flow formations may appear between the vortex pair and eventually wrap around the main vortices, leading to turbulence build inside the system and rapid circulation decay~\cite{Holzapfel2001Feb, Holzapfel2003Jun}.
 \end{itemize}
-
 \begin{figure}[h]
@@
 
-\subsection{Wake Vortex Separation}
-A second aircraft following the aircraft generating wake vortices can experience loss of lift or an induced rolling moment and velocity fluctuations (Figure~\ref{fig:vortex_encounter}).
-
+\section{Wake Vortex Separation}
+Distance separation between aircraft is necessary to mitigate the effect of the wake vortices.
+This is especially true for aircraft in approach for landing or take-off, "since practically all serious problems associated with wake hazard occur in the vicinity of airports"~\cite{dup._donamdson_vortex_1975}.
+The wake vortices generated by a leading aircraft (leader) can cause loss of lift or an induced rolling moment and velocity fluctuations in a following aircraft (follower) if the follower enters the wake turbulence region near the leader~(Figure~\ref{fig:vortex_encounter}).
 This can lead to the up-set and potentially loss of control of an aircraft following the flight path of a preceding aircraft.
@@
 The distance between aircraft pairs in view of safety is referred to as wake vortex separation. 
-
 \begin{figure}[h]
@@
     \includegraphics[width=0.8\textwidth]{graphics/reaction_in_wake.jpg}
-    \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft.
-(A) induced rolling moment, (B) upward motion, (C) loss of lift~\cite[p.~33]{Hallock2018Apr}} \label{fig:vortex_encounter}
+    \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft: induced rolling moment (A), upward motion (B), loss of lift (C)~\cite[p.~33]{Hallock2018Apr}} \label{fig:vortex_encounter}
 \end{figure}
 
-Traditional separation standards, introduced in the seventies, are defined by ICAO  on the basis of certificated maximum take-off mass (MTOM) as shown in Table~\ref{tab:WTC}.
+Traditional separation standards, introduced in the seventies, are defined by ICAO on the basis of certificated maximum take-off mass (MTOM) of aircraft (Table~\ref{tab:WTC}) and must be observed when the airport operates under instrument flight rules (IFR)~\cite{gerz_commercial_2002}.
+The IFR allow properly equipped aircraft to be flown under impaired visibility conditions~\cite{icao_annex_2005}.
 The prescribed categories are three i.e.
 Heavy, Medium and Light.
-In addition for aircraft in the order of $560000$~kg ICAO provides a subcategory to the Heavy types called Super Heavy.
-Some states have chosen to further increase the number of categories to adapt the wake turbulence categories (WTC) to specific airport requirements.
-Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing B757 into a separate category and increases the upper weight limit of the Light category.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
-
+The Heavy category includes aircraft with MTOM larger than $136.000$~kg.
+The aircraft in the Medium category have MTOM between $7.000$~and~$136.000$~kg and the Light category includes the aircraft with maximum take-off mass less than or equal to $7.000$~kg.
+In addition, for aircraft in the order of $560.000$~kg, ICAO provides a subcategory to the Heavy types called Super Heavy. \\
+Some states have chosen to further increase the number of the wake turbulence categories in light of safe airport operational experience.
+Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing B757 into a separate category and increases the upper weight limit of the Light category.
+The Boeing 757s are known to generate unusually high core vortex speed.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
 \begin{table}[h]
@@
 
-    ~    & A380-800    & $ > 136000 $  & \multicolumn{2}{c|}{$ 136000-7000 $ } & \multicolumn{2}{c}{$ \leq 7000 $} \\ \hline
+    ~    & A380-800    & > 136.000  & \multicolumn{2}{c|}{ 136.000 -- 7.000 } & \multicolumn{2}{c}{$ \leq 7.000$} \\ \hline
 
@@
 
-    ~    & A380-800    & $ > 136000 $  & $ 136000-104000 $     & $ 104000-40000 $      & $ 40000-17000 $ & $ \leq 17000 $   \\ \hline
+    ~    & A380-800    & > 136.000  & 136.000 -- 104.000     & 104.000 -- 40.000      & 40.000 -- 17.000 & $ \leq 17.000 $   \\ \hline
 
-    ICAO \& FAA (US)   & Super      & Heavy (H) & B757   & Large   & \multicolumn{2}{c}{Small} \\
+    FAA (US)   & Super      & Heavy & B757   & Large   & \multicolumn{2}{c}{Small} \\
 
-    ~    & A380        & $ > 136000 $  & ~                 & $ 136000-18600 $      & \multicolumn{2}{c}{$ \leq 18600 $} \\ 
+    ~    & A380        &  > 136.000   & ~                 &  136.000 -- 18.600       & \multicolumn{2}{c}{$ \leq 18.600 $} \\ 
     \end{tabular}}
-    \caption[ICAO wake turbulence categories based on maximum take-off mass]{ICAO wake turbulence categories based on maximum take-off mass (MTOM) in kg~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, kolos2013influence}.} \label{tab:WTC}
+    \caption[ICAO wake turbulence categories based on maximum take-off mass]{ICAO wake turbulence categories of transport aircraft based on maximum take-off mass (MTOM) in~kg~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, kolos2013influence}.} \label{tab:WTC}
 \end{table}

### 2019-01-08 introduction.tex
still on WTC and WTS explanations to make it understandable

@@
 ICAO has specified wake separation minima requirements that are applied to aircraft in the flight-path of another aircraft.
-Those requirements classify each aircraft in a wake turbulence category (WTC) and the wake separation is applied to a following aircraft based on the wake category of a leading aircraft.
+Those requirements classify each aircraft in a wake turbulence category (WTC) and the wake separation is applied to a following aircraft associated with the wake category of a leading aircraft.
 Thus the aircraft form a pair composed of a "leader" and a "follower", that observes a certain wake separation distance in nautical miles (NM).
@@
 Even though those separations are considered reliable, they are often conservative and outdated for newer aircraft models and as such generate unnecessary over-separation in some cases.
-In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} or just RECAT has been devised by EUROCONTROL~\cite{EUROCONTROL_who_is}, an intergovernmental organisation assisting in safe and efficient air traffic operations.
-RECAT is based on the ICAO wake turbulence categories, but takes into account the wake specifications of current aircraft models.
+In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} or just RECAT has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an intergovernmental organisation assisting in safe and efficient air traffic operations.
+RECAT is based on the ICAO wake turbulence longitudinal separation minima on approach and departure, but takes into account the wake specifications of current aircraft models.
 The RECAT scheme splits aircraft into six wake categories, generally referred to as CAT-A, CAT-B, CAT-C, CAT-D, CAT-E and CAT-F.
-Mostly, re-categorisation leads to decreased wake separation requirements for aircraft pairs, which in turn can lead to accommodating more landing and departures per time interval and increase airport capacity (throughput).\\
+Mostly, re-categorisation leads to decreased wake separation minima requirements for aircraft pairs, which in turn can lead to accommodating more landing and departures in peak traffic periods and increase airport capacity (throughput).\\
 Apart from wake separation, the other major element that affects capacity is runway occupancy time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
@@
 This section briefly reviews the necessity of adopting a wake vortex separation scheme for aircraft from the perspective of fluid dynamics.
-Wake vortex characteristics of conventional aircraft have been studied for several decades now and "it is generally acknowledged that the near wake of a large jet airliner poses a significant hazard to any smaller aircraft that follows it into of out of an air terminal"~\cite{dup._donamdson_vortex_1975}.
+Wake vortex characteristics of conventional aircraft have been studied for several decades now and "it is generally acknowledged that the near wake of a large jet airliner poses a significant hazard to any smaller aircraft that follows it into of out of an air terminal"~\cite[p.~5]{dup._donamdson_vortex_1975}.
 
@@
 Distance separation between aircraft is necessary to mitigate the effect of the wake vortices.
-This is especially true for aircraft in approach for landing or take-off, "since practically all serious problems associated with wake hazard occur in the vicinity of airports"~\cite{dup._donamdson_vortex_1975}.
+This is especially true for aircraft in approach for landing or take-off, "since practically all serious problems associated with wake hazard occur in the vicinity of airports"~\cite[p.~5]{dup._donamdson_vortex_1975}.
 The wake vortices generated by a leading aircraft (leader) can cause loss of lift or an induced rolling moment and velocity fluctuations in a following aircraft (follower) if the follower enters the wake turbulence region near the leader~(Figure~\ref{fig:vortex_encounter}).
@@
 
-Traditional separation standards, introduced in the seventies, are defined by ICAO on the basis of certificated maximum take-off mass (MTOM) of aircraft (Table~\ref{tab:WTC}) and must be observed when the airport operates under instrument flight rules (IFR)~\cite{gerz_commercial_2002}.
-The IFR allow properly equipped aircraft to be flown under impaired visibility conditions~\cite{icao_annex_2005}.
+Traditional separation standards, introduced in the seventies, are defined by ICAO on the basis of certificated maximum take-off mass (MTOM) of aircraft.
+The MTOM criteria allocates each aircraft in a wake turbulence category.
 The prescribed categories are three i.e.
@@
 The aircraft in the Medium category have MTOM between $7.000$~and~$136.000$~kg and the Light category includes the aircraft with maximum take-off mass less than or equal to $7.000$~kg.
-In addition, for aircraft in the order of $560.000$~kg, ICAO provides a subcategory to the Heavy types called Super Heavy. \\
+In addition, for aircraft in the order of $560.000$~kg, ICAO provides a subcategory to the Heavy types called Super Heavy (Table~\ref{tab:WTC}). \\
 Some states have chosen to further increase the number of the wake turbulence categories in light of safe airport operational experience.
-Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing B757 into a separate category and increases the upper weight limit of the Light category.
-The Boeing 757s are known to generate unusually high core vortex speed.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
-\begin{table}[h]
+Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing~B757 into a separate category and increases the upper weight limit of the Small (Light) category to $18.600$~kg.
+The Boeing~757s are known to generate unusually high core vortex speed.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
+\begin{table}[ht]
     \centering
@@
     \end{tabular}}
-    \caption[ICAO wake turbulence categories based on maximum take-off mass]{ICAO wake turbulence categories of transport aircraft based on maximum take-off mass (MTOM) in~kg~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, kolos2013influence}.} \label{tab:WTC}
+    \caption[ICAO wake turbulence categories based on maximum take-off mass]{ICAO wake turbulence categories of transport aircraft based on maximum take-off mass (MTOM) in~kg~\cite{doc4444full, uk_aeronautical_information_services_wake_2017, kolos2013influence}.} \label{tab:WTC}
 \end{table}
 
-The ICAO wake turbulence categories (Table~\ref{tab:ICAO_WTC}) outline the worst case for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018}.
+Under standard ICAO criteria, wake separation for each category, can be charted as a minimum distance separation matrix~(Table~\ref{tab:ICAO_WTC}).
+The The aircraft categories are paired as leader and follower \fxnote{figure out how to make this understandable, WTC to WTS}
+
+Those separation must be observed, for arrival and departure, when the airport operates under instrument flight rules (IFR).
+The IFR allow properly equipped aircraft to be flown under impaired visibility conditions.
+When visual flight rules (VFR) are in effect (flight visibility 5 km, clear of clouds and in sight of the surface), the separation may be decided by the pilot.~\cite{gerz_commercial_2002, icao_annex_2005} \\
+
+The ICAO wake turbulence categories  outline the worst case for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018, rooseleer2015recat}.
 The separation distances for instrumental flight rules are observed for the follower aircraft and are conventionally expressed in nautical miles (NM).

### 2019-01-09 introduction.tex
explained recat scheme

@@
 
-\section{Wake Vortex Separation}
+\section{Wake Turbulence Separation Minima}
 Distance separation between aircraft is necessary to mitigate the effect of the wake vortices.
 This is especially true for aircraft in approach for landing or take-off, "since practically all serious problems associated with wake hazard occur in the vicinity of airports"~\cite[p.~5]{dup._donamdson_vortex_1975}.
-The wake vortices generated by a leading aircraft (leader) can cause loss of lift or an induced rolling moment and velocity fluctuations in a following aircraft (follower) if the follower enters the wake turbulence region near the leader~(Figure~\ref{fig:vortex_encounter}).
+The wake vortices generated by a leading aircraft (leader) can cause loss of lift or an induced rolling moment and velocity fluctuations in a following aircraft (follower), if the follower enters the wake turbulence region near the leader~(Figure~\ref{fig:vortex_encounter}).
 This can lead to the up-set and potentially loss of control of an aircraft following the flight path of a preceding aircraft.
@@
 
-Traditional separation standards, introduced in the seventies, are defined by ICAO on the basis of certificated maximum take-off mass (MTOM) of aircraft.
+Traditional separation standards, introduced in the seventies, are defined by ICAO based on certificated maximum take-off mass (MTOM) of aircraft.
 The MTOM criteria allocates each aircraft in a wake turbulence category.
@@
 In addition, for aircraft in the order of $560.000$~kg, ICAO provides a subcategory to the Heavy types called Super Heavy (Table~\ref{tab:WTC}). \\
-Some states have chosen to further increase the number of the wake turbulence categories in light of safe airport operational experience.
-Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing~B757 into a separate category and increases the upper weight limit of the Small (Light) category to $18.600$~kg.
-The Boeing~757s are known to generate unusually high core vortex speed.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
 \begin{table}[ht]
@@
 
-Under standard ICAO criteria, wake separation for each category, can be charted as a minimum distance separation matrix~(Table~\ref{tab:ICAO_WTC}).
-The The aircraft categories are paired as leader and follower \fxnote{figure out how to make this understandable, WTC to WTS}
-
-Those separation must be observed, for arrival and departure, when the airport operates under instrument flight rules (IFR).
-The IFR allow properly equipped aircraft to be flown under impaired visibility conditions.
-When visual flight rules (VFR) are in effect (flight visibility 5 km, clear of clouds and in sight of the surface), the separation may be decided by the pilot.~\cite{gerz_commercial_2002, icao_annex_2005} \\
-
-The ICAO wake turbulence categories  outline the worst case for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018, rooseleer2015recat}.
-The separation distances for instrumental flight rules are observed for the follower aircraft and are conventionally expressed in nautical miles (NM).
-The separation reverts to radar separation minimum (MRS) when wake turbulence restrictions are not required.
-As  prescribed  by  ICAO  the  horizontal MRS is 3~NM (or a reduced separation minimum of 2.5~NM may be applied under given conditions described in ICAO Doc 4444 PANS-AM~\cite{doc44444}), which translates to 79~s (66~s) time separation between leader and follower at final approach speed of 70~m/s.
-
+Some states have chosen to further increase the number of the ICAO wake turbulence categories in light of safe airport operational experience.
+Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing~B757 into a separate category and increases the upper weight limit of the Small (Light) category to $18.600$~kg.
+The Boeing~757s are known to generate unusually high core vortex speed.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}\\
+Under standard ICAO criteria, each aircraft is allocated in a wake category and the categories form pairs: a combination of a leader and a follower aircraft.
+Each pair is assigned a distance separation minima, depending on the wake turbulence generated by the leader and applied to the follower.
+Conventionally the separations are expressed in nautical miles (NM).
+The leader-follower combinations can be charted as a wake separation matrix~(Table~\ref{tab:ICAO_WTC}), in which the elements indicate the minimum wake separation for each pair.\\
 % Please add the following required packages to your document preamble:
@@
 }
-\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter~\cite{noauthor_recat_2018, rooseleer2015recat}.} \label{tab:ICAO_WTC}
+\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter.(*) indicate radar separation minimum (MRS)~\cite{noauthor_recat_2018, rooseleer2015recat}.} \label{tab:ICAO_WTC}
+\end{table}
+
+The wake separation must be observed, for arrival and departure, when the airport operates under instrument flight rules (IFR).
+The IFR allow properly equipped aircraft to be flown under impaired visibility conditions.
+When visual flight rules (VFR) are in effect (flight visibility 5 km, clear of clouds and in sight of the surface), the separation may be decided by the pilot within observed safety limits.~\cite{gerz_commercial_2002, icao_annex_2005} \\
+When wake separation requirements are not specified and a surveillance system like radar is used, the longitudinal separation minimum may be reduced by the air traffic management (ATM), if the surveillance systems' capabilities at a given location permit this~\cite{MRS_separation_standard}.
+The radar separation minimum (MRS) for aircraft established on the same final approach track is 3~NM (or 2,5~NM under given conditions) as prescribed by ICAO Doc 4444 PANS-AM~\cite{doc44444}.\\
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=1\textwidth]{graphics/Criteria_RECAT.png}
+    \caption[RECAT-EU categorisation criteria]{Categorisation process and criteria for assigning an existing aircraft type into RECAT-EU scheme~\cite[p.~15]{rooseleer2015recat}} \label{fig:RECAT_criteria}
+\end{figure}
+The ICAO wake turbulence separation rules outline the "worst case" for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018, rooseleer2015recat}.
+The categorisation was implemented over 40 years ago and has become out-of-date in view of the newer aircraft models.
+Consequently EUROCONTROL has developed the European Wake Vortex Re-categorisation (RECAT-EU) as a more precise categorisation of aircraft compared to ICAO WTC.
+The objective was to safely help increase airport capacity and reduce delays by redefining wake turbulence categories and their associated separation
+minima. RECAT-EU is based on the ICAO scheme and takes into account maximum take-off weight (MTOW) and the wing span of the aircraft (Figure~\ref{fig:RECAT_criteria}).
+
+According to EUROCONTROL the expected immediate benefits from RECAT-EU deployment are: increased runway throughput around 5\% and operational efficiency during peak periods.
+The runway capacity increase with RECAT-EU "comes from the categorisation and separation reduction for aircraft types which are predominant in European traffic"~\cite[p.~22]{rooseleer2015recat}.
+Implementation of RECAT-EU will mean a minimum ATM system update and does not require deployment of new technologies.
+
+The resulting categories following the criteria for re-categorisation are six:\\
+\begin{tabular}{l}
+    \textbullet \space CAT A -- "Super Heavy" \\
+    \textbullet \space CAT B -- "Upper Heavy"\\
+    \textbullet \space CAT C -- "Lower Heavy"\\
+    \textbullet \space CAT D -- "Upper Medium"\\
+    \textbullet \space CAT E -- "Lower Medium"\\
+    \textbullet \space CAT F -- "Light"\\ 
+\end{tabular}
+
+The RECAT-EU wake separation matrix combining leader and follower into pairs with assigned wake separation minima for each pair is shown in Table~\ref{tab:RECAT-dist}.
+
+\begin{table}[h]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
+\hline
+\multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
+\multicolumn{2}{|c|}{}                                 & CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
+\multirow{6}{*}{\rotatebox[origin=c]{90}{Leader}}            & CAT-A            & 3 NM   & 4 NM  & 5 NM  & 5 NM  & 6 NM  & 8 NM  \\ \cline{2-8} 
+                                    & CAT-B            &    (*)    & 3 NM  & 4 NM  & 4 NM  & 5 NM  & 7 NM  \\ \cline{2-8} 
+                                    & CAT-C            &    (*)    & (*)   & 3 NM  & 3 NM  & 4 NM  & 6 NM  \\ \cline{2-8} 
+                                    & CAT-D            &    (*)    &   (*)    &    (*)   &   (*)    &   (*)    & 5 NM  \\ \cline{2-8} 
+                                    & CAT-E            &    (*)    &   (*)    &   (*)    &   (*)    &   (*)    & 4 NM  \\ \cline{2-8} 
+                                    & CAT-F            &    (*)    &   (*)    &    (*)   &   (*)    &   (*)    & 3 NM  \\ \hline
+\end{tabular}%
+}
+\caption[RECAT-EU distance-based separation minima]{RECAT-EU wake turbulence distance-based separation minima on approach and departure.
+(*) means minimum radar separation (MRS), set at 2.5 NM, applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat, noauthor_recat_2018}.}
+\label{tab:RECAT-dist}
 \end{table}
@@
 %This logo will scale according to the width of the text on the page.
-
-% Please add the following required packages to your document preamble:
-% \usepackage{multirow}
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-\multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
-\multicolumn{2}{|c|}{}                                 & CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
-\multirow{6}{*}{\rotatebox[origin=c]{90}{Leader}}            & CAT-A            & 3 NM   & 4 NM  & 5 NM  & 5 NM  & 6 NM  & 8 NM  \\ \cline{2-8} 
-                                    & CAT-B            &    (*)    & 3 NM  & 4 NM  & 4 NM  & 5 NM  & 7 NM  \\ \cline{2-8} 
-                                    & CAT-C            &    (*)    & (*)   & 3 NM  & 3 NM  & 4 NM  & 6 NM  \\ \cline{2-8} 
-                                    & CAT-D            &    (*)    &   (*)    &    (*)   &   (*)    &   (*)    & 5 NM  \\ \cline{2-8} 
-                                    & CAT-E            &    (*)    &   (*)    &   (*)    &   (*)    &   (*)    & 4 NM  \\ \cline{2-8} 
-                                    & CAT-F            &    (*)    &   (*)    &    (*)   &   (*)    &   (*)    & 3 NM  \\ \hline
-\end{tabular}%
-}
-\caption[RECAT-EU distance-based separation minima]{RECAT-EU wake turbulence distance-based separation minima on approach and departure.
-(*) means minimum radar separation (MRS), set at 2.5 NM, applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat, noauthor_recat_2018}.}
-\label{tab:RECAT-dist}
-\end{table}
 

### 2019-01-10 introduction.tex
re-organised  the structure of Intro

@@
 The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West).
-There has been an unforeseen increase of passengers travelling through Keflavík Airport in the recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
+There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
 For 2018 the passenger numbers show an yearly growth of 13.44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
@@
 Airport capacity is generally defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology}.
-\\
+
 A significant constraint of airport capacity is caused by distance separation regulations between aircraft in flight due to wake turbulence.
@@
 The European Aviation Safety Agency (EASA) provides oversight of those requirements and their application by the Air Traffic Management (ATM).
-\\
+
 The distance separation due to wake turbulence is especially critical during take-off and on approach for landing.
@@
 Generally the ICAO wake categories are three - Heavy, Medium and Light.
-The classification is based primarily on the weight of the aircraft, but this can vary slightly for different states.\\
+The classification is based primarily on the weight of the aircraft, but this can vary slightly for different states.
+
 The ICAO WTC categorisation is based on observations and calculations performed on aircraft models that are some 50 years old.
-Even though those separations are considered reliable, they are often conservative and outdated for newer aircraft models and as such generate unnecessary over-separation in some cases.
+Even though those separations are considered reliable, they are often conservative and outdated in view of newer aircraft models and as such generate unnecessary over-separation in some cases.
 In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} or just RECAT has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an intergovernmental organisation assisting in safe and efficient air traffic operations.
+
 RECAT is based on the ICAO wake turbulence longitudinal separation minima on approach and departure, but takes into account the wake specifications of current aircraft models.
 The RECAT scheme splits aircraft into six wake categories, generally referred to as CAT-A, CAT-B, CAT-C, CAT-D, CAT-E and CAT-F.
-Mostly, re-categorisation leads to decreased wake separation minima requirements for aircraft pairs, which in turn can lead to accommodating more landing and departures in peak traffic periods and increase airport capacity (throughput).\\
+Re-categorisation leads mostly to decreased wake separation minima requirements for aircraft pairs, which in turn can lead to accommodating more landing and departures in peak traffic periods and increase airport capacity (throughput).
+
 Apart from wake separation, the other major element that affects capacity is runway occupancy time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
 The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, per se, can limit the throughput of a runway.
-The emphasis in this project will be on runway occupancy times for arriving aircraft (AROT) during peak traffic hours at the airport\\
+The emphasis in this project will be on runway occupancy times for arriving aircraft (AROT) during peak traffic hours at the airport.
+The peak traffic hours are determined by the time intervals when the airport operates at high capacity load.
+
 The runway occupancy time and the wake separation requirements are the two main metrics that directly influence the throughput of a runway.
@@
 
-\section{Background and Literature Review}
-This section briefly reviews the necessity of adopting a wake vortex separation scheme for aircraft from the perspective of fluid dynamics.
-Wake vortex characteristics of conventional aircraft have been studied for several decades now and "it is generally acknowledged that the near wake of a large jet airliner poses a significant hazard to any smaller aircraft that follows it into of out of an air terminal"~\cite[p.~5]{dup._donamdson_vortex_1975}.
+\section{Runway Capacity\label{sec:runway_capacity}}
+
+ICAO generally defines airport capacity as the number of movements per unit of time that can be accepted during different meteorological conditions~\cite{airport_capacity_methodology}. 
+Several measures are currently used to estimate the amount of aircraft movements on the runways of an airport in a specified time interval.
+
+R.
+de Neufville~\cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure is maximum throughput capacity.
+It indicates the average number of movements that can be performed on the runway system in one hour in the presence of continuous demand, while adhering to all the separation requirements imposed by the (ATM) system~\cite{de_neufville_airport_2013}.
+
+The Capacity Envelope method used by Isavia to estimate the throughput capacity at BIKF takes into consideration the number of arrivals and departures within a selected time frame.
+This approach allows to measure the throughput capacity and model the maximum possible values or the potential of the airfield.
+A summary of the capacity envelope for a period of one year is shown in Figure~\ref{fig:capacity_evnelope} for a 15~minute time interval.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.8\textwidth]{graphics/fig_Capacity_Envelope_2017-10-04_to_2018-11-30_15min_occurrences_limits.png}
+    \caption[Capacity envelope for BIKF]{Capacity envelope at BIKF since October 2017.
+The measured time interval is 15 minutes and includes both the morning and afternoon peaks.
+The capacity limits of the airfield are also defined and noted on the figure for various sequence modes.
+The mixed modes indicate equal arrivals and departures (Mixed~Mode~A) or two times more arrivals than departures (Mixed Mode B), the latter being a tendency during the afternoon peaks.
+The figure is from Isavia's internal GUI Víkingaskipið.}  \label{fig:capacity_evnelope}
+\end{figure}
+
+This approach indicates a maximum measured capacity of four arrivals to four departures in a "Mixed~Mode~A".
+This equilibrium mode is defined as the primary measure for the throughput capacity of BIKF.
+The values are in accord with the defined peak hour intervals in section~\ref{sec:ROT}
+
+There are many aspects of a runway system that can affect the number of aircraft that can land or depart from an airfield and some of those are listed below~\cite{de_neufville_airport_2013, kim_validation_2010}.
+\begin{itemize}
+    \item Number and geometric layout of the runways.
+    \item State and performance of the ATM system.
+    \item Separation requirements between aircraft pairs imposed by the ATM.
+    \item Weather conditions (visibility, precipitation, cloud ceiling). 
+    \item Wind direction and strength.
+    \item Mix of aircraft using the airport.
+    \item Sequence of movement on each runway (arrivals -- departures split)
+    \item Type and location of taxiway exits from the runway.
+    \item Runway occupancy time.
+    \item Controller workload.
+    \item Noise-related constraints and other environmental considerations.
+\end{itemize}
+
+Two of the above mentioned factors are principal in determining airfield capacity, namely the wake vortex separation requirements and runway occupancy time~\cite{kolos2013influence}.
+Those are in turn influenced by traffic mix, seasonal factors and runway exits, to mention a few.
+The following Chapter~\ref{cha:methods} on Methods will deal with some of those aspects.
+
+\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur, Talk to Hjalti about the 2,5~-~3~NM MRS issue!}
+
+\section{Wake Turbulence Separation}
+This section briefly reviews the necessity of adopting a wake vortex separation scheme for aircraft from the perspective of fluid dynamics and the overall benefits of the RECAT-EU wake separation scheme in comparison to the ICAO WTC.
 
@@
 When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and are combined into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
-Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards.\\
+Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards.
+
 The effect of wake vortices on air masses is known as wake turbulence. 
@@
 
-\subsection{Ground Effect and Decay Process}
+\subsection{Ground Effect and Vortex Decay Rate}
 Wake vortices tend to descend slowly to an altitude of about one half of the initial separation, near ground on approach for landing.
@@
 This "re-bounce" effect is caused by the presence of the ground and carries with it the formation of a second pair of induced vortices outside and bellow the main vortex.
+
 In the presence of stable cross-wind conditions the decay of the downwind vortex is not identical to the upwind vortex and the descent rate may vary significantly~\cite{Hallock2018Apr}. 
@@
 Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$~m and $518$~m and in some cases even $762$~m, which is the basis for the "2500-foot" rule separation in parallel runway configurations~\cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
-The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5$~m/s for the first $30$~s after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m and below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}.
+
+The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere is shown to be around $1.5-2.5$~m/s for the first $30$~seconds after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m and below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}.
+
 Aerodynamic properties of the aircraft govern the vortex roll-up process and ambient atmospheric conditions dominate the behaviour of the vortices forcing instability and eventual decay of the vortices~\cite{Hallock2018Apr}.
@@
 
-\section{Wake Turbulence Separation Minima}
+Wake vortex characteristics of conventional aircraft have been studied for several decades now and "it is generally acknowledged that the near wake of a large jet airliner poses a significant hazard to any smaller aircraft that follows it into of out of an air terminal"~\cite[p.~5]{dup._donamdson_vortex_1975}.
 Distance separation between aircraft is necessary to mitigate the effect of the wake vortices.
@@
     \includegraphics[width=0.8\textwidth]{graphics/reaction_in_wake.jpg}
-    \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft: induced rolling moment (A), upward motion (B), loss of lift (C)~\cite[p.~33]{Hallock2018Apr}} \label{fig:vortex_encounter}
+    \caption[Wake vortex encounter]{Effect of wake vortex flow field on an aircraft: induced rolling moment (A), upward motion (B), loss of lift (C), yaw motion (D)~\cite[p.~33]{Hallock2018Apr}} 
+    \label{fig:vortex_encounter}
 \end{figure}
 
-Traditional separation standards, introduced in the seventies, are defined by ICAO based on certificated maximum take-off mass (MTOM) of aircraft.
+\subsection{Wake Categorisation}
+Traditional separation standards, introduced in the 1970s, are defined by ICAO based on certificated maximum take-off mass (MTOM) of aircraft.
 The MTOM criteria allocates each aircraft in a wake turbulence category.
@@
 The aircraft in the Medium category have MTOM between $7.000$~and~$136.000$~kg and the Light category includes the aircraft with maximum take-off mass less than or equal to $7.000$~kg.
-In addition, for aircraft in the order of $560.000$~kg, ICAO provides a subcategory to the Heavy types called Super Heavy (Table~\ref{tab:WTC}). \\
+In addition, for aircraft in the order of $560.000$~kg, ICAO provides a subcategory to the Heavy types called Super Heavy (Table~\ref{tab:WTC}).
+
 \begin{table}[ht]
@@
 Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing~B757 into a separate category and increases the upper weight limit of the Small (Light) category to $18.600$~kg.
-The Boeing~757s are known to generate unusually high core vortex speed.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}\\
+The Boeing~757s are known to generate unusually high core vortex speed.~\cite{icao_wtc, uk_aeronautical_information_services_wake_2017, noauthor_recat_2018}
+
 Under standard ICAO criteria, each aircraft is allocated in a wake category and the categories form pairs: a combination of a leader and a follower aircraft.
@@
 Conventionally the separations are expressed in nautical miles (NM).
-The leader-follower combinations can be charted as a wake separation matrix~(Table~\ref{tab:ICAO_WTC}), in which the elements indicate the minimum wake separation for each pair.\\
+The leader-follower combinations can be charted as a wake separation matrix~(Table~\ref{tab:ICAO_WTC}), in which the elements indicate the minimum wake separation for each pair.
+
 % Please add the following required packages to your document preamble:
@@
 The IFR allow properly equipped aircraft to be flown under impaired visibility conditions.
-When visual flight rules (VFR) are in effect (flight visibility 5 km, clear of clouds and in sight of the surface), the separation may be decided by the pilot within observed safety limits.~\cite{gerz_commercial_2002, icao_annex_2005} \\
-When wake separation requirements are not specified and a surveillance system like radar is used, the longitudinal separation minimum may be reduced by the air traffic management (ATM), if the surveillance systems' capabilities at a given location permit this~\cite{MRS_separation_standard}.
-The radar separation minimum (MRS) for aircraft established on the same final approach track is 3~NM (or 2,5~NM under given conditions) as prescribed by ICAO Doc 4444 PANS-AM~\cite{doc44444}.\\
+When visual flight rules (VFR) are in effect (flight visibility 5 km, clear of clouds and in sight of the surface), the separation may be decided by the pilot within observed safety limits.~\cite{gerz_commercial_2002, icao_annex_2005}
+
+When wake separation requirements are not specified and a surveillance system like radar is used, the longitudinal separation minimum may be reduced by the air traffic management (ATM), if the surveillance system's capabilities at a given location permit this~\cite{MRS_separation_standard}.
+The radar separation minimum (MRS) for aircraft established on the same final approach track is 3~NM (or 2,5~NM under given conditions) as prescribed by ICAO Doc 4444 PANS-AM~\cite{doc44444}.
 
@@
 \end{figure}
+
 The ICAO wake turbulence separation rules outline the "worst case" for each category and thus generate over-separation in many cases~\cite{noauthor_recat_2018, rooseleer2015recat}.
@@
 
-The resulting categories following the criteria for re-categorisation are six:\\
+The resulting categories following the re-categorisation criteria from Figure~\ref{fig:RECAT_criteria} are six:
+
 \begin{tabular}{l}
@@
 The RECAT-EU wake separation matrix combining leader and follower into pairs with assigned wake separation minima for each pair is shown in Table~\ref{tab:RECAT-dist}.
+The application of instrumental or visual flight rules and the MRS procedures remain identical to the ICAO WTC scheme.
+The introduction of RECAT-EU alters only the spacing between
+certain aircraft in a pair, which is applied to arrivals as well as departures.
 
@@
 \caption[RECAT-EU distance-based separation minima]{RECAT-EU wake turbulence distance-based separation minima on approach and departure.
-(*) means minimum radar separation (MRS), set at 2.5 NM, applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat, noauthor_recat_2018}.}
+(*) indicates minimum radar separation (MRS), set at 3~NM (2,5~NM under given conditions), applicable as per current ICAO doc 4444 provisions \cite{doc44444, rooseleer2015recat, noauthor_recat_2018}.}
 \label{tab:RECAT-dist}
@@
 
-It must be noted that the division of ICAO WTC into the RECAT scheme (Table~\ref{tab:RECAT-dist}) is not achieved by splitting each category in half.
-The criteria used for categorisation of existing and new aircraft types are provided in detail in~\cite{rooseleer2015recat}.
-The case from the traffic mix at BIKF presents the arrangement shown in Table~\ref{tab:wtc2recat_division}.
-A more detailed illustration of the re-categorisation is presented in the upcoming section \ref{ssec:traffic_mix} on aircraft traffic mix (Figure~\ref{fig:post_fast_exit_mix_pie_v2}).
+The noticeable changes in wake separation under RECAT-EU from ICAO WTC are systematised in Table~\ref{tab:delta_distance_wtc2recat}.
+A beneficial change in wake separation experiences for example a Heavy leader -- Medium follower aircraft pair, which after re-categorisation form a CAT-C -- CAT-D pair.
+In this case, the separation from the leader can be reduced by up to 2~NM.
 
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-% \usepackage[table,xcdraw]{xcolor}
-% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
 \begin{table}[h]
 \centering
-\resizebox{0.7\textwidth}{!}{%
-\begin{tabular}{clcllc}
-\multicolumn{3}{c}{\cellcolor[HTML]{34CDF9}HEAVY} &  &  & \cellcolor[HTML]{FD6864}LIGHT \\ \hline
-\multicolumn{1}{|c|}{\cellcolor[HTML]{34CDF9}CAT-A} & \multicolumn{1}{c|}{\cellcolor[HTML]{34CDF9}CAT-B} & \multicolumn{1}{c|}{\cellcolor[HTML]{32CB00}CAT-C} & \multicolumn{1}{c|}{\cellcolor[HTML]{F8FF00}CAT-D} & \multicolumn{1}{c|}{\cellcolor[HTML]{F8FF00}CAT-E} & \multicolumn{1}{c|}{\cellcolor[HTML]{FFC702}CAT-F} \\ \hline
-\multicolumn{1}{l}{} &  & \multicolumn{4}{c}{\cellcolor[HTML]{F8FF00}MEDIUM}
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
+\hline
+\multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                                                                                                                   \\ \cline{3-8}
+\multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A                             & CAT-B                                & CAT-C                         & CAT-D                         & CAT-E                         & CAT-F                                                \\ \hline
+                                                        & CAT-A & \cellcolor[HTML]{FD6864}(+0,5 NM) & \cellcolor[HTML]{67FD9A}-2 NM        & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
+                                                        & CAT-B &                                   & \cellcolor[HTML]{67FD9A}-1 NM        &                               & \cellcolor[HTML]{67FD9A}-1 NM &                               & \cellcolor[HTML]{FD6864}+1 NM                        \\ \cline{2-8}
+                                                        & CAT-C &                                   & \cellcolor[HTML]{67FD9A}-1 (-1,5) NM & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
+                                                        & CAT-D &                                   &                                      &                               &                               &                               &                                                      \\ \cline{2-8}
+                                                        & CAT-E &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{67FD9A}{\color[HTML]{000000} -1 NM} \\ \cline{2-8}
+\multirow{-6}{*}{\rotatebox[origin=c]{90}{Leader}}                                & CAT-F &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{FD6864}(+0,5 NM)                    \\ \hline
 \end{tabular}%
 }
-\caption[Transition from ICAO WTC to RECAT-EU categories]{Transition from ICAO WTC to RECAT-EU categories.
-The categorisation process and criteria for assigning an existing aircraft type into RECAT-EU scheme is illustrated in detail in~\cite{rooseleer2015recat}}.
-\label{tab:wtc2recat_division}
+\caption[Difference in wake separation minima between ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)~\cite{rooseleer2015recat}}
+\label{tab:delta_distance_wtc2recat}
 \end{table}
 
-The noticeable changes in Table~\ref{tab:RECAT-dist} from Table~\ref{tab:ICAO_WTC} affect a follower aircraft, whose relative position from the ICAO scheme is shifted to the left in the new table after re-categorisation.
-Such benefit experiences for example a Heavy follower aircraft that is re-categorised as CAT-C, in which case the separation from the leader is reduced with up to 3 NM.
-
-%There is an RU logo in Figure~\ref{fig:ru-logo}.
-%This logo will scale according to the width of the text on the page.
-
-The different category code names in this study will also be referred to on a single letter basis for simplification e.g.
+The different category code names in this study will be referred to on a single letter basis for simplification e.g.
 "CAT-C" will be abbreviated to "C", "CAT-D" to "D", "Heavy" to "H", Medium to "M" and so forth.
-This applies to pair categories by abbreviating e.g.
-"CAT-C leader - CAT-D follower"  to "C-D pair" or simply "C-D", and "Medium-Medium" pair to "M-M".
+This applies to pair categories as well by abbreviating e.g.
+"CAT-C leader -- CAT-D follower"  to "C-D pair" or just "C-D", and "Medium--Medium" pair to "M-M".
 
@@
 
-The Runway Occupancy Time (ROT) is defined by Eurocontrol as the amount of time that each aircraft occupies the runway~\cite{ROT_definition}. \\
+The Runway Occupancy Time (ROT) is defined by EUROCONTROL as the length of time that each aircraft occupies the runway~\cite{ROT_definition}. 
 This project focuses on Arrivals Runway Occupancy Time (AROT) in peak traffic hours, that is the time interval between the aircraft crossing the threshold and its tail vacating the runway~\cite{AROT_definition}.
@@
 
-\section{Runway Capacity\label{sec:runway_capacity}}
-
-ICAO generally defines airport capacity as the number of movements per unit of time that can be accepted during different meteorological conditions~\cite{airport_capacity_methodology}. 
-Several measures are currently used to estimate the amount of aircraft movements on the runways of an airport in a specified time interval.
-R.
-de Neufville~\cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure is maximum throughput capacity.
-It indicates the average number of movements that can be performed on the runway system in one hour in the presence of continuous demand, while adhering to all the separation requirements imposed by the air traffic management (ATM) system~\cite{de_neufville_airport_2013}.
-The Capacity Envelope method used by Isavia to estimate the throughput capacity at BIKF takes into consideration the number of arrivals and departures within a selected time frame.
-This approach allows to measure the throughput capacity and model the maximum possible values or the potential of the airfield.
-A summary of the capacity envelope for a period of one year is shown in Figure~\ref{fig:capacity_evnelope} for a 15~minute time interval.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.8\textwidth]{graphics/fig_Capacity_Envelope_2017-10-04_to_2018-11-30_15min_occurrences_limits.png}
-    \caption[Capacity envelope for BIKF]{Capacity envelope at BIKF since October 2017.
-The measured time interval is 15 minutes and includes both the morning and afternoon peaks.
-The capacity limits of the airfield are also defined and noted on the figure for various sequence modes.
-The mixed modes indicate equal arrivals and departures (Mixed~Mode~A) or two times more arrivals than departures (Mixed Mode B), the later being a tendency during the afternoon peaks.
-The figure is from Isavia´s internal GUI Víkingaskipið.}  \label{fig:capacity_evnelope}
-\end{figure}
-
-This approach indicates a maximum measured capacity of four arrivals to four departures in a "Mixed~Mode~A".
-This equilibrium mode is defined as the primary measure for the throughput capacity of BIKF.
-The values are in accord with the defined peak hour intervals in the previous section~\ref{sec:ROT}
-
-There are many aspects of a runway system that can affect the number of aircraft that can land or depart from an airfield and some of those are listed bellow~\cite{de_neufville_airport_2013, kim_validation_2010}.
-\begin{itemize}
-    \item Number and geometric layout of the runways.
-    \item State and performance of the ATM system.
-    \item Separation requirements between aircraft pairs imposed by the ATM.
-    \item Weather conditions (visibility, precipitation, cloud ceiling). 
-    \item Wind direction and strength.
-    \item Mix of aircraft using the airport.
-    \item Sequence of movement on each runway (arrivals - departures split)
-    \item Type and location of taxiway exits from the runway.
-    \item Runway occupancy time.
-    \item Controller workload.
-    \item Noise-related constraints and other environmental considerations.
-\end{itemize}
-
-Two of the above mentioned factors are principal in determining airfield capacity, namely the wake vortex separation requirements and runway occupancy~\cite{kolos2013influence}.
-Those are in turn influenced by traffic mix, seasonal factors and runway exits, to mention a few.
-The following chapter~\ref{cha:methods} on Methods  will deal with some of those aspects.
-
-\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur, Talk to Hjalti about the 2,5~-~3~NM MRS issue!}
-
 \section{Runway Occupancy and Landing Time Interval Study Objective}\label{sec:study_objective}
@@
 The proposed transition by Eurocontrol from ICAO WTS to RECAT-EU scheme suggests a decrease in distance separation for a number of the aircraft pairs (Table~\ref{tab:delta_distance_wtc2recat}).
-
-% Please add the following required packages to your document preamble:
-% \usepackage{multirow}
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-\multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                                                                                                                   \\ \cline{3-8}
-\multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A                             & CAT-B                                & CAT-C                         & CAT-D                         & CAT-E                         & CAT-F                                                \\ \hline
-                                                        & CAT-A & \cellcolor[HTML]{FD6864}(+0,5 NM) & \cellcolor[HTML]{67FD9A}-2 NM        & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
-                                                        & CAT-B &                                   & \cellcolor[HTML]{67FD9A}-1 NM        &                               & \cellcolor[HTML]{67FD9A}-1 NM &                               & \cellcolor[HTML]{FD6864}+1 NM                        \\ \cline{2-8}
-                                                        & CAT-C &                                   & \cellcolor[HTML]{67FD9A}-1 (-1,5) NM & \cellcolor[HTML]{67FD9A}-1 NM & \cellcolor[HTML]{67FD9A}-2 NM & \cellcolor[HTML]{67FD9A}-1 NM &                                                      \\ \cline{2-8}
-                                                        & CAT-D &                                   &                                      &                               &                               &                               &                                                      \\ \cline{2-8}
-                                                        & CAT-E &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{67FD9A}{\color[HTML]{000000} -1 NM} \\ \cline{2-8}
-\multirow{-6}{*}{\rotatebox[origin=c]{90}{Leader}}                                & CAT-F &                                   &                                      &                               &                               &                               & \cellcolor[HTML]{FD6864}(+0,5 NM)                    \\ \hline
-\end{tabular}%
-}
-\caption[Difference in wake separation minima between ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)~\cite{rooseleer2015recat}}
-\label{tab:delta_distance_wtc2recat}
-\end{table}
-
-% % Please add the following required packages to your document preamble:
-% % \usepackage{multirow}
-% % \usepackage{graphicx}
-% \begin{table}[]
-% \centering
-% \resizebox{\textwidth}{!}{%
-% \begin{tabular}{|c|c|c|c|c|c|c|c|}
-% \hline
-% \multicolumn{2}{|c|}{}                                          & \multicolumn{6}{c|}{Follower}                                                                                                             \\ \cline{3-8}
-% \multicolumn{2}{|c|}{\multirow{-2}{*}{RECAT-EU scheme}} & CAT-A & CAT-B                        & CAT-C & CAT-D                        & CAT-E                        & CAT-F                        \\ \hline
-%                                                         & CAT-A &       & \cellcolor[HTML]{67FD9A}-20s &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8}
-%                                                         & CAT-B &       &                              &       & \cellcolor[HTML]{67FD9A}-20s &                              & \cellcolor[HTML]{FD6864}+20s \\ \cline{2-8}
-%                                                         & CAT-C &       &                              &       & \cellcolor[HTML]{67FD9A}-40s & \cellcolor[HTML]{67FD9A}-20s &                              \\ \cline{2-8}
-%                                                         & CAT-D &       &                              &       &                              &                              &                              \\ \cline{2-8}
-%                                                         & CAT-E &       &                              &       &                              &                              & \cellcolor[HTML]{67FD9A}-20s \\ \cline{2-8}
-% \multirow{-6}{*}{Leader}                                & CAT-F &       &                              &       &                              &                              & \cellcolor[HTML]{FD6864}+20s \\ \hline
-% \end{tabular}
-% }
-% \caption[Time difference between reference ICAO and RECAT-EU schemes]{Time difference in wake separation on departure between reference ICAO and RECAT-EU schemes(full proposal) \cite{rooseleer2015recat}}
-% \label{tab:delta_times_wtc2recat}
-% \end{table}
 

### 2019-01-11 appendix_1.tex
some changes on Methods, mainly grammar and syntax .

--- 2019-01-04 appendix_1.tex
@@
-\chapter{Runway Usage}\label{cha:RWY_usage}
+\chapter{Runway Usage}\label{app:RWY_usage}
 
@@
 The summer traffic favoured RWY-19 with 52.8\%~(Figure~\ref{fig:runway_usage_summer}), while during the winter season some of those arrivals were diverted towards RWY-01 and RWY-10~(Figure~\ref{fig:runway_usage_summer}).
-RWY-28 was least used despite its fast-exit TWY B-1.}
+RWY-28 was least used despite its rapid-exit TWY B-1.}
     \label{fig:runway_usage}
@@
 \clearpage
-\chapter{Arrival Runway Occupancy Times\label{cha:AROTs}}
+\chapter{Arrival Runway Occupancy Times\label{app:AROTs}}
 
@@
 }
-\caption[AROTs for the air traffic mix by RECAT]{AROT statistics for the air traffic mix at BIKF by RECAT-EU categories. The count is the number of landings in peak hours since October 2017}
+\caption[AROTs for the air traffic mix by RECAT]{AROT statistics for the air traffic mix at BIKF by RECAT-EU categories. The count is the number of landings during peak hours since October 2017}
 \label{tab:AROT_RECAT_stats}
@@
 }
-\caption[AROTs RWY-01 pre fast exit by season]{AROTs for runway RWY-01 pre-fast-exit TWY A-1 by season}
+\caption[AROTs RWY-01 pre rapid-exit by season]{AROTs for runway RWY-01 pre-rapid-exit TWY A-1 by season}
 \label{tab:season_AROT_stats_RWY01_pre_fast_exit}
@@
 }
-\caption[AROTs RWY-01 post fast exit by season]{AROTs for runway RWY-01 post-fast-exit TWY A-1 by season}
+\caption[AROTs RWY-01 post rapid-exit by season]{AROTs for runway RWY-01 post-rapid-exit TWY A-1 by season}
 \label{tab:season_AROT_stats_RWY01_post_fast_exit}
@@
 }
-\caption[AROTs RWY-28 pre fast exit by season]{AROTs for runway RWY-28 pre-fast-exit TWY B-1 by season}
+\caption[AROTs RWY-28 pre rapid-exit by season]{AROTs for runway RWY-28 pre-rapid-exit TWY B-1 by season}
 \label{tab:season_AROT_stats_RWY28_pre_fast_exit}
@@
 }
-\caption[AROTs RWY-28 post fast exit by season]{AROTs for runway RWY-28 post-fast-exit TWY B-1 by season}
+\caption[AROTs RWY-28 post rapid-exit by season]{AROTs for runway RWY-28 post-rapid-exit TWY B-1 by season}
 \label{tab:season_AROT_stats_RWY28_post_fast_exit}
@@
 }
-\caption[AROTs for the air traffic mix by runway for the summer]{AROT statistics for the air traffic mix at KEF by runway for the summer of 2018. The count is the number of landings in peak hours.}
+\caption[AROTs for the air traffic mix by runway for the summer]{AROT statistics for the air traffic mix at KEF by runway for the summer of 2018.
+The count is the number of landings during peak hours.}
 \label{my-label3}
@@
 \caption[AROTs for the air traffic mix by runway for the winter]{AROT statistics for the air traffic mix at KEF by runway for the winter season.
-The count is the number of landings in peak hours since October 2017}
+The count is the number of landings during peak hours since October 2017}
 \label{my-label4}
@@
 \caption[AROTs for the air traffic mix by month]{AROT statistics for the air traffic mix at BIKF by month.
-The count is the number of landings in peak hours from October 2017 til November 2018.
+The count is the number of landings during peak hours from October 2017 til November 2018.
 The colour fields indicate a subjective separation of the data into summer and winter season, based on mean AROT value.
@@
 \clearpage
-\chapter{Inter-arrival time and distance separation}
+\chapter{Inter-arrival time and distance separation}\label{app:time_and_dist_sep}
 
@@
     \includegraphics[width=0.5\textwidth]{graphics/fig_HM_to_DD_pairs_dist_separ.png}
-    \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs to D-D pairs into the RECAT scheme.}
+    \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT-EU scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs to D-D pairs into the RECAT-EU scheme.}
     \label{fig:HM_to_DD_pairs_dist_separ}
@@
     \caption[AROT and LTI of C-C pairs from ICAO H-M pairs on RWY~19]{AROT and landing time intervals of C-C pairs originating from ICAO H-M pairs for RWY-19 during the summer season.
-The dashed red line indicates the RECAT reference time separation for the C-C pairs.}
+The dashed red line indicates the RECAT-EU reference time separation for the C-C pairs.}
     \label{fig:best_CC_from_HM_pairs_time_sep}
@@
 \clearpage
-\chapter{BIKF Schematic}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.8\textwidth]{graphics/BIKF_schematic.png}
-    \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways \cite{BIKF_scheme}.}
-    \label{fig:BIKF_schematic}
-\end{figure}
-
-\clearpage
-\chapter{BIKF Wind Rose}
+\chapter{BIKF Wind Rose}\label{app:wind_rose}
 

### 2019-01-11 introduction.tex
some changes on Methods, mainly grammar and syntax .

@@
 There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
-For 2018 the passenger numbers show an yearly growth of 13.44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
+For 2018 the passenger numbers show a yearly growth of 13.44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
 The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods.
@@
 The distance separation due to wake turbulence is especially critical during take-off and on approach for landing.
-ICAO has specified wake separation minima requirements that are applied to aircraft in the flight-path of another aircraft.
-Those requirements classify each aircraft in a wake turbulence category (WTC) and the wake separation is applied to a following aircraft associated with the wake category of a leading aircraft.
-Thus the aircraft form a pair composed of a "leader" and a "follower", that observes a certain wake separation distance in nautical miles (NM).
+ICAO has specified wake turbulence separation minima requirements that are applied to aircraft in the flight-path of another aircraft.
+Those requirements classify each aircraft in a wake turbulence category (WTC) and the wake turbulence separation is applied to a following aircraft associated with the wake category of a leading aircraft.
+Thus the aircraft form a pair composed of a "leader" and a "follower", that observe a certain spacing in nautical miles (NM).
 Generally the ICAO wake categories are three - Heavy, Medium and Light.
@@
 Even though those separations are considered reliable, they are often conservative and outdated in view of newer aircraft models and as such generate unnecessary over-separation in some cases.
-In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} or just RECAT has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an intergovernmental organisation assisting in safe and efficient air traffic operations.
+In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an intergovernmental organisation assisting in safe and efficient air traffic operations.
 
-RECAT is based on the ICAO wake turbulence longitudinal separation minima on approach and departure, but takes into account the wake specifications of current aircraft models.
-The RECAT scheme splits aircraft into six wake categories, generally referred to as CAT-A, CAT-B, CAT-C, CAT-D, CAT-E and CAT-F.
-Re-categorisation leads mostly to decreased wake separation minima requirements for aircraft pairs, which in turn can lead to accommodating more landing and departures in peak traffic periods and increase airport capacity (throughput).
+RECAT-EU is based on the ICAO wake turbulence longitudinal separation minima on approach and departure, but takes into account the wake specifications of current aircraft models.
+The RECAT-EU scheme splits aircraft into six wake categories, generally referred to as CAT-A, CAT-B, CAT-C, CAT-D, CAT-E and CAT-F.
+Re-categorisation leads mostly to decreased wake turbulence separation minima requirements for aircraft pairs, which in turn can lead to accommodating more landing and departures during peak traffic periods and increase airport capacity (throughput).
 
-Apart from wake separation, the other major element that affects capacity is runway occupancy time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
-The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, per se, can limit the throughput of a runway.
+Apart from wake turbulence separation, the other major factor that affects capacity is runway occupancy time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
+The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, per se, does limit the throughput of a runway.
 The emphasis in this project will be on runway occupancy times for arriving aircraft (AROT) during peak traffic hours at the airport.
-The peak traffic hours are determined by the time intervals when the airport operates at high capacity load.
+The peak traffic hours are determined by the time intervals when the airport operates at high capacity loads.
 
-The runway occupancy time and the wake separation requirements are the two main metrics that directly influence the throughput of a runway.
-The aim of this study is to estimate the constraints for implementing the RECAT wake turbulence categorisation and separation minima for Keflavík International Airport, the effect this can produce for airfield throughput and when is the runway occupancy a limiting factor for increased capacity.
+The arrival runway occupancy time and the wake turbulence separation requirements are the two main metrics that directly influence throughput.
+The wake turbulence distance separation is transferred into the time domain and defined as the landing time interval (LTI), to be able to compare the two metrics and their ranges.
+The landing time interval quantifies the time separation between aircraft in a pair (derived from the distance separation between aircraft in a pair and the final approach speed of the follower).
+
+The aim of this study is to estimate the constraints for implementing the RECAT-EU wake turbulence categorisation and separation minima for Keflavík International Airport, the effect this can produce for airfield throughput and when is the runway occupancy a limiting factor for increased capacity.
 
@@
     \includegraphics[width=0.8\textwidth]{graphics/fig_Capacity_Envelope_2017-10-04_to_2018-11-30_15min_occurrences_limits.png}
-    \caption[Capacity envelope for BIKF]{Capacity envelope at BIKF since October 2017.
+    \caption[Capacity envelope for BIKF]{Capacity envelope at BIKF since October 2017 during peak hours.
 The measured time interval is 15 minutes and includes both the morning and afternoon peaks.
 The capacity limits of the airfield are also defined and noted on the figure for various sequence modes.
-The mixed modes indicate equal arrivals and departures (Mixed~Mode~A) or two times more arrivals than departures (Mixed Mode B), the latter being a tendency during the afternoon peaks.
+The mixed modes indicate: equal arrivals and departures (Mixed~Mode~A) or two times more arrivals than departures (Mixed Mode B), the latter being a tendency during the afternoon peaks.
 The figure is from Isavia's internal GUI Víkingaskipið.}  \label{fig:capacity_evnelope}
@@
 This equilibrium mode is defined as the primary measure for the throughput capacity of BIKF.
-The values are in accord with the defined peak hour intervals in section~\ref{sec:ROT}
+The values are in accord with the defined peak hour intervals in section~\ref{sec:arot_and_study_objective}
 
@@
 
-\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur, Talk to Hjalti about the 2,5~-~3~NM MRS issue!}
+\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur!}
 
 \section{Wake Turbulence Separation}
-This section briefly reviews the necessity of adopting a wake vortex separation scheme for aircraft from the perspective of fluid dynamics and the overall benefits of the RECAT-EU wake separation scheme in comparison to the ICAO WTC.
+This section briefly reviews the necessity of adopting a wake vortex separation scheme for aircraft from the perspective of fluid dynamics and the overall benefits of the RECAT-EU wake turbulence separation scheme in comparison to the ICAO WTC.
 
@@
 Upon reaching this point the descent will stop and then ascend slowly.
-This "re-bounce" effect is caused by the presence of the ground and carries with it the formation of a second pair of induced vortices outside and bellow the main vortex.
+This "re-bounce" effect is caused by the presence of the ground and carries with it the formation of a second pair of induced vortices outside and below the main vortex.
 
@@
 The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure~\ref{fig:vortex_ground_effect}).
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.8\textwidth]{graphics/Hallock_vortex_evolution.jpg}
+    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect~\cite[p.~29]{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
+\end{figure}
+
 Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$~m and $518$~m and in some cases even $762$~m, which is the basis for the "2500-foot" rule separation in parallel runway configurations~\cite{Hallock2018Apr, hallock2004summary, hallock2003wake}.
-
-The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere is shown to be around $1.5-2.5$~m/s for the first $30$~seconds after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m and below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}.
+The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere is shown to be around $1.5-2.5$~m/s for the first $30$~seconds after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}.
 
@@
 \end{itemize}
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.8\textwidth]{graphics/Hallock_vortex_evolution.jpg}
-    \caption[Wake vortex and ground effect]{Vortex evolution and ground effect~\cite[p.~29]{Hallock2018Apr}.} \label{fig:vortex_ground_effect}
-\end{figure}
 
 Wake vortex characteristics of conventional aircraft have been studied for several decades now and "it is generally acknowledged that the near wake of a large jet airliner poses a significant hazard to any smaller aircraft that follows it into of out of an air terminal"~\cite[p.~5]{dup._donamdson_vortex_1975}.
+
 Distance separation between aircraft is necessary to mitigate the effect of the wake vortices.
@@
 The wake vortices generated by a leading aircraft (leader) can cause loss of lift or an induced rolling moment and velocity fluctuations in a following aircraft (follower), if the follower enters the wake turbulence region near the leader~(Figure~\ref{fig:vortex_encounter}).
-This can lead to the up-set and potentially loss of control of an aircraft following the flight path of a preceding aircraft.
-To diminish the effect of such vortices the trailing aircraft must be maintained at a safe distance behind the leading aircraft as the vortices spread laterally to either side of the flight path and are dissipated~\cite{Breitsamter2011Feb}.
-The distance between aircraft pairs in view of safety is referred to as wake vortex separation. 
+
 \begin{figure}[h]
@@
 \end{figure}
+
+This can lead to the up-set and potentially loss of control of an aircraft following the flight path of a preceding aircraft.
+To diminish the effect of such vortices the trailing aircraft must be maintained at a safe distance behind the leading aircraft as the vortices spread laterally to either side of the flight path and are dissipated~\cite{Breitsamter2011Feb}.
+The distance between aircraft pairs in view of safety is referred to as wake vortex separation.
 
@@
 Conventionally the separations are expressed in nautical miles (NM).
-The leader-follower combinations can be charted as a wake separation matrix~(Table~\ref{tab:ICAO_WTC}), in which the elements indicate the minimum wake separation for each pair.
+The leader-follower combinations can be charted as a wake turbulence separation matrix~(Table~\ref{tab:ICAO_WTC}), in which the elements indicate the minimum spacing for each pair.
 
@@
 
-The wake separation must be observed, for arrival and departure, when the airport operates under instrument flight rules (IFR).
+The wake turbulence separation must be observed, for arrival and departure, when the airport operates under instrument flight rules (IFR).
 The IFR allow properly equipped aircraft to be flown under impaired visibility conditions.
@@
 
-When wake separation requirements are not specified and a surveillance system like radar is used, the longitudinal separation minimum may be reduced by the air traffic management (ATM), if the surveillance system's capabilities at a given location permit this~\cite{MRS_separation_standard}.
+When wake turbulence separation requirements are not specified and a surveillance system like radar is used, the longitudinal separation minimum may be reduced by the air traffic management (ATM), if the surveillance system's capabilities at a given location permit this~\cite{MRS_separation_standard}.
 The radar separation minimum (MRS) for aircraft established on the same final approach track is 3~NM (or 2,5~NM under given conditions) as prescribed by ICAO Doc 4444 PANS-AM~\cite{doc44444}.
@@
 
-The RECAT-EU wake separation matrix combining leader and follower into pairs with assigned wake separation minima for each pair is shown in Table~\ref{tab:RECAT-dist}.
+The RECAT-EU wake turbulence separation matrix combining leader and follower into pairs with assigned minimum spacing for each pair is shown in Table~\ref{tab:RECAT-dist}.
 The application of instrumental or visual flight rules and the MRS procedures remain identical to the ICAO WTC scheme.
@@
 
-The noticeable changes in wake separation under RECAT-EU from ICAO WTC are systematised in Table~\ref{tab:delta_distance_wtc2recat}.
-A beneficial change in wake separation experiences for example a Heavy leader -- Medium follower aircraft pair, which after re-categorisation form a CAT-C -- CAT-D pair.
+The noticeable changes in wake turbulence separation under RECAT-EU from ICAO WTC are systematised in Table~\ref{tab:delta_distance_wtc2recat}.
+A beneficial change in wake turbulence separation experiences for example a Heavy leader -- Medium follower aircraft pair, which after re-categorisation form a CAT-C -- CAT-D pair.
 In this case, the separation from the leader can be reduced by up to 2~NM.
@@
 }
-\caption[Difference in wake separation minima between ICAO and RECAT-EU schemes]{Difference in wake separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)~\cite{rooseleer2015recat}}
+\caption[Difference in wake turbulence separation minima between ICAO and RECAT-EU schemes]{Difference in wake turbulence separation minima on approach between reference ICAO and RECAT-EU schemes (full proposal)~\cite{rooseleer2015recat}}
 \label{tab:delta_distance_wtc2recat}
@@
 
-\section{Runway Occupancy Time\label{sec:ROT}}
+\section{Arrival Runway Occupancy Time and Landing Time Interval}\label{sec:arot_and_study_objective}
 
-The Runway Occupancy Time (ROT) is defined by EUROCONTROL as the length of time that each aircraft occupies the runway~\cite{ROT_definition}. 
-This project focuses on Arrivals Runway Occupancy Time (AROT) in peak traffic hours, that is the time interval between the aircraft crossing the threshold and its tail vacating the runway~\cite{AROT_definition}.
+The runway occupancy time (ROT) is defined by EUROCONTROL as the length of time that each aircraft occupies the runway~\cite{ROT_definition}. 
+This project focuses on arrivals runway occupancy time (AROT) during peak traffic hours, that is the time interval defined by an aircraft crossing the threshold and its tail vacating the runway~\cite{AROT_definition}.
 The threshold is the beginning of that portion of the runway that is available for landing.
-The peak traffic hours are characterised by the time intervals when the airport operates at high load.
+
+The peak traffic hours are characterised by the time intervals when the airport operates at high loads.
 High load interval or window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavík Airport.
@@
 The last flight in a high load window that fulfils these requirements is not considered to be a part of the peak as its behaviour is not affected by an instantaneous next flight after it.
-The AROT metric is essential for the project because a following aircraft is not allowed to land on the same runway before it has been vacated by the leading aircraft.
-The interval to the next landing aircraft is specified as the Landing Time Interval (LTI) which is directly linked to the inter-arrival distance or in other words the wake turbulence separation.
 
@@
 
-\section{Runway Occupancy and Landing Time Interval Study Objective}\label{sec:study_objective}
-The goal of examining the arrival runway occupancy times and the landing time intervals (LTI) is to determine whether the implementation of the RECAT-EU scheme at BIKF under certain conditions, would affect the throughput of the airfield.
-LTI is the a measure of the time separation between aircraft pairs derived from the distance separation between the aircraft and the final approach speed (time travelled is distance travelled divided by velocity).
-% The LTI is directly influenced by the distance separation of arrival pairs and the final approach speed of the aircraft .
-Additionally the analysis could indicate whether the limiting element is the runway occupancy or the separation requirement.
-The proposed transition by Eurocontrol from ICAO WTS to RECAT-EU scheme suggests a decrease in distance separation for a number of the aircraft pairs (Table~\ref{tab:delta_distance_wtc2recat}).
+The AROT metric is essential for the project because a following aircraft is not allowed to land on the same runway before it has been vacated by the leading aircraft.
+The interval to the next landing aircraft is specified as the landing time interval (LTI) which is directly linked to the inter-arrival distance or in other words the wake turbulence separation.
+The LTI is a metric used to quantify the time separation between aircraft in a pair, derived from the distance separation between the aircraft and the final approach speed of the following aircraft.% (time travelled is distance travelled divided by velocity).
 
-The implementation of RECAT shifts the wake separation minima for arrival and departure pairs thus creating the potential for reduced time separation between the aircraft.
-The reduction of the time interval between landings facilitates increase in the throughput capacity of the runways.
-Figure~\ref{fig:AROT_LTI_rwy19_H_M} may be used to illustrate the hypothesis.
-The current ICAO time reference separation, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT scheme.
-This relocation will be more noticeable for RECAT C-C and C-D pairs formed from the ICAO HEAVY-MEDIUM pairs, in which case the separation is reduced from 5 NM to 3 NM.
-A shift of the reference separation line creates the potential for shift in the distribution of the LTI, as long as it does not overlap with the ROT.
-Furthermore study on the runway capacity of significantly larger airports~\cite{kolos2013influence} suggests that the frequency distribution of LTI tends to compress or squeeze to the right of the reference line with less standard deviation about the mean when the air traffic is intensified.
+The goal of examining the arrival runway occupancy times and the landing time intervals is to determine whether the implementation of the RECAT-EU scheme at BIKF under certain conditions, would affect the throughput of the airfield.
+Additionally the analysis could indicate whether the limiting factor is the runway occupancy time or the separation requirement.
+The proposed transition by EUROCONTROL from ICAO WTS to RECAT-EU scheme suggests a decrease in distance separation for a certain aircraft pairs (Table~\ref{tab:delta_distance_wtc2recat}).
+
+The implementation of RECAT-EU shifts the wake turbulence separation minima for arrival pairs, creating the potential for reduced spacing between aircraft, hence reducing the time interval between landings (LTI) and facilitating increased throughput capacity of the runways.
 
@@
 
+Figure~\ref{fig:AROT_LTI_rwy19_H_M} may be used to illustrate the hypothesis.
+The current ICAO time reference separation, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT-EU scheme.
+This relocation will be more noticeable for C-C and C-D pairs formed from the ICAO H-M pairs, in which case the separation is reduced from 5 NM to 3 NM.
+A shift of the reference separation line creates the potential for a shift in the distribution of the LTI, as long as it does not overlap with the AROT.
+Furthermore, study on the runway capacity of significantly larger airports~\cite{kolos2013influence} suggests that the frequency distribution of LTI tends to compress or squeeze to the right of the reference line with less standard deviation about the mean when the air traffic is intensified.
+
 %%% Local Variables: 

### 2019-01-13 introduction.tex
minor changes

@@
     \item State and performance of the ATM system.
-    \item Separation requirements between aircraft pairs imposed by the ATM.
+    \item Wake Separation requirements between aircraft pairs imposed by the ATM.
     \item Weather conditions (visibility, precipitation, cloud ceiling). 
@@
 
-Two of the above mentioned factors are principal in determining airfield capacity, namely the wake vortex separation requirements and runway occupancy time~\cite{kolos2013influence}.
+Two of the above mentioned factors are principal in determining airfield capacity, namely the wake turbulence separation requirements and runway occupancy time~\cite{kolos2013influence}.
 Those are in turn influenced by traffic mix, seasonal factors and runway exits, to mention a few.
@@
 \section{Wake Turbulence Separation}
-This section briefly reviews the necessity of adopting a wake vortex separation scheme for aircraft from the perspective of fluid dynamics and the overall benefits of the RECAT-EU wake turbulence separation scheme in comparison to the ICAO WTC.
+This section briefly reviews the necessity of adopting a wake turbulence separation scheme for aircraft from the perspective of fluid dynamics and the overall benefits of the RECAT-EU wake turbulence separation scheme in comparison to the ICAO WTC.
 

### 2019-01-15 introduction.tex
finished data explanation and started traffic mix fixes

@@
 
-Keflavík International Airport (IATA: KEF, ICAO: BIKF) is the major gateway to Iceland currently with two functioning runways.
-The runways operate in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West).
+Keflavík International Airport (IATA: KEF, ICAO: BIKF) is the major gateway to Iceland and is serviced by Isavia~\cite{Isavia_about}.
+The airport currently has two functioning runways operating in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West).
 There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
@@
 The peak traffic hours are characterised by the time intervals when the airport operates at high loads.
-High load interval or window is at least 15 minutes in length and has four or more flights arriving or departing from Keflavík Airport.
+High load interval or window is at least 15 minutes in length and has five or more flights arriving or departing from Keflavík Airport.
 The time between flights during the high load interval is set at $\leq$4~minutes (Figure~\ref{fig:Peak_Diagram}).
-This value is based on statistical data by Isavia in order to achieve two distinct peak hours: one in the morning and one in the afternoon.
+This value is based on statistical estimations by Isavia in order to achieve two distinct peak hours: one in the morning and one in the afternoon.
 The last flight in a high load window that fulfils these requirements is not considered to be a part of the peak as its behaviour is not affected by an instantaneous next flight after it.

### 2019-01-17 appendix_1.tex
started fixing Methods
fixed up to seasonal on AROT

@@
 \chapter{Arrival Runway Occupancy Times\label{app:AROTs}}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
-                          & \multicolumn{1}{c|}{}   & \multicolumn{7}{c|}{AROT [s]} \\ \hline
-\multicolumn{1}{|l|}{RECAT-EU} & count & mean & std & min & 25\% & 50\% & 75\% & max \\ \hline
-\multicolumn{1}{|l|}{CAT-A}    & 2     & 119  & 19  & 105 & 112  & 119  & 125  & 132 \\ \hline
-\multicolumn{1}{|l|}{CAT-B}    & 53    & 81   & 20  & 51  & 65   & 77   & 91   & 136 \\ \hline
-\multicolumn{1}{|l|}{CAT-C}    & 3436  & 78   & 17  & 46  & 65   & 75   & 89   & 155 \\ \hline
-\multicolumn{1}{|l|}{CAT-D}    & 1029  & 75   & 18  & 46  & 62   & 71   & 87   & 155 \\ \hline
-\multicolumn{1}{|l|}{CAT-E}    & 95   & 88   & 24  & 46  & 71   & 85   & 101  & 151 \\ \hline
-\multicolumn{1}{|l|}{CAT-F}    & 49    & 84   & 22  & 48  & 72   & 81   & 91   & 155 \\ \hline
-\end{tabular}%
-}
-\caption[AROTs for the air traffic mix by RECAT]{AROT statistics for the air traffic mix at BIKF by RECAT-EU categories. The count is the number of landings during peak hours since October 2017}
-\label{tab:AROT_RECAT_stats}
-\end{table}
 
@@
 \clearpage
+\chapter{ADS-B Standard Data Items}\label{app:adsb_items}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.7\textwidth]{graphics/ads_b_items.png}
+    \caption[ADS-B Standard Data Items]{ADS-B Standard Data Items \cite[p. 8]{ASTERIX_ADS-B_specs}.}
+    \label{fig:adsb_items}
+\end{figure}
+
+\clearpage
 \chapter{BIKF Wind Rose}\label{app:wind_rose}

### 2019-01-17 introduction.tex
fixed up to seasonal on AROT

@@
 
-\subsection{Wake Categorisation}
+\subsection{Wake Turbulence Categorisation}
 Traditional separation standards, introduced in the 1970s, are defined by ICAO based on certificated maximum take-off mass (MTOM) of aircraft.
@@
 Additionally the analysis could indicate whether the limiting factor is the runway occupancy time or the separation requirement.
-The proposed transition by EUROCONTROL from ICAO WTS to RECAT-EU scheme suggests a decrease in distance separation for a certain aircraft pairs (Table~\ref{tab:delta_distance_wtc2recat}).
-
-The implementation of RECAT-EU shifts the wake turbulence separation minima for arrival pairs, creating the potential for reduced spacing between aircraft, hence reducing the time interval between landings (LTI) and facilitating increased throughput capacity of the runways.
+The proposed transition by EUROCONTROL from ICAO WTS to RECAT-EU scheme suggests reduced distance separation for certain aircraft pairs (Table~\ref{tab:delta_distance_wtc2recat}).
+The reduction is reveled by shifting the wake turbulence separation minima for arrival pairs, creating the potential for reduced spacing between aircraft, hence reducing the time interval between landings (LTI) and facilitating increased throughput capacity of the runways.
 

### 2019-01-21 appendix_1.tex
fixed numbers and figures to them from BIKF traffic db, added operator
fixed seasonal and usage to account for rwy with rapid-exit on one hand and without rapid exit ot the other

@@
 \cline{3-9}
-\multicolumn{2}{l|}{RWY01}  & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
-\multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{764} & 85 & 20 & 46 & 70 & 86 & 99 & 153 \\ \hline
-\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{263} & 100 & 17 & 55 & 87 & 98 & 111 & 156 \\ \hline
-\end{tabular}%
-}
-\caption[AROTs RWY-01 pre rapid-exit by season]{AROTs for runway RWY-01 pre-rapid-exit TWY A-1 by season}
-\label{tab:season_AROT_stats_RWY01_pre_fast_exit}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
-\multicolumn{2}{l|}{RWY01, TWY A-1}  & \multicolumn{7}{c|}{AROT  {[}s{]}} \\ \hline
-\multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{1879} & 84 & 22 & 46 & 64 & 86 & 98 & 152 \\ \hline
-\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{284} & 88 & 26 & 47 & 63 & 90 & 106 & 153 \\ \hline
-\end{tabular}%
-}
-\caption[AROTs RWY-01 post rapid-exit by season]{AROTs for runway RWY-01 post-rapid-exit TWY A-1 by season}
-\label{tab:season_AROT_stats_RWY01_post_fast_exit}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
-\multicolumn{2}{l|}{RWY28} & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
-\multicolumn{1}{|l|}{Season} & count & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & \multicolumn{1}{r|}{553} & 99 & 12 & 46 & 91 & 98 & 106 & 153 \\ \hline
-\multicolumn{1}{|l|}{Winter} & \multicolumn{1}{r|}{249} & 110 & 13 & 60 & 100 & 108 & 117 & 155 \\ \hline
-\end{tabular}%
-}
-\caption[AROTs RWY-28 pre rapid-exit by season]{AROTs for runway RWY-28 pre-rapid-exit TWY B-1 by season}
-\label{tab:season_AROT_stats_RWY28_pre_fast_exit}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
-\multicolumn{2}{l|}{RWY28, TWY B-1} & \multicolumn{7}{c|}{AROT {[}s{]}} \\ \hline
-\multicolumn{1}{|l|}{Season} & {count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{Summer} & 401 & 68 & 14 & 49 & 60 & 66 & 72 & 155 \\ \hline
-\multicolumn{1}{|l|}{Winter} & 69 & 73 & 13 & 49 & 63 & 72 & 79 & 105 \\ \hline
-\end{tabular}%
-}
-\caption[AROTs RWY-28 post rapid-exit by season]{AROTs for runway RWY-28 post-rapid-exit TWY B-1 by season}
-\label{tab:season_AROT_stats_RWY28_post_fast_exit}
-\end{table}
-
-% Please add the following required packages to your document preamble:
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{lr|r|r|r|r|r|r|r|}
-\cline{3-9}
                              & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
@@
 \multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
-\multicolumn{1}{|l|}{RWY 01} & 284 & 88 & 26 & 47 & 63 & 90 & 106 & 153 \\ \hline
-\multicolumn{1}{|l|}{RWY 10} & 333 & 93 & 12 & 70 & 85 & 91 & 99 & 144 \\ \hline
-\multicolumn{1}{|l|}{RWY 19} & 251 & 72 & 12 & 51 & 65 & 70 & 75 & 140 \\ \hline
+\multicolumn{1}{|l|}{RWY 01} & 290 & 88 & 26 & 47 & 63 & 90 & 106 & 153 \\ \hline
+\multicolumn{1}{|l|}{RWY 10} & 338 & 93 & 12 & 70 & 85 & 91 & 100 & 144 \\ \hline
+\multicolumn{1}{|l|}{RWY 19} & 259 & 72 & 13 & 51 & 65 & 69 & 75 & 140 \\ \hline
 \multicolumn{1}{|l|}{RWY 28} & 69 & 73 & 13 & 49 & 63 & 72 & 79 & 105 \\ \hline
@@
 \rowcolor[HTML]{DAE8FC} 
-\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}November} & 220 & 84 & 23 & 48 & 66 & 82 & 99 & 151 \\ \hline
+\multicolumn{1}{|l|}{\cellcolor[HTML]{DAE8FC}November} & 239 & 84 & 23 & 48 & 66 & 82 & 99 & 151 \\ \hline
 \rowcolor[HTML]{DAE8FC} 

### 2019-01-23 introduction.tex
fixes of inter-arrival times and LTI

@@
 The interval to the next landing aircraft is specified as the landing time interval (LTI) which is directly linked to the inter-arrival distance or in other words the wake turbulence separation.
-The LTI is a metric used to quantify the time separation between aircraft in a pair, derived from the distance separation between the aircraft and the final approach speed of the following aircraft.% (time travelled is distance travelled divided by velocity).
+The LTI is a metric used to quantify the time separation between aircraft in a pair, derived from the distance separation between the aircraft and the final approach speed of the following aircraft.
+One way the conflict between the AROT and LTI can manifest itself is through a missed approach (ICAO: bulked landing), when an aircraft landing cannot be completed due to runway incursion \cite{doc44444}: the incorrect presence of another aircraft on the runway designated for landing.% (time travelled is distance travelled divided by velocity).
 

### 2019-01-24 appendix_1.tex
fixing icao to recat pairs distribution and procents

@@
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.5\textwidth]{graphics/fig_HM_to_DD_pairs_dist_separ.png}
-    \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT-EU scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs to D-D pairs into the RECAT-EU scheme.}
-    \label{fig:HM_to_DD_pairs_dist_separ}
-\end{figure}
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=0.5\textwidth]{graphics/fig_HM_to_DD_pairs_dist_separ.png}
+%     \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT-EU scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs to D-D pairs into the RECAT-EU scheme.}
+%     \label{fig:HM_to_DD_pairs_dist_separ}
+% \end{figure}
 

### 2019-01-24 introduction.tex
fixing icao to recat pairs distribution and procents

@@
 There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
-For 2018 the passenger numbers show a yearly growth of 13.44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
+For 2018 the passenger numbers show a yearly growth of 13,44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
 The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods.

### 2019-02-01 introduction.tex
fixing LTI

@@
 The proposed transition by EUROCONTROL from ICAO WTS to RECAT-EU scheme suggests reduced distance separation for certain aircraft pairs (Table~\ref{tab:delta_distance_wtc2recat}).
-The reduction is reveled by shifting the wake turbulence separation minima for arrival pairs, creating the potential for reduced spacing between aircraft, hence reducing the time interval between landings (LTI) and facilitating increased throughput capacity of the runways.
+This reduction, revealed by shifting the wake turbulence separation minima for arrival pairs, creates potential for reduced spacing between aircraft, hence reducing the time interval between landings (LTI) and facilitating increased throughput capacity of the runways.
 

### 2019-02-04 introduction.tex
fixes on LTI and 2 std on LTI AROT plots

@@
 The following Chapter~\ref{cha:methods} on Methods will deal with some of those aspects.
-
-\fxnote{equipment specifications in the tower that might influence the WTC separtarion decision-talk to Haraldur!}
 
@@
 Figure~\ref{fig:AROT_LTI_rwy19_H_M} may be used to illustrate the hypothesis.
-The current ICAO time reference separation, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT-EU scheme.
+The current ICAO time reference separation, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT-EU scheme, following a decreased wake separation requirement.
 This relocation will be more noticeable for C-C and C-D pairs formed from the ICAO H-M pairs, in which case the separation is reduced from 5 NM to 3 NM.

### 2019-02-12 appendix_1.tex
fixes on interarrival distances
fixing constraints to capacity from arot and lti

@@
 \clearpage
-\chapter{Inter-arrival time and distance separation}\label{app:time_and_dist_sep}
+\chapter{Landing time interval and AROT}\label{app:lti_and_arot}
 
@@
 \end{table}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=1.0\textwidth]{graphics/fig_best_CC_from_HM_pairs_time_sep.png}
-    \caption[AROT and LTI of C-C pairs from ICAO H-M pairs on RWY~19]{AROT and landing time intervals of C-C pairs originating from ICAO H-M pairs for RWY-19 during the summer season.
-The dashed red line indicates the RECAT-EU reference time separation for the C-C pairs.}
-    \label{fig:best_CC_from_HM_pairs_time_sep}
-\end{figure}
 

### 2019-02-14 appendix_1.tex
fixes  of capacity constraints

@@
 
-\clearpage
-\chapter{Landing time interval and AROT}\label{app:lti_and_arot}
+% \clearpage
+% \chapter{Landing time interval and AROT}\label{app:lti_and_arot}
 
@@
 % \usepackage{multirow}
-% \usepackage{graphicx}
-\begin{table}[h]
-\centering
-\resizebox{0.8\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-\multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
-\multicolumn{2}{|c|}{}& CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
-\multirow{6}{*}{\rotatebox[origin=c]{90}{Leader}}& CAT-A&& 100s  & 120s  & 140s  & 160s  & 180s  \\ \cline{2-8}
-& CAT-B&&&& 100s  & 120s  & 140s  \\ \cline{2-8} 
-& CAT-C&&&& 80s   & 100s  & 120s  \\ \cline{2-8} 
-& CAT-D&&&&&& 120s  \\ \cline{2-8} 
-& CAT-E&&&&&& 100s  \\ \cline{2-8} 
-& CAT-F&&&&&& 80s   \\ \hline
-\end{tabular}%
-}
-\caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure~\cite{rooseleer2015recat}}
-\label{tab:RECAT-time}
-\end{table}
+% % \usepackage{graphicx}
+% \begin{table}[h]
+% \centering
+% \resizebox{0.8\textwidth}{!}{%
+% \begin{tabular}{|c|c|c|c|c|c|c|c|}
+% \hline
+% \multicolumn{2}{|c|}{\multirow{2}{*}{RECAT-EU scheme}} & \multicolumn{6}{c|}{Follower}                   \\ \cline{3-8} 
+% \multicolumn{2}{|c|}{}& CAT-A & CAT-B & CAT-C & CAT-D & CAT-E & CAT-F \\ \hline
+% \multirow{6}{*}{\rotatebox[origin=c]{90}{Leader}}& CAT-A&& 100s  & 120s  & 140s  & 160s  & 180s  \\ \cline{2-8}
+% & CAT-B&&&& 100s  & 120s  & 140s  \\ \cline{2-8} 
+% & CAT-C&&&& 80s   & 100s  & 120s  \\ \cline{2-8} 
+% & CAT-D&&&&&& 120s  \\ \cline{2-8} 
+% & CAT-E&&&&&& 100s  \\ \cline{2-8} 
+% & CAT-F&&&&&& 80s   \\ \hline
+% \end{tabular}%
+% }
+% \caption[RECAT-EU time-based separation minima]{RECAT-EU WT time-based separation minima on approach and departure~\cite{rooseleer2015recat}}
+% \label{tab:RECAT-time}
+% \end{table}
 

### 2019-02-15 introduction.tex
LTI and reference times fixes
fixes on constraints

@@
 \centering
-\resizebox{\textwidth}{!}{%
+\resizebox{.8\textwidth}{!}{%
 \begin{tabular}{|c|c|c|c|c|c|}
@@
 \centering
-\resizebox{\textwidth}{!}{%
+\resizebox{.8\textwidth}{!}{%
 \begin{tabular}{|c|c|c|c|c|c|c|c|}
@@
 \centering
-\resizebox{\textwidth}{!}{%
+\resizebox{.9\textwidth}{!}{%
 \begin{tabular}{|c|c|c|c|c|c|c|c|}
@@
 The current ICAO time reference separation, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT-EU scheme, following a decreased wake separation requirement.
-This relocation will be more noticeable for C-C and C-D pairs formed from the ICAO H-M pairs, in which case the separation is reduced from 5 NM to 3 NM.
+This relocation will be more noticeable for C-C and C-D pairs formed from the ICAO H-M pairs, in which case the separation is reduced from 5~NM to 3~NM.
 A shift of the reference separation line creates the potential for a shift in the distribution of the LTI, as long as it does not overlap with the AROT.

### 2019-04-11 introduction.tex
fixing figure positions within the text
upper case fixes of abbreviations + some  'Arman fixes

@@
 For 2018 the passenger numbers show a yearly growth of 13,44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
-The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods.
-Airport capacity is generally defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology}.
+The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods, here airport capacity is defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology}.
 
@@
 ICAO has specified wake turbulence separation minima requirements that are applied to aircraft in the flight-path of another aircraft.
-Those requirements classify each aircraft in a wake turbulence category (WTC) and the wake turbulence separation is applied to a following aircraft associated with the wake category of a leading aircraft.
-Thus the aircraft form a pair composed of a "leader" and a "follower", that observe a certain spacing in nautical miles (NM).
+Those requirements classify each aircraft in a Wake Turbulence Category (WTC) and the wake turbulence separation is applied to a following aircraft associated with the wake category of a leading aircraft.
+Thus the aircraft form a pair composed of a "leader" and a "follower", that observe a certain spacing in Nautical Miles (NM).
 Generally the ICAO wake categories are three - Heavy, Medium and Light.
-The classification is based primarily on the weight of the aircraft, but this can vary slightly for different states.
+The classification is based primarily on the weight of the aircraft, but this can vary slightly for different countries.
 
@@
 
-Apart from wake turbulence separation, the other major factor that affects capacity is runway occupancy time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
+Apart from wake turbulence separation, the other major factor that affects capacity is Runway Occupancy Time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
 The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, per se, does limit the throughput of a runway.
@@
 The arrival runway occupancy time and the wake turbulence separation requirements are the two main metrics that directly influence throughput.
-The wake turbulence distance separation is transferred into the time domain and defined as the landing time interval (LTI), to be able to compare the two metrics and their ranges.
+The wake turbulence distance separation is transferred into the time domain and defined as the Landing Time Interval (LTI), to be able to compare the two metrics and their ranges.
 The landing time interval quantifies the time separation between aircraft in a pair (derived from the distance separation between aircraft in a pair and the final approach speed of the follower).
@@
 
-R.
-de Neufville~\cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure is maximum throughput capacity.
+Richard de Neufville~\cite{de_neufville_airport_2013} states that the capacity of runway systems determines the ultimate capacity of an airport and that its principal measure is maximum throughput capacity.
 It indicates the average number of movements that can be performed on the runway system in one hour in the presence of continuous demand, while adhering to all the separation requirements imposed by the (ATM) system~\cite{de_neufville_airport_2013}.
@@
 \subsection{Wake Turbulence Categorisation}
-Traditional separation standards, introduced in the 1970s, are defined by ICAO based on certificated maximum take-off mass (MTOM) of aircraft.
+Traditional separation standards, introduced in the 1970s, are defined by ICAO based on certificated Maximum Take-Off Mass (MTOM) of aircraft.
 The MTOM criteria allocates each aircraft in a wake turbulence category.
@@
 
-Some states have chosen to further increase the number of the ICAO wake turbulence categories in light of safe airport operational experience.
+Some countries have chosen to further increase the number of the ICAO wake turbulence categories in light of safe airport operational experience.
 Aerodromes in the UK split the Medium and Light categories each into two subcategories (Table~\ref{tab:WTC}), while the Federal Aviation Administration (FAA) places Boeing~B757 into a separate category and increases the upper weight limit of the Small (Light) category to $18.600$~kg.
@@
 Each pair is assigned a distance separation minima, depending on the wake turbulence generated by the leader and applied to the follower.
-Conventionally the separations are expressed in nautical miles (NM).
+Conventionally the separations are expressed in Nautical Miles (NM).
 The leader-follower combinations can be charted as a wake turbulence separation matrix~(Table~\ref{tab:ICAO_WTC}), in which the elements indicate the minimum spacing for each pair.
@@
 }
-\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter.(*) indicate radar separation minimum (MRS)~\cite{noauthor_recat_2018, rooseleer2015recat}.} \label{tab:ICAO_WTC}
+\caption[ICAO wake turbulence categories and separation minima]{ICAO wake turbulence categories and separation minima to avoid wake vortex encounter.(*) indicate Radar Separation Minimum (MRS)~\cite{noauthor_recat_2018, rooseleer2015recat}.} \label{tab:ICAO_WTC}
 \end{table}
 
-The wake turbulence separation must be observed, for arrival and departure, when the airport operates under instrument flight rules (IFR).
+The wake turbulence separation must be observed, for arrival and departure, when the airport operates under Instrument Flight Rules (IFR).
 The IFR allow properly equipped aircraft to be flown under impaired visibility conditions.
-When visual flight rules (VFR) are in effect (flight visibility 5 km, clear of clouds and in sight of the surface), the separation may be decided by the pilot within observed safety limits.~\cite{gerz_commercial_2002, icao_annex_2005}
+When Visual Flight Rules (VFR) are in effect (flight visibility 5 km, clear of clouds and in sight of the surface), the separation may be decided by the pilot within observed safety limits.~\cite{gerz_commercial_2002, icao_annex_2005}
 
-When wake turbulence separation requirements are not specified and a surveillance system like radar is used, the longitudinal separation minimum may be reduced by the air traffic management (ATM), if the surveillance system's capabilities at a given location permit this~\cite{MRS_separation_standard}.
-The radar separation minimum (MRS) for aircraft established on the same final approach track is 3~NM (or 2,5~NM under given conditions) as prescribed by ICAO Doc 4444 PANS-AM~\cite{doc44444}.
+When wake turbulence separation requirements are not specified and a surveillance system like radar is used, the longitudinal separation minimum may be reduced by the Air Traffic Management (ATM), if the surveillance system's capabilities at a given location permit this~\cite{MRS_separation_standard}.
+The Radar Separation Minimum (MRS) for aircraft established on the same final approach track is 3~NM (or 2,5~NM under given conditions) as prescribed by ICAO Doc 4444 PANS-AM~\cite{doc44444}.
 
@@
 The objective was to safely help increase airport capacity and reduce delays by redefining wake turbulence categories and their associated separation
-minima. RECAT-EU is based on the ICAO scheme and takes into account maximum take-off weight (MTOW) and the wing span of the aircraft (Figure~\ref{fig:RECAT_criteria}).
+minima. RECAT-EU is based on the ICAO scheme and takes into account Maximum Take-Off Weight (MTOW) and the wing span of the aircraft (Figure~\ref{fig:RECAT_criteria}).
 
@@
 
-The runway occupancy time (ROT) is defined by EUROCONTROL as the length of time that each aircraft occupies the runway~\cite{ROT_definition}. 
-This project focuses on arrivals runway occupancy time (AROT) during peak traffic hours, that is the time interval defined by an aircraft crossing the threshold and its tail vacating the runway~\cite{AROT_definition}.
+The Runway Occupancy Time (ROT) is defined by EUROCONTROL as the length of time that each aircraft occupies the runway~\cite{ROT_definition}. 
+This project focuses on Arrival Runway Occupancy Time (AROT) during peak traffic hours, that is the time interval defined by an aircraft crossing the threshold and its tail vacating the runway~\cite{AROT_definition}.
 The threshold is the beginning of that portion of the runway that is available for landing.
@@
 The AROT metric is essential for the project because a following aircraft is not allowed to land on the same runway before it has been vacated by the leading aircraft.
-The interval to the next landing aircraft is specified as the landing time interval (LTI) which is directly linked to the inter-arrival distance or in other words the wake turbulence separation.
+The interval to the next landing aircraft is specified as the Landing Time Interval (LTI) which is directly linked to the inter-arrival distance or in other words the wake turbulence separation.
 The LTI is a metric used to quantify the time separation between aircraft in a pair, derived from the distance separation between the aircraft and the final approach speed of the following aircraft.
@@
 
-Figure~\ref{fig:AROT_LTI_rwy19_H_M} may be used to illustrate the hypothesis.
+The hypothesis may be illustrated with Figure~\ref{fig:AROT_LTI_rwy19_H_M}.
 The current ICAO time reference separation, indicated by the red vertical line for BIKF RWY-19 will be relocated to the left for some of the aircraft pairs under the RECAT-EU scheme, following a decreased wake separation requirement.

### 2019-04-25 introduction.tex
second revision fixes up to data manipulation

@@
 
-Keflavík International Airport (IATA: KEF, ICAO: BIKF) is the major gateway to Iceland and is serviced by Isavia~\cite{Isavia_about}.
-The airport currently has two functioning runways operating in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West).
-There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
-For 2018 the passenger numbers show a yearly growth of 13,44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}.
-The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods, here airport capacity is defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology}.
+Keflavík International Airport (IATA: KEF, ICAO: BIKF) is the major gateway to Iceland and is operated by Isavia~\cite{Isavia_about}, a shareholding company solely owned by the Icelandic state.
+The airport currently has two functioning runways operating in both directions, designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West).
+There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over 9,8~million passengers and 97.432 air transport movements in 2018~\cite{isavia_facts_2017}.
+Where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}.
+For 2018 the passenger numbers show a yearly growth of 12.0\% from the previous year~\cite{isavia_pass_statistics_2018}.
+The increase in traffic at Keflavík Airport does experience constrained capacity of the runways during peak traffic periods, here airport capacity is defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology} without causing delays on the ground or in the air.
 
 A significant constraint of airport capacity is caused by distance separation regulations between aircraft in flight due to wake turbulence.
-The term "wake turbulence" describes the effect of the rotating air masses generated behind the wing tips of large jet aircraft~\cite{doc4444full}.
+The term "wake turbulence" describes the effect of the rotating air masses generated behind the wins of large jet aircraft~\cite{doc4444full}.
 A smaller aircraft can experience loss of control in close proximity behind a larger aircraft because of those swirling air masses.
 The general flight safety requirements for maintaining safe distance between aircraft are established by the International Civil Aviation Organisation (ICAO)~\cite{doc4444full}.
-The European Aviation Safety Agency (EASA) provides oversight of those requirements and their application by the Air Traffic Management (ATM).
-
+The European Aviation Safety Agency (EASA) provides the specific definition and oversight of those requirements and their application by Air Traffic Management (ATM) providers in the European Economic Area (EAA).
 The distance separation due to wake turbulence is especially critical during take-off and on approach for landing.
 ICAO has specified wake turbulence separation minima requirements that are applied to aircraft in the flight-path of another aircraft.
-Those requirements classify each aircraft in a Wake Turbulence Category (WTC) and the wake turbulence separation is applied to a following aircraft associated with the wake category of a leading aircraft.
+Those requirements classify each aircraft into a Wake Turbulence Category (WTC) and the wake turbulence separation is applied to a following aircraft associated with the wake category of a leading aircraft.
 Thus the aircraft form a pair composed of a "leader" and a "follower", that observe a certain spacing in Nautical Miles (NM).
 Generally the ICAO wake categories are three - Heavy, Medium and Light.
-The classification is based primarily on the weight of the aircraft, but this can vary slightly for different countries.
+The classification is based primarily on the weight of the aircraft, but this can vary slightly for different countries or regions such as the EEA.
 
-The ICAO WTC categorisation is based on observations and calculations performed on aircraft models that are some 50 years old.
+The ICAO WTC categorisation is based on observations and calculations performed on aircraft models that are in some cases 50 years old.
 Even though those separations are considered reliable, they are often conservative and outdated in view of newer aircraft models and as such generate unnecessary over-separation in some cases.
-In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an intergovernmental organisation assisting in safe and efficient air traffic operations.
+In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an inter-European organisation with wide responsibility for ensuring safe and efficient air traffic operations in Europe including the operation of the Central Flow Management Unit (CFMU) in Brussels and the Maastricht Area Control Center.
 
@@
 
-Apart from wake turbulence separation, the other major factor that affects capacity is Runway Occupancy Time (ROT), which is the time interval an aircraft occupies a runway during landing or departure.
-The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, per se, does limit the throughput of a runway.
+Apart from wake turbulence separation, the other major factor that affects capacity is Runway Occupancy Time (ROT), which is the time interval that an aircraft occupies the runway during landing or departure.
+The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, consequently does limit the throughput of a runway.
 The emphasis in this project will be on runway occupancy times for arriving aircraft (AROT) during peak traffic hours at the airport.
@@
 
-The aim of this study is to estimate the constraints for implementing the RECAT-EU wake turbulence categorisation and separation minima for Keflavík International Airport, the effect this can produce for airfield throughput and when is the runway occupancy a limiting factor for increased capacity.
+The aim of this study is to estimate the constraints for implementing the RECAT-EU wake turbulence categorisation and separation minima for Keflavík International Airport, the effect this can produce for airfield throughput and when the runway occupancy is a limiting factor for increased capacity.
 

### 2019-04-28 introduction.tex
fixes up to constraints for capacity

@@
 Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards.
-When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and are combined into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
+When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and are combined into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_airbus_safety, Breitsamter2011Feb, gerz_commercial_2002}.
 Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards.
@@
     \caption[Wake vortex roll-up process]{Wing vortex evolution and roll-up process.
-Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one~\cite[p.~043]{magazine_aibus_safety}} \label{fig:vortex_develop}
+Two main vortices form behind the aircraft turning in opposite directions, clockwise behind the left wing (seen from behind) and anti-clockwise behind the right one~\cite[p.~043]{magazine_airbus_safety}} \label{fig:vortex_develop}
 \end{figure}
@@
 The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure~\ref{fig:vortex_ground_effect}).
-
 \begin{figure}[h]
